### 2020-01-15 convivial-2020/tex/manuscript.tex
Flesh out some of later sections, try latex it

\documentclass[english,submission]{programming}
\usepackage[backend=biber]{biblatex} % Use Biblatex
\addbibresource{example.bib}

\begin{document}

\paperdetails{
  perspective=engineering,
  area={General-purpose programming}
}

\title{What does it take to create with domain-appropriate tools? A case study on the ``OROM'' system.}
\author{Joel Jakubovic}
\affiliation{University of Kent, Canterbury}

\keywords{paper, showcase, lorem ipsum}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002944.10011122.10003459</concept_id>
<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}`

\maketitle

\begin{abstract}
  This paper shows...
\end{abstract}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

As someone who can code, I have already passed the first and most
important hurdle for making full use of the potential of my computer.
However, even in this supposedly empowered state, I am still far away
from feeling the relationship between myself and software as between
artisan and material, free to shape it into any form with effort
proportional to complexity.

One would have thought that software-creation acts like hypothetical
super-intelligent AI. That is: even though we start from a primitive
base in the 50s (or even today), there would surely be a recursive
process of self-improvement, building better software-creation tools
with the existing ones, until an ``expressivity singularity'' where
software becomes a workable material as described.

However, this didn't happen. Or at least, it is happening glacially
slowly. The brute fact is that whenever you want to create software, you
go to a text editor and figure out how to translate your design into
that. The text editors, being software, were written with the help of
previous text editors, and so on. It's undeniable that text editors have
improved, even if you think it peaked with Emacs. We just don't seem
able to go beyond them.

\begin{quote}
You discuss what's wrong with text editor in the next few paragraphs,
but I think it would be useful to give some brief summary here so that
readers don't think that ``we need to go beyond text editors'' is
unmotivated - I think the summary of the text below is that you want to
create software using tools that are closer to the domain of the
software (text is far from everything; visual editors would be closer to
a more visual domain) This general formulation might also nicely frame
some of the things you say next.
\end{quote}

Amdahl's Law generalises the following idea: even when you spend hours
of effort doubling the performance of a component used 1\% of the time,
your reward is a system overall improved by a mere 0.5\%.

Text coding is certainly ubiquitous, the 99\% case in programming, so
you might wonder where I'm going with this. Well, a small improvement to
text editing, if adopted by everyone, certainly does have a massive
\emph{intermediate} effect -- but this only \emph{matters} to the extent
that text was helping the programmers in the first place. If my goal is
to draw or animate pictures, or create a digital synth from a frequency
spectrogram, then giving me the ability to auto-indent my SVG markup is
rather underwhelming as a productivity increase, as it doesn't target
the core of the enterprise that makes it so hard.

My experience of coding, most of the projects requiring shapes (such as
GUIs), leads me to conclude that no matter how much I improve my skill
at a particular language, knowledge of libraries or even general coding
ability, my predicament stays the same. Our basic method of creating
software is optimised for an ever-diminishing proportion of the software
we actually want to make; ill-optimised for the graphics, layout,
interactivity and and basic physics -- more on this later -- that we
usually require.

Whenever I work on these I feel stuck in a box I know I can never escape
from: that box is the text editor, a fixed conduit through which all
\emph{fundamental} changes to my program must pass. It's not a part of
the system I am building, so I can't even make use of features of the
thing I'm developing, to make its own development easier.

Surely the trick is to \emph{use} coding to build something \emph{better
than it}. And then use that, to build something even better. But there
is an enormous breadth and depth of philosophies here, along with many
failed historical attempts to do better -- or at least, ones that failed
to catch on. But even worse than this is that in my very \emph{language}
here I am making the same mistake as the text editor -- speaking in
unqualified terms of ``better'' and ``worse'' as if there really is a
one-size-fits-all solution to software creation!

Of course what we \emph{really} want is the ability for people to create
\emph{in the way that they think is best} in their particular context --
to equip them to feasibly create the tools that suit them for the thing
they want to make. And second-order tools that suit them for making the
first-order tools, etc. It would do no good to replace text-imperialism
with anything-else-imperialism, which is one common interpretation of
calls for alternatives. If someone wants to type out pictures in ASCII,
let them -- whether they do it for a challenge, or even if they find
that more natural for themselves. But equally, if I want to do it
another way, then please give me that affordance.

\begin{quote}
Is programming like `craft'? I guess some artisans make their own tools,
but most likely not in the higher-order sense you are discussing here.
This suggests that programming is different in some sense - but in what
sense?
\end{quote}

This is how I segue into the software artefact for which I have been
attempting to build a natural representation. True, it is a programmer's
artefact, but it is still representative of what any normal person has
to do, insofar as:

\begin{enumerate}
\def\labelenumi{\alph{enumi})}
\tightlist
\item
  Wanting to create a piece of software (for whatever reason)
\item
  Having in mind a natural way to represent it as it's being built.
\end{enumerate}

\begin{quote}
I think it would be good to say (fairly generally) what kind of system
do you imagine we should be using (and goals that your OROM/SVG is
getting closer to). This also clarifies why nothing like that exists
already. I suppose you want something that lets you program in a way
that is ``close to the problem domain'' (i.e.~visually if the domain is
visual) and something that has no arbitrary boundaries
(i.e.~meta-circular?). Boxer is the former, but not the latter;
Smalltalk/LISP is the latter but not the former.
\end{quote}

\hypertarget{the-orom-system}{%
\section{The OROM system}\label{the-orom-system}}

When I first read the paper Open, Reusable Object Models, I was hooked
on its idea of a small but expressive starting system that could be
self-improved into anything. It describes a late-bound, Smalltalk-style
objects and messaging environment that the authors call ``Id'' -- but
which I refer to as OROM, for pronunciation and Googleability.

\begin{quote}
How much does the reader need to know about OROM for the next
paragraphs? I think they need to know there are objects with references
that send messages to each other - this will be clear to those who know
what `Smalltalk-style' means, but not others. It would be useful to give
a very brief summary before getting to the next part of the story.
\end{quote}

The paper as a publication follows the usual patterns. It consists of
mostly prose, several code listings, some diagrams, and full C sources
for a sample implementation at the end. Insofar as this is normal, I
cannot criticise it. But reading this paper, and trying to grasp their
ideas, brought to my attention how inadequate this norm can sometimes
be.

\begin{quote}
The above feels a bit more chatty than necessary - you could probably
shorten the paragraph into just one sentence without losing much.
\end{quote}

In order to understand it in the first place, I repeatedly found myself
drawing extra diagrams. For example, the first acts of the running
system boil down to initialising the three or so objects. This consists
of allocating memory, interpreting it as a C struct and then filling in
fields in a mundane manner. I had great difficulty grasping the
specifics in my head, but when I drew tables in the style of the paper's
diagrams I readily saw what what going on.

Still, areas like messaging semantics could be quite confusing. After
all, should we expect to be able to predict the entire future evolution
of a dynamical system from a static description of its initial state,
such as source code? Is this not why debuggers exist? Having
``de-compiled'' English text and C source code into object diagrams, it
is a shame to have to compile it all back to struct member assignments.

Worst of all, the reference system does not even have text I/O when it
is run, let alone some sort of GUI. That is, the (un)intended user
interface for this project is a C debugger! Faced with the necessity of
adding \emph{some} UI, it seemed a waste of effort to end up with a
system that must be continually polled for its current state at a
terminal prompt. If I naturally think of this system as 2D tables, why
can't that be how the running system looks? I do not have to keep
polling my eyes for what state my diagrams are in.

But further than that -- why can't the system be \emph{built} out of
tables in the first place? Shouldn't this be the main takeaway from the
amount of time we spend prototyping, explaining and designing software
as diagrams on paper? Why must the ``natural representation'' be
restricted to the finished product?

Thus was my natural representation decided. My first attempt to make it
a reality was a partial success: a webpage made of HTML tables, evolved
via JavaScript.

\hypertarget{orom-as-html-tables}{%
\subsubsection{OROM as HTML tables}\label{orom-as-html-tables}}

(OROM/HTML screenshot)

(Describe basic encoding of obj-dicts as tables, use of css class, etc)

Unfortunately, the choice of ordinary HTML as a substrate proved rather
two-edged. I was grateful for the browser's management of graphical
layout, resizable text fields, and keeping the DOM tree synchronised
with what you actually see. This last property enabled me to make the
decision to \emph{directly} encode much of the system state in the DOM,
achieving basic liveness (the thing on the screen is the actual thing)
for the keys and values of obj-dicts.

On the other hand, the browser requires many features to do its job of
rendering complex web pages. And sadly, as its client, I could only make
use of those capabilities which the W3C had decided, at the time of
authorship, were worth the effort exposing in JavaScript. For anything
else, the browser is a black box, and this was very frustrating in the
following case.

\hypertarget{the-radical-concept-of-arrows-that-stay-on-the-shapes}{%
\paragraph{The Radical Concept of Arrows That Stay On The
Shapes}\label{the-radical-concept-of-arrows-that-stay-on-the-shapes}}

A key aspect of the OROM system is that there is an object \emph{graph}.
That is, obj-dicts can have entries pointing to other obj-dicts, without
restriction to a tree structure. Drawing arrows between things to denote
this is a no-brainer, and I wanted it in my substrate for OROM to match
the diagrams.

A small problem to surmount first: even though I could hijack the
\texttt{\textless{}table\textgreater{}} for its display properties, what
element could I hijack to make arrows between arbitrary points? Luckily,
there was SVG at my disposal, which could be persuaded to display over
the HTML. However, another key feature of my intended substrate was to
be able to rearrange and resize the boxes as desired. So I would also
need to detect changes to the position and size of an element.

Bizarrely, there is no such facility provided for HTML elements. This,
despite the fact the browser \emph{needs} this functionality hence it
must reside \emph{somewhere} inside the black box.

Reluctantly, I stuck with my plan B: each object has a numerical ID and
pointers are just fields containing a number, followed using a deref()
function.

At this point it was starting to look like a mistake not to have done
the whole thing in SVG. I would gain full graphical freedom, though also
lose some benefits of the browser's managing it on my behalf. I already
knew from experience the surprising complexity of a DIY approach to
layout, model-view updates, and interactivity. But such an exercise
would be an opportunity to carefully observe this tedium, and
crystallise some of my intuitions about why it is so consistently
frustrating.

\hypertarget{orom-as-files-and-directories}{%
\subsubsection{OROM as files and
directories}\label{orom-as-files-and-directories}}

\begin{quote}
I would probably put this after the SVG part or perhaps even later once
you discuss more general principles (like the idea of polyfilling to
make a domain substrate into a programmable system). Then you can give
this as another example.
\end{quote}

\hypertarget{orom-as-svg-trees}{%
\subsubsection{OROM as SVG trees}\label{orom-as-svg-trees}}

(Screenshot of OROM/SVG showing boxes, code and arrows)

In this version, obj-dicts are encoded as nested SVG
\texttt{\textless{}rect\textgreater{}}s and other elements, reminiscent
of diSessa's Boxer. This was a significant departure from the table
representation, and even though SVG supports (some) nested HTML via
\texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
the possibility of multiple levels of nesting.

OROM/SVG more or less realises my desired substrate for implementing
OROM. Needless to say, this version was far more challenging and took
much longer to reach a satisfactory state. However, it is precisely this
drudgery that brings me to a better understanding of this paper's
question: \emph{what has it taken?} I shall discuss this in the form of
broad patterns or themes that stand out to me from my development
experience, hammered home by these OROM projects.

\begin{quote}
Why is OROM/SVG more challenging? I think it would be useful to sketch
here, because it reveals some useful principles. There is more
polyfilling (arrows and resizing is not in the SVG substrate itself),
but at least it's possible to add those (unlike in HTML)
\end{quote}

\hypertarget{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}{%
\subsection{Inevitable Requirements Of Most Software (and the work we
must do to meet
them)}\label{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}}

As programmers, we have a maxim along the lines of: when you find
yourself repeating the same thing over and over again, factor it into a
``thing'' and let the computer do the duplication. Some consider the
``design pattern'' to be simply what a ``language feature'' looks like
when it's not part of the syntax.

\begin{quote}
The jump from very general abstraction idea to design patterns and
language features in the above is not exactly clear - it would be good
to help the reader understand why are you talking about this here.
\end{quote}

As I developed OROM/SVG I found myself implementing many such patterns.
It seems that languages are often at the wrong level of abstraction for
the requirements of modern software, necessitating the same boilerplate
per project just to get up to basic functionality. This burden either
falls on the author, or on the wider community to build and maintain
higher-level frameworks, syntax extensions, etc.

By contrast, these languages seem much better adapted to e.g.~batch-mode
printf()-centric applications. That they fail for the \emph{common case}
should be concerning. I will now present what seem to be the inevitable
demands of this common case.

\begin{quote}
I'm not sure I understand what the above opening is trying to say - I
suppose it might be because it is quite high-level/general. What is this
section about more specifically? Is it about what certain kind of
software consists of and what we need to build that? I would also be a
bit careful about things like ``most modern software'' below. It would
be good to instead characterize what kind of software are you interested
in - end-user graphical systems? (a microservice or something like that
does not need SVG\ldots{})
\end{quote}

\hypertarget{as-a-mere-consumer}{%
\subsubsection{As A ``Mere Consumer''}\label{as-a-mere-consumer}}

I begin by describing expectations of software that even ``end-users''
hold, consigned as they are to more or less accept what we give them.

\hypertarget{retained-mode-vector-graphics}{%
\paragraph{Retained-Mode Vector
Graphics}\label{retained-mode-vector-graphics}}

Most software is designed for the subset of people who have a colour
display they can perceive. So right away it is going to require ways to
draw coloured shapes. There are usually libraries for this (though note:
not part of the language), but some only provide \emph{immediate mode}:
commands to instantaneously rasterise pixels to a buffer. This is not
enough for modern software, as we often expect animation, or at least to
see things change as we interact. Most often we wish to see \emph{small
changes} to the \emph{same} shapes, rather than completely different
shapes altogether; the reification that this requires to persist between
frames, is known as \emph{retained mode.}

On this requirement, SVG fits the bill very well. Although it is not
part of the JavaScript language \emph{per se}, it is a standard and
widely supported technology of the Web \emph{platform}. We can observe
that anyone with a browser \emph{in principle} has access to a powerful
vector graphics editor -- just one with no GUI. I will return to this
later.

The SVG tree keeps the nice properties of the DOM, such as updating the
display when shape parameters are changed. This is well-adapted to
``I/O-bound'' software like mine, where things change only in response
to user input. If I wanted animation, this boils down to a regular
``advance simulation'' signal, and would require setting up some
rendering loop. Alternatively, there is the W3C's chosen ontology of CSS
animations, but see section ??.

\hypertarget{basic-assumptions-about-physical-objects}{%
\paragraph{Basic Assumptions About Physical
Objects}\label{basic-assumptions-about-physical-objects}}

In any software making use of vector graphics, there is usually some
level of ``physics'' expected by users. This need not be nearly as
exhaustive as the word ``physics'' might imply, as in e.g.~physics
engines for games; I feel it is important to recognise it for what it is
instead of conceiving physics as an inherently complex thing to be found
only in specialised simulations.

\begin{quote}
People might not find this easy to believe - but I think you can
convince them by saying that pretty much all mobile phone apps have some
sort of ``phone touch physics'' (with sliding, opening menus etc.).
\end{quote}

All humans learn a basic set of expectations about the things they see
around them. Some of these, such as ``things fall down'', are not
generally appropriate to software UIs -- perhaps because the screen has
a role in our lives a more of a table work surface rather than a
vertical wall, even if it is vertical in real life.

The level of physics in software tends to not involve force, or mass, or
very much at all, merely position and space. One thing that all usable
software must do, for example, is not crush many visually complex
shapes, such as lines of text, into the same region, since it becomes
unreadable. Such concepts of ``solid objects do not intersect'' or
``only things at different layers may overlap'' are basic rules
inherited from the real world of graphical presentation.

I feel the need to point this out, because by default the computer does
not know even the most obvious things about how space works, so we must
laboriously algorithmise this intuitive concept. This is not only true
in the case of 2-dimensional visual domains, but even in the
1-dimensional case of memory allocation. The physics of 1D memory are
something like this:

\begin{itemize}
\tightlist
\item
  This number range 0000-FFFF is like a space, numbers = points
\item
  Every point has at most one owner block
\item
  These blocks are contiguous, finite ranges (i.e.~1D boxes)
\end{itemize}

Hence the boilerplate involved to realise this in any domain with
something resembling space. Far from being a niche topic in games and
graphics, spatial partitioning algorithms and data structures have
surprising relevance to more ordinary software. Both memory allocation
and graphical layout are essential to today's; shame that only one of
those has been recognised as such and entered the runtime of modern
languages.

\hypertarget{translationally-rigid-bodies}{%
\subparagraph{Translationally Rigid
Bodies}\label{translationally-rigid-bodies}}

When you have both a screen and a pointing device (e.g.~touch or mouse),
immediately it becomes worth having ways to move things around in at
least a minimally realistic way. We can debate the appropriateness of
Direct Manipulation (DM) for various situations. But it does make a lot
of sense in simple cases, such moving around subdivisions of space
(e.g.~windows) or elements of a graphical design.

In OROM, the obvious candidate for this is the obj-dicts, plus all
nested boxes in OROM/SVG. If I move the top-level rect, then I expect
its children to move with it. This is simply the translational physics
of rigid bodies: this set of points all move together. Of course, proper
rigid bodies might also rotate and have mass, but this is usually
undesirable for UI elements.

Translational rigidity can be expressed as the points X and Y always
having the same displacement from each other. Or, when one point is
moved, the rest also move by the same delta. This is a problem of
preserving the relationship over time, which was a significant area of
OROM/SVG.

\hypertarget{maintaining-relationships-over-time}{%
\paragraph{Maintaining Relationships Over
Time}\label{maintaining-relationships-over-time}}

The model of state-mutation present in most imperative languages is what
I call ``dumb'' state. The language provides an affordance to change any
part of the state to a new value, but nothing else.

What more could there be? Well, in \emph{every} software system there
are certain rules, or ``invariants'' of internal consistency, such as
``translational rigidity'' above. Often, changes to any part of the
system are permissible, but only if connected or dependent parts of the
state change in response.

The job of keeping track of who depends on whom can fall either on the
programmer, or the computer. If the programmer has to do this, they can
only go so far managing and simulating in their head. As systems grow
more complex, it is only natural to try and make the computer more
intelligent to do this work. What I am building up to is that whenever
we (or I) consider constraints, or ``reactive'' programming, or the
``Observer'' pattern as things you only wheel out on special occasions,
we only deceive ourselves into doing the same work less explicitly. It
seems that such ``live state'' should be the expected common case for
software development.

If a platform does not provide a means to causally link and unlink bits
of live state, then this must form part of the standard boilerplate.

Such was the case in OROM/SVG. I implemented the system in an OOP
fashion, and the Observable class is the most widely used. It wraps a
current value and a list of subscribers, notifying them when it changes.
It can be seen that this supplanted JavaScript's own ``dumb'' state by
the frequency of the change() function throughout the code.

Getting an object to follow the mouse pointer, for example when
dragging, is conceptually very simple: an ``always equal'' relation. In
OROM/SVG founded on live-state, this can be expressed in much the same
way:

\begin{verbatim}
subscribe(object.position, pointer.position);
\end{verbatim}

(By default, an Observable A responds to a change from Observable B by
adopting B's new value.)

\hypertarget{nut-cracking-with-sledgehammers}{%
\paragraph{Nut-Cracking With
Sledgehammers}\label{nut-cracking-with-sledgehammers}}

Speaking about vector graphics, physics, layout and constraint
maintenance might give the impression of high \emph{conceptual}
complexity at the heart of even simple software. This is not quite true,
which makes it all the worse that there is yet still immense
\emph{incidental} or implementation complexity.

We are conditioned to only think of these in their most general forms.
The vector graphics I use in OROM/SVG are just rects, lines, circles and
text; a fraction of the full capability of SVG. The ``basic physics'' I
use is dwarfed by fully general 2D or 3D physics engines. The only
layout algorithm I had the patience to implement was a simple way to
expand a list of boxes to fit in a new child at the bottom; the
affordance to place and size boxes \emph{manually} staves off the rest
for the time being! Yet search for material on layout algorithms and it
can seem like Fully General Linear Inequality Solvers like Cassowary are
all there is.

The inevitable requirements I suggest here, do \emph{not} necessitate
Fully General anything. In fact, such representations would make it more
cumbersome to express what I wanted in OROM. As the old wisdom goes,
there's no point expressing a simple regex search as an arbitrary Turing
machine; my boxes don't \emph{have} a moment of inertia or a mass and I
don't \emph{need} a linear optimisation solver for my space management
-- for the time being.

Unfortunately, there doesn't seem to be much interest on the
smaller-scale instances of these problems. I suggest the name
``geometric physics'' for the subset of physics I have mentioned.

\begin{quote}
I think the above section talks about 3 things - (i) it describes some
of the things that you had to implement for OROM/SVG, (ii) it talks
about some general patterns that arise there (like shapes and physics)
and (iii) it mentions some implementation details. I found it a bit
confusing to have these all mixed together. It reads a bit as if you
were jumping from one random topic that you encountered when
implementing this - but that makes the section very specific to your
implementation. How can we make it clear that there are more general
principles here? I think it might be better if you started with some
larger example (hypothetical is fine!) to illustrate what kind of things
you'd like to build - e.g. imagine you're building PowerPoint in OROM -
what would be some of the things you'd need? (for both the end
application and the system that you're using to build it). This would
then nicely motivate the more specific topics (graphics, physics,
relationships).
\end{quote}

\hypertarget{as-a-developer}{%
\subsubsection{As A ``Developer''}\label{as-a-developer}}

I now turn to some things that were perhaps inevitable, in any case
useful, as the \emph{implementor} of OROM/SVG.

\hypertarget{practicality-vs.-high-aesthetics}{%
\paragraph{Practicality vs.~High
Aesthetics}\label{practicality-vs.-high-aesthetics}}

There is the view that insists on provably-correct code written with
strange Unicode symbols, and at the opposite end there is the attitude
of ``get the job done'', high style be damned. Each has its merits, but
for my task I quickly saw the latter route to be most suitable.

My favourite example of this has been the ``encoding'' of obj-dicts as
DOM trees containing particular arrangements of children. In OROM/HTML,
I would gleefully get/set state by calling code like this:

(Listing involving CSS, element querying / munging)

and in OROM/SVG the story is similar. It has the advantage of the
powerful developer tools included in browsers, and avoids the need to
manually synchronise separate Model and View for such encoded state.
With a little work, it also allowed the system a large degree of
externalisability -- see Section ??.

Contrast this with an approach that had to re-synthesise bits of the DOM
whenever something changed. Admittedly this does still happen to a
certain extent, but at least it was minimised where obvious.

Another example of practicality was in the issue of positioning and
sizing. In Section ?? I mentioned that I was able to sidestep some
layout work by offloading it to the user. While building the system I
tended to want to draw boxes into existence, and for that my brain's
existing aesthetic algorithms can decide where they go. There are areas
where the machine must be able to figure this out itself -- I don't want
to be constantly bothered by prompts to place boxes as they are
allocated by some running code -- but for one-offs it is a very
convenient way to do a little less yak-shearing.

\begin{quote}
I think you could adapt the above discussion a bit and focus less on
what you don't want (high modernism) and more on what you want - what is
that? I think it's simplifying the problem along a different axis - to a
certain domain with relatively simple interactions and aesthetics. (So
rather than trying to make minimal formal calculus that's provably
correct, you're trying to make minimal working interaction model that's
interesting)
\end{quote}

\hypertarget{positioning-and-sizing}{%
\paragraph{Positioning and Sizing}\label{positioning-and-sizing}}

The simple desire to move and resize boxes with the mouse motivated a
lot of the live-state and geometric physics ideas. This problem could be
considered a microcosm of OROM/SVG: what's a natural way I
\emph{conceive} of this behaviour, and could I implement it that way?

\begin{quote}
The above point (moving motivated live state and physics ideas) would be
a lot more obvious if you followed the structure with motivating example
first followed by an analysis\ldots{}
\end{quote}

It starts with a consideration of translational rigidity. In its most
primitive form, this is a relation between two points. Thus it is
natural to draw a line or ``rod'' between them. It seems that this rod
transmits changes in one of its endpoints directly to the other
endpoint. Since they both feel the same deltas, the displacement vector
between them is preserved.

(Diagram of delta transmission)

I like to see what I'm doing, so I wanted these rods to be visible and
thus somehow \emph{present in the SVG}. This was not hard; in SVG I draw
a \texttt{\textless{}line\textgreater{}}, though the background
machinery of the Rod class does all the work. There is also a Point
class that is used wherever manipulable points (SVG circles) are
intended.

Resizing (of boxes at least) could be achieved through rods that stay
horizontal or vertical. In the language of ``small differences'' spoken
by my live-state infrastructure, this is expressed as a rod
``transmitting deltas'' in the vertical and horizontal, ``absorbing''
the other component into itself. Mirroring a DOM rect to these rods is
as simple as subscribing its width and height to horizontal and vertical
rods' length Observables. This way, boxes can be resized from whatever
corner is convenient.

(Box demonstrating all three rod types)

I colour such ``half-rigid'' rods yellow and fully rigid rods red. The
``absorb all endpoint changes'' case is coloured green.

\begin{quote}
I don't think you explained what `half-rigid' means yet.
\end{quote}

Unfortunately, with these rods came possibly the most frustrating
technical challenges of the entire system. Initially I hoped to move
boxes as rigid bodies by temporarily making their border rods rigid.
However, the four border rods form a cyclic graph, as rods are not
directed. This, coupled with the unintended depth-first semantics of
Observable notification (a result of JavaScript function calls in a
loop), led to duplicate deltas applied twice and other nightmares. This
is the tip of an entire research iceberg stretching from Functional
Reactive Programming to internet routing and distributed algorithms.

I did manage to surmount this through kludging and compromise. But I do
not know whether these problems are merely a consequence of some design
decision I could change to escape it, or if they are intrinsic to my
(modest) UI goal.

\begin{quote}
This section reads more as an implementation detail - even though it
makes some interesting points about the physics \& rods system. However,
it might better to have it in a section that's more about implementation
than about general principles. Or perhaps you could restructure Section
3 to have (i) example (ii) general lessons (like physics etc.) and (iii)
specific points about implementation (e.g.~your concrete physics system
with rods, what you had to `polyfill')
\end{quote}

\hypertarget{visible-coordinate-systems}{%
\paragraph{Visible Coordinate
Systems}\label{visible-coordinate-systems}}

Rigidity in a flat world of sibling shapes is somewhat straightforward.
However, rigidity in the SVG world is more involved.

First of all, SVG shapes (e.g.~\texttt{\textless{}rect\textgreater{}})
are strictly leaf nodes, so if I wish to nest boxes within boxes, the
visible ``box'' must be a mere \emph{accessory} to the nestable element,
in my case a \texttt{\textless{}g\textgreater{}} (group). This means
that instead of resizing the x, y, width, height attributes of the
\texttt{\textless{}rect\textgreater{}}, only its width and height change
along with the transform attribute of its \emph{parent}
\texttt{\textless{}g\textgreater{}}. This was not too bad; just
subscribe this attribute, instead of the
\texttt{\textless{}rect\textgreater{}} position, to the top-left Point
handle.

All child elements of a node transform with it, so already SVG has baked
in a basic facility for translational rigidity. This is only available
as a tree hierarchy, but it is still useful. However, it conflicts with
my early decision to have Point objects all share the global co-ordinate
system. This way, connecting ``this point'' to the mouse pointer, etc.
is not infuriating to express. Still, it was necessary in the case of
certain elements -- especially those which must transcend the tree
structure altogether, like arrows between boxes -- to bite this bullet,
one way or another.

\begin{quote}
Maybe mention that this is caused by the mismatch between the substrate
and the programming system (substrate is tree, but programming is
graphs)?
\end{quote}

Again, I return to how we tend to work things out in the freedom of
paper. Co-ordinate systems, here merely positionally displaced, have
their origins here and there and have vectors between them. The rods
thus far let me visually express relations between global Points; now
was a question of expressing one global Point as a displacement from
another (the \texttt{\textless{}g\textgreater{}} transform). New rod
Observables p2\_from\_p1 and p1\_from\_p2 do the vector subtraction,
which can then be propagated as local co-ordinates to children. It is
nicer to express the relation (as well as see it!) this way -- see
Figure ?? above.

\hypertarget{context-appropriate-ontologies}{%
\paragraph{Context-appropriate
ontologies}\label{context-appropriate-ontologies}}

\ldots{}

Talk about: fixed ontologies (a rect IS x,y,w,h, and nothing else!)
imposed on programmer

Example from 3.1.1. mention: CSS animations?

``One size fits all'' rears its ugly head again.

I want: top-left, top-right, bot-left, bot-right \ldots{}

Instead of explaining ``the x y is the top left blah blah'' in English
in the W3C spec, they should make these relationships \emph{machine
readable} with information like:

A Rect is

\begin{itemize}
\tightlist
\item
  a Polygon
\item
  defined by 4 degrees of freedom x y width height
\item
  where the four vertices are

  \begin{itemize}
  \tightlist
  \item
    \texttt{{[}x,y{]}} ``top-left''
  \item
    \texttt{{[}x+width,y{]}} ``top-right''
  \item
    etc
  \end{itemize}
\end{itemize}

Reference to paper related to Antranig involving knitting\ldots{}?

\ldots{}

\hypertarget{extensional-functions}{%
\paragraph{Extensional Functions}\label{extensional-functions}}

aka dicts, Maps, std::vector, \ldots{}

\ldots{} \ldots{} Some parts of the OROM authors' C code were confusing
until I realised they were just the guts of a basic associative-array
implementation. \ldots{} \ldots{}

Inevitability / ubiquity of state descriptions that are at trees of
strings with pointers

Antranig ``natural co-ordinates''?

OROM/SVG as Boxer-like extensional functions

\hypertarget{externalisability}{%
\paragraph{Externalisability}\label{externalisability}}

Reference Antranig's stuff on this. How I dealt with event listeners etc

Necessity to development: frustrating to find bug, tear down, refresh,
lose work, build it up again!

Inspect Element: copy to save works, arbitrary edits \emph{should} work
(cuz it's on demand)

Work required to parse SVG trees into objects ``on demand''

Started heavyweight, ended lightweight (generate rods on demand, let SVG
move children)

\hypertarget{polyfilling}{%
\paragraph{Polyfilling}\label{polyfilling}}

Very advantageous to permit this e.g.~JavaScript

Stupid lack of XYZ in JS or SVG due to come out in the next spec, etc

Syntax sugar (SVG tree ops, vector ops)

\hypertarget{the-orom-system-as-a-part-of-the-solution}{%
\subsection{The OROM system as a part of the
solution}\label{the-orom-system-as-a-part-of-the-solution}}

Meta-circularity as empowerment, not just ``cool''

The allure

Other themes e.g.~homoiconicity of code and data

Problems: cognitive vs.~``official'' complexity

Maxwell's eqns metaphor: omits the years of study compressed into the
math symbols!

And yet, the allure of the Maxwell's Eqns of Software?

``Taming'' SVG - ability to use SVG in a vaguely sane way (vector
graphics editor)

Similarly for e.g.~3D OpenGL, Web Audio, sockets \ldots{}

\hypertarget{tensions-between-philosophies}{%
\subsubsection{Tensions between
philosophies}\label{tensions-between-philosophies}}

\hypertarget{stable-or-emergent-antranig-against-oop-dispatch}{%
\paragraph{Stable or emergent? (Antranig against OOP
dispatch)}\label{stable-or-emergent-antranig-against-oop-dispatch}}

\hypertarget{turing-complete-antranig-against-this}{%
\paragraph{Turing-complete? (Antranig against
this)}\label{turing-complete-antranig-against-this}}

\hypertarget{messaging-vs-read-write}{%
\paragraph{Messaging vs Read-Write?}\label{messaging-vs-read-write}}

\hypertarget{misc}{%
\section{Misc}\label{misc}}

A note about DIY and libraries

My other problem was one of cost-benefit. My aim was to \ldots{}

\acks
I want to thank ....

\printbibliography
\end{document}

### 2020-01-15 convivial-2020/tex/manuscript.tex
Abstract criteria

@@
 \addbibresource{example.bib}
+
+\lstdefinelanguage{JavaScript}{
+  keywords={break, case, catch, continue, debugger, default, delete, do, else,
+  finally, for, function, if, in, instanceof, new, return, switch, this, throw,
+  try, typeof, var, void, while, with, let},
+  morecomment=[l]{//},
+  morecomment=[s]{/*}{*/},
+  morestring=[b]',
+  morestring=[b]",
+  sensitive=true
+}
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 
-\title{What does it take to create with domain-appropriate tools? A case study on the ``OROM'' system.}
+\title{What does it take to create with domain-appropriate tools?}
+\subtitle{A case study on the ``OROM'' system}
 \author{Joel Jakubovic}
@@
 \end{abstract}
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
   improved, even if you think it peaked with Emacs. We just don't seem
-able to go beyond them.
-
-\begin{quote}
-You discuss what's wrong with text editor in the next few paragraphs,
-but I think it would be useful to give some brief summary here so that
-readers don't think that ``we need to go beyond text editors'' is
-unmotivated - I think the summary of the text below is that you want to
-create software using tools that are closer to the domain of the
-software (text is far from everything; visual editors would be closer to
-a more visual domain) This general formulation might also nicely frame
-some of the things you say next.
-\end{quote}
+  able to go beyond them where it matters, such as visual domains
+  ill-fitted to monospaced ASCII.
 
@@
   of effort doubling the performance of a component used 1\% of the time,
-your reward is a system overall improved by a mere 0.5\%.
-
-Text coding is certainly ubiquitous, the 99\% case in programming, so
-you might wonder where I'm going with this. Well, a small improvement to
+  your reward is a system overall improved by a mere 0.5\%. Now, text
+  coding is certainly ubiquitous, the 99\% case in programming, so you
+  might wonder where I'm going with this. Well, a small improvement to
   text editing, if adopted by everyone, certainly does have a massive
-\emph{intermediate} effect -- but this only \emph{matters} to the extent
-that text was helping the programmers in the first place. If my goal is
-to draw or animate pictures, or create a digital synth from a frequency
-spectrogram, then giving me the ability to auto-indent my SVG markup is
-rather underwhelming as a productivity increase, as it doesn't target
-the core of the enterprise that makes it so hard.
+  \emph{intermediate} effect --- but this only \emph{matters} to the
+  extent that text was helping the programmers in the first place. If my
+  goal is to draw or animate pictures, or create a digital synth from a
+  frequency spectrogram, then giving me the ability to auto-indent my SVG
+  markup is rather underwhelming as a productivity increase, as it doesn't
+  target the core of the enterprise that makes it so hard.
 
@@
   we actually want to make; ill-optimised for the graphics, layout,
-interactivity and and basic physics -- more on this later -- that we
+  interactivity and and basic physics --- more on this later --- that we
   usually require.
@@
   is an enormous breadth and depth of philosophies here, along with many
-failed historical attempts to do better -- or at least, ones that failed
-to catch on. But even worse than this is that in my very \emph{language}
-here I am making the same mistake as the text editor -- speaking in
-unqualified terms of ``better'' and ``worse'' as if there really is a
-one-size-fits-all solution to software creation!
+  failed historical attempts to do better --- or at least, ones that
+  failed to catch on. But even worse than this is that in my very
+  \emph{language} here I am making the same mistake as the text editor ---
+  speaking in unqualified terms of ``better'' and ``worse'' as if there
+  really is a one-size-fits-all solution to software creation!
 
   Of course what we \emph{really} want is the ability for people to create
-\emph{in the way that they think is best} in their particular context --
-to equip them to feasibly create the tools that suit them for the thing
-they want to make. And second-order tools that suit them for making the
-first-order tools, etc. It would do no good to replace text-imperialism
-with anything-else-imperialism, which is one common interpretation of
-calls for alternatives. If someone wants to type out pictures in ASCII,
-let them -- whether they do it for a challenge, or even if they find
-that more natural for themselves. But equally, if I want to do it
-another way, then please give me that affordance.
+  \emph{in the way that they think is best} in their particular context
+  --- to equip them to feasibly create the tools that suit them for the
+  thing they want to make. And second-order tools that suit them for
+  making the first-order tools, etc. It would do no good to replace
+  text-imperialism with anything-else-imperialism, which is one
+  interpretation of calls for alternatives. This dream goes beyond
+  familiar senses of what constitutes a ``craft'', as far as a strong
+  melding of tool and material.
 
-\begin{quote}
-Is programming like `craft'? I guess some artisans make their own tools,
-but most likely not in the higher-order sense you are discussing here.
-This suggests that programming is different in some sense - but in what
-sense?
-\end{quote}
+  Certainly in the real world there is only one ``material'' --- matter
+  --- and humanity as a whole does create ``higher-order'' tools. A
+  (future) software craft, however, could be expected to give this power
+  to \emph{individuals} instead of the community alone. The
+  one-size-fits-all approach is an understandable consequence of few
+  designers catering to many users --- perhaps due to the necessity to
+  specialise in topics, languages, libraries, and tools. Their complexity
+  is part of the general difficulty of software creation, exacerbated by
+  the one-size-fits-all approach, closing the circle.
 
-This is how I segue into the software artefact for which I have been
-attempting to build a natural representation. True, it is a programmer's
-artefact, but it is still representative of what any normal person has
-to do, insofar as:
+  What we need is some system that not only lets us create software in a
+  way that is ``close to the problem domain'' as decided by the
+  user-developer, but also can augment or change itself to adapt to a
+  different ``way of creating''. Existing systems seem to only have one of
+  these properties without the other. Smalltalk and LISP try to minimise
+  arbitrary commitments of language \emph{semantics} to be adaptable, but
+  their being textual languages is a fairly tough commitment to break out
+  of. And it is not so hard to make a specific, \emph{hard-baked} visual
+  or alternative programming tool --- but it is hard to make it
+  re-programmable \emph{without} having to go back to \emph{its} textual
+  source code.
+
+  Again, if someone wants to type out pictures in ASCII, let them ---
+  whether they do it for a challenge, or even if they find that more
+  natural for themselves. But equally, if I want to do it another way,
+  then please give me that affordance. This is how I segue into the
+  software artefact for which I have been attempting to build a natural
+  representation. True, it is a programmer's artefact, but it is still
+  representative of what any normal person has to do, insofar as:
 
@@
 
-\begin{quote}
-I think it would be good to say (fairly generally) what kind of system
-do you imagine we should be using (and goals that your OROM/SVG is
-getting closer to). This also clarifies why nothing like that exists
-already. I suppose you want something that lets you program in a way
-that is ``close to the problem domain'' (i.e.~visually if the domain is
-visual) and something that has no arbitrary boundaries
-(i.e.~meta-circular?). Boxer is the former, but not the latter;
-Smalltalk/LISP is the latter but not the former.
-\end{quote}
-
   \hypertarget{the-orom-system}{%
@@
   on its idea of a small but expressive starting system that could be
-self-improved into anything. It describes a late-bound, Smalltalk-style
-objects and messaging environment that the authors call ``Id'' -- but
-which I refer to as OROM, for pronunciation and Googleability.
+  self-improved into anything. It describes a late-bound\footnote{Fewer
+    commitments; more things determined by runtime conditions},
+  Smalltalk-style\footnote{OOP with more emphasis on object instances and
+    messaging (methods) as in a distributed system; less emphasis on class
+    hierarchies implementing traditional data structures} objects and
+  messaging environment that the authors call ``Id'' --- but which I refer
+  to as OROM, for pronunciation and Googleability.
 
-\begin{quote}
-How much does the reader need to know about OROM for the next
-paragraphs? I think they need to know there are objects with references
-that send messages to each other - this will be clear to those who know
-what `Smalltalk-style' means, but not others. It would be useful to give
-a very brief summary before getting to the next part of the story.
-\end{quote}
+  In summary, an object is a key-value mappings of state which can change
+  as a result of messages received by it. A message is sent by first
+  \emph{bind}-ing its name to its implementation, i.e.~specific code which
+  is then run in the context of the receiver. ``Bind'' is accomplished via
+  another message send, this time to the receiver's \emph{vtable}, or
+  ``behaviour''. This is another object whose state defines a particular
+  way of associating ``message name'' to ``implementation code''. This
+  continues up the vtable-chain, and terminates at a base case. The system
+  as a whole is ``bootstrapped'' into existence by its initialisation
+  code. Each step of this code makes use of any parts of the system set up
+  by the previous steps.
 
-The paper as a publication follows the usual patterns. It consists of
-mostly prose, several code listings, some diagrams, and full C sources
-for a sample implementation at the end. Insofar as this is normal, I
-cannot criticise it. But reading this paper, and trying to grasp their
-ideas, brought to my attention how inadequate this norm can sometimes
-be.
+  The paper itself consists of mostly prose, several code listings, and
+  full C sources for a sample implementation at the end. It also provides
+  several diagrams. But to understand it, I repeatedly found myself
+  drawing \emph{extra} diagrams. For example, the first acts of the
+  running system boil down to initialising the three or so objects. This
+  consists of allocating memory, interpreting it as a C struct and then
+  filling in fields in a mundane manner. I had great difficulty following
+  the specifics in my head, but when I drew tables in the style of their
+  diagrams I readily saw what what going on. I personally would have
+  preferred these to have been in the paper in the first place, but I am
+  not necessarily representative of those who read it. The
+  ``one-size-fits-all'' approach is perhaps unavoidable for static media
+  like print.
 
-\begin{quote}
-The above feels a bit more chatty than necessary - you could probably
-shorten the paragraph into just one sentence without losing much.
-\end{quote}
-
-In order to understand it in the first place, I repeatedly found myself
-drawing extra diagrams. For example, the first acts of the running
-system boil down to initialising the three or so objects. This consists
-of allocating memory, interpreting it as a C struct and then filling in
-fields in a mundane manner. I had great difficulty grasping the
-specifics in my head, but when I drew tables in the style of the paper's
-diagrams I readily saw what what going on.
-
-Still, areas like messaging semantics could be quite confusing. After
-all, should we expect to be able to predict the entire future evolution
-of a dynamical system from a static description of its initial state,
-such as source code? Is this not why debuggers exist? Having
+  Other areas like messaging semantics could still be quite confusing.
+  After all, should we expect to be able to predict the entire future
+  evolution of a dynamical system from a static description of its initial
+  state, such as source code? Is this not why debuggers exist? Having
   ``de-compiled'' English text and C source code into object diagrams, it
@@
 
-But further than that -- why can't the system be \emph{built} out of
+  But further than that --- why can't the system be \emph{built} out of
   tables in the first place? Shouldn't this be the main takeaway from the
@@
   \hypertarget{orom-as-html-tables}{%
-\subsubsection{OROM as HTML tables}\label{orom-as-html-tables}}
+  \subsection{OROM as HTML tables}\label{orom-as-html-tables}}
 
@@
 
-(Describe basic encoding of obj-dicts as tables, use of css class, etc)
+  I was grateful for the browser's management of graphical layout,
+  resizable text fields, and keeping the DOM tree synchronised with what
+  you actually see. This last property enabled me to make the decision to
+  \emph{directly} encode much of the system state in the DOM, achieving
+  basic liveness (the thing on the screen is the actual thing) for the
+  keys and values of obj-dicts. For each obj-dict I used a two-column
+  \texttt{\textless{}table\textgreater{}} within a
+  \texttt{\textless{}div\textgreater{}}, matching the diagrams, and gave
+  the rows CSS class names matching the keys for easy lookup. Sending
+  messages relied on the JavaScript console, but existing values in text
+  boxes (such as method implementations) could be edited directly. Also
+  thanks to the browser, I could easily make text boxes and
+  \texttt{\textless{}div\textgreater{}}s resizable as necessary.
 
-Unfortunately, the choice of ordinary HTML as a substrate proved rather
-two-edged. I was grateful for the browser's management of graphical
-layout, resizable text fields, and keeping the DOM tree synchronised
-with what you actually see. This last property enabled me to make the
-decision to \emph{directly} encode much of the system state in the DOM,
-achieving basic liveness (the thing on the screen is the actual thing)
-for the keys and values of obj-dicts.
-
-On the other hand, the browser requires many features to do its job of
-rendering complex web pages. And sadly, as its client, I could only make
-use of those capabilities which the W3C had decided, at the time of
-authorship, were worth the effort exposing in JavaScript. For anything
-else, the browser is a black box, and this was very frustrating in the
-following case.
+  However, the choice of ordinary HTML as a substrate proved rather
+  two-edged. The browser requires many features to do its job of rendering
+  complex web pages. And sadly, as its client, I could only make use of
+  those capabilities which the W3C had decided, at the time of authorship,
+  were worth the effort exposing in JavaScript. For anything else, the
+  browser is a black box, and this was very frustrating in the following
+  case.
 
   \hypertarget{the-radical-concept-of-arrows-that-stay-on-the-shapes}{%
-\paragraph{The Radical Concept of Arrows That Stay On The
+  \subsubsection{The Radical Concept of Arrows That Stay On The
   Shapes}\label{the-radical-concept-of-arrows-that-stay-on-the-shapes}}
@@
   Reluctantly, I stuck with my plan B: each object has a numerical ID and
-pointers are just fields containing a number, followed using a deref()
-function.
+  pointers are just fields containing a number, followed using a
+  \texttt{deref()} function.
 
@@
   \hypertarget{orom-as-files-and-directories}{%
-\subsubsection{OROM as files and
+  \subsection{OROM as files and
   directories}\label{orom-as-files-and-directories}}
@@
   \hypertarget{orom-as-svg-trees}{%
-\subsubsection{OROM as SVG trees}\label{orom-as-svg-trees}}
+  \subsection{OROM as SVG trees}\label{orom-as-svg-trees}}
 
@@
   \hypertarget{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}{%
-\subsection{Inevitable Requirements Of Most Software (and the work we
+  \section{Inevitable Requirements Of Most Software (and the work we
   must do to meet
@@
   \hypertarget{as-a-mere-consumer}{%
-\subsubsection{As A ``Mere Consumer''}\label{as-a-mere-consumer}}
+  \subsection{As A ``Mere Consumer''}\label{as-a-mere-consumer}}
 
@@
   \hypertarget{retained-mode-vector-graphics}{%
-\paragraph{Retained-Mode Vector
+  \subsubsection{Retained-Mode Vector
   Graphics}\label{retained-mode-vector-graphics}}
@@
   that anyone with a browser \emph{in principle} has access to a powerful
-vector graphics editor -- just one with no GUI. I will return to this
+  vector graphics editor --- just one with no GUI. I will return to this
   later.
@@
   \hypertarget{basic-assumptions-about-physical-objects}{%
-\paragraph{Basic Assumptions About Physical
+  \subsubsection{Basic Assumptions About Physical
   Objects}\label{basic-assumptions-about-physical-objects}}
@@
   around them. Some of these, such as ``things fall down'', are not
-generally appropriate to software UIs -- perhaps because the screen has
+  generally appropriate to software UIs --- perhaps because the screen has
   a role in our lives a more of a table work surface rather than a
@@
   \hypertarget{maintaining-relationships-over-time}{%
-\paragraph{Maintaining Relationships Over
+  \subsubsection{Maintaining Relationships Over
   Time}\label{maintaining-relationships-over-time}}
@@
   \hypertarget{nut-cracking-with-sledgehammers}{%
-\paragraph{Nut-Cracking With
+  \subsubsection{Nut-Cracking With
   Sledgehammers}\label{nut-cracking-with-sledgehammers}}
@@
   don't \emph{need} a linear optimisation solver for my space management
--- for the time being.
+  --- for the time being.
 
@@
   \hypertarget{as-a-developer}{%
-\subsubsection{As A ``Developer''}\label{as-a-developer}}
+  \subsection{As A ``Developer''}\label{as-a-developer}}
 
@@
   \hypertarget{practicality-vs.-high-aesthetics}{%
-\paragraph{Practicality vs.~High
+  \subsubsection{Practicality vs.~High
   Aesthetics}\label{practicality-vs.-high-aesthetics}}
@@
 
-(Listing involving CSS, element querying / munging)
+  \begin{verbatim}
+  Entity.prototype.addState = function(key) {
+    let tr = document.createElement('tr');
+    tr.setAttribute('class', key);
 
-and in OROM/SVG the story is similar. It has the advantage of the
-powerful developer tools included in browsers, and avoids the need to
-manually synchronise separate Model and View for such encoded state.
-With a little work, it also allowed the system a large degree of
-externalisability -- see Section ??.
+    let td = document.createElement('td');
+    td.setAttribute('class', 'key');
+    td.textContent = key;
+    tr.appendChild(td);
+    ...
+  \end{verbatim}
+
+  and in OROM/SVG the story is similar. It takes advantage of the powerful
+  developer tools included in browsers (e.g.~the DOM inspector), and
+  avoids the need to manually synchronise separate Model and View for such
+  encoded state. With a little work, it also allowed the system a large
+  degree of externalisability --- see Section ??. I consider this as
+  practical, expedient, ``hijacking'' of these substrates because it
+  either goes against their intended use (HTML tables for presentation
+  rather than semantics) or are simply ``odd'' (using certain arrangements
+  of shapes to ``encode'' machine-readable semantics).
 
@@
   existing aesthetic algorithms can decide where they go. There are areas
-where the machine must be able to figure this out itself -- I don't want
-to be constantly bothered by prompts to place boxes as they are
-allocated by some running code -- but for one-offs it is a very
+  where the machine must be able to figure this out itself --- I don't
+  want to be constantly bothered by prompts to place boxes as they are
+  allocated by some running code --- but for one-offs it is a very
   convenient way to do a little less yak-shearing.
@@
   \hypertarget{positioning-and-sizing}{%
-\paragraph{Positioning and Sizing}\label{positioning-and-sizing}}
+  \subsubsection{Positioning and Sizing}\label{positioning-and-sizing}}
 
@@
   \hypertarget{visible-coordinate-systems}{%
-\paragraph{Visible Coordinate
+  \subsubsection{Visible Coordinate
   Systems}\label{visible-coordinate-systems}}
@@
   is not infuriating to express. Still, it was necessary in the case of
-certain elements -- especially those which must transcend the tree
-structure altogether, like arrows between boxes -- to bite this bullet,
+  certain elements --- especially those which must transcend the tree
+  structure altogether, like arrows between boxes --- to bite this bullet,
   one way or another.
@@
   another (the \texttt{\textless{}g\textgreater{}} transform). New rod
-Observables p2\_from\_p1 and p1\_from\_p2 do the vector subtraction,
-which can then be propagated as local co-ordinates to children. It is
-nicer to express the relation (as well as see it!) this way -- see
-Figure ?? above.
+  Observables \texttt{p2\_from\_p1} and \texttt{p1\_from\_p2} do the
+  vector subtraction, which can then be propagated as local co-ordinates
+  to children. It is nicer to express the relation (as well as see it!)
+  this way --- see Figure ?? above.
 
   \hypertarget{context-appropriate-ontologies}{%
-\paragraph{Context-appropriate
+  \subsubsection{Context-appropriate
   ontologies}\label{context-appropriate-ontologies}}
 
-\ldots{}
+  Each API has its own conventions, including a way of naming and
+  structuring expressions --- an ontology. The one-size-fits-all approach
+  is exemplified in such interfaces. For example, in SVG we express a
+  rectangle by
+  \texttt{\textless{}rect\ x="10"\ y="10"\ width="600"\ height="400"\textgreater{}}.
+  The SVG specification \emph{defines} to the user that a rect \emph{is} a
+  top-left corner, a width, and a height --- and that's it. However, this
+  SVG-approved parametrisation of a rectangle is far from the only
+  one(1){[}https://www.shift-society.org/salon/papers/2018/critiques/critique-semprola.pdf{]},
+  and thus is, unsurprisingly, ill-fitted to some contexts.
 
-Talk about: fixed ontologies (a rect IS x,y,w,h, and nothing else!)
-imposed on programmer
+  For example, I find it natural to resize boxes by dragging any of their
+  four corners, so I wanted this in OROM/SVG. In this context, a
+  ``rectangle'' is \emph{seen as} four points: top-left, top-right,
+  bottom-left, bottom-right. Obviously this is not a \emph{minimal}
+  description, since given e.g.~the top-left and bottom-right, the other
+  two points can be inferred. But one way or another, to be able to drag
+  any of them, all four points must be present at \emph{some} level. This
+  alternative ontology was polyfilled in the form of a ``rect controls''
+  class that can be attached manually to any SVG
+  \texttt{\textless{}rect\textgreater{}}. The SVG \texttt{x} and
+  \texttt{y} are subscribed to the top-left Point; \texttt{width} and
+  \texttt{height} are subscribed to Rod lengths.
 
-Example from 3.1.1. mention: CSS animations?
+  Another example is to be found in the DOM's event listener model.
+  Conceptually, many ``events'' are in fact changes to the state of some
+  physical (or conceptual) device. Some of these genuinely are different.
+  But both keyboard keys and mouse buttons, for example, have two states
+  --- pressed, not pressed --- which ought to make them interchangeable to
+  an extent. This is essentially the logic behind ``key remapping'' in
+  many PC games. But the official ontology of the Web makes it nontrivial
+  to do this.
 
-``One size fits all'' rears its ugly head again.
+  To start with, the situation is modelled not as a time-varying value
+  changing, but instead as a sort of Cartesian product of event listeners.
+  Rather than, say, a \texttt{left\_mouse\_button} piece of live-state, we
+  get \texttt{onmousedown} and \texttt{onmouseup}. Thankfully we do not
+  have \texttt{onAdown}/\texttt{onAup}, \texttt{onBdown/onBup}, \ldots{}
+  all the way to \texttt{onZdown}/\texttt{onZup}, just
+  \texttt{onkeydown}/\texttt{onkeyup}. But this is just one of the many
+  possible ways to slice this 3-dimensional\footnote{Device (mouse,
+    keyboard), sub-device (button, key), state (up, down)} space.
 
-I want: top-left, top-right, bot-left, bot-right \ldots{}
+  In OROM/SVG I did not quite want to re-map keys, but I did often want to
+  have things follow the mouse when dragged. In order to do this, I
+  reified the mouse pointer and its position, letting me write
+  \texttt{subscribe(point.position,\ pointer.position)}. The Point has an
+  \texttt{is-considering-me?} Observable wrapping
+  \texttt{onmouseover}/\texttt{onmouseout}, and the LMB\footnote{left
+    mouse button} is reified into \texttt{left\_mouse\_button\_is\_down}
+  to be explicit. The aforementioned subscription is set up whenever
+  \texttt{is-considering-me?} and \texttt{left\_mouse\_button\_is\_down}
+  become true, and torn down otherwise. I tended to think of this in the
+  form ``subscribe to pointer \emph{only when} pointer is-considering-me
+  \emph{and} LMB is down'', but I could live with this as a future
+  polyfill in JavaScript.
 
-Instead of explaining ``the x y is the top left blah blah'' in English
-in the W3C spec, they should make these relationships \emph{machine
-readable} with information like:
+  It takes some frustration and experience to get used to the idea that
+  you have a right to polyfill in an alternative representation. Before I
+  came to this conclusion, I used to twist my head around translating my
+  intention into the \texttt{x,y,width,height} parameters, and
+  un-translating when reading back the code I had produced. Once you get
+  used to having to adapt your mental imagery to a single way of doing
+  things (i.e.~learning to code), it makes sense to simply expect to see
+  more of it --- especially when you know you are new, surrounded by
+  veterans who see no problem, etc. But nowadays I take the position that
+  these are simply \emph{widespread} failings of our way of doing things,
+  instead of \emph{our} failure to adapt to the way software is.
 
-A Rect is
+  What would it look like to support multiple ontologies? There are the
+  two answers: anticipate the possibilities ahead of time, or support
+  users adding their own. The latter seems to imply polyfilling. As for
+  the former: in general, anticipating the diversity of ways someone might
+  look at the world is doomed to fail. But in the case of fairly
+  \emph{formalised} concepts such as geometrical shapes or mathematics, I
+  could suggest that simple \emph{under-specification} is the root of the
+  problem in SVG. Instead of explaining in English that ``\texttt{x} and
+  \texttt{y} are the co-ordinates of the top left-hand corner\ldots{}'',
+  it might be better to make these relations machine-readable or
+  \emph{embodied} in the API. For the rect this might look like:
 
-\begin{itemize}
-\tightlist
-\item
-  a Polygon
-\item
-  defined by 4 degrees of freedom x y width height
-\item
-  where the four vertices are
+  A \texttt{rect} is\ldots{} - a \texttt{polygon} (defined elsewhere) -
+  defined by 4 \emph{degrees of freedom} \texttt{x\ y\ width\ height}
+  (internal representation) - where there are 4 \emph{vertices}, all
+  \texttt{point}s + \texttt{{[}x,y{]}} ``top-left'' +
+  \texttt{{[}x+width,y{]}} ``top-right'' + \texttt{{[}x,y+height{]}}
+  ``bot-left'' + \texttt{{[}x+width,y+height{]}} ``bot-right''
 
-  \begin{itemize}
-  \tightlist
-  \item
-    \texttt{{[}x,y{]}} ``top-left''
-  \item
-    \texttt{{[}x+width,y{]}} ``top-right''
-  \item
-    etc
-  \end{itemize}
-\end{itemize}
+  The hope is that if we then specify enough information --- say, the
+  bot-left and top-right --- then the runtime has enough information to
+  derive its internal 4 degrees of freedom. In a crude sense I have
+  successfully anticipated the most obvious ontologies of a rectangle
+  here. However, I missed out the ``centre plus half-width and
+  half-height'' formulation, among many others.
 
-Reference to paper related to Antranig involving knitting\ldots{}?
-
-\ldots{}
+  Is this a futile effort even for precise mathematical knowledge
+  structures, or could enough formalisation of Euclidean geometry in the
+  Web platform put an end to this sort of polyfilling?\footnote{This
+    relates to my belief that type systems and other auto-reasoning tools
+    are trapped in a doomed quest for ``closed-form'' AI, representable as
+    a LaTeX formula.}
 
   \hypertarget{extensional-functions}{%
-\paragraph{Extensional Functions}\label{extensional-functions}}
+  \subsubsection{Extensional Functions}\label{extensional-functions}}
 
-aka dicts, Maps, std::vector, \ldots{}
+  Time and time again we come across the same pattern of partitioning
+  system state: trees or graphs of dictonaries, a.k.a. Maps, a.k.a.
+  associative arrays. I am talking about filesystem paths
+  \texttt{/path/to/some/file}, Python / Java modules
+  \texttt{com.example.pkg.subpkg}, JavaScript objects
+  \texttt{window.my\_obj.component}. The common case is an association of
+  a textual (string) name, to an arbitrary value. I find it useful to see
+  this as a mathematical ``function'' defined \emph{extensionally} by
+  listing its input/output mappings -- this opposed to an
+  \emph{intensional} definition such as
+  \texttt{x\ \textbackslash{}mapsto\ 2x+3}, or a computer program.
 
-\ldots{} \ldots{} Some parts of the OROM authors' C code were confusing
+  Extensional functions are perhaps the most basic form of Knowledge
+  Representation, and match natural language very well. ``The bicycle's
+  wheels' spokes are silver'' straightforwardly translates to a function
+  equation
+  \texttt{root\ (bicycle)\ (wheels)\ (spokes)\ (colour)\ =\ root\ (silver)}.
+  That is, whatever object is the output of \texttt{silver} in the
+  top-level \texttt{root} function, the output of \texttt{colour} (in the
+  function on the left) points to the same object. The ability to
+  partition a system in this way enables what
+  (2){[}https://github.com/amb26/papers/blob/master/ppig-2016a/ppig2016a.pdf{]}
+  calls a ``natural co-ordinate system'' for a piece of software, crucial
+  for understanding and adaptability by others (so long as it is
+  externalised).
+
+  It seems that this way of expressing the ``parts'' of a system is an
+  inevitable requirement of any programming substrate. Some languages,
+  such as C, do have static, \emph{compile-time} associative arrays
+  (\texttt{struct}s). In my experience this is not enough, and it's
+  necessary to bring in a library or clutter the code with a home-grown
+  approximation. Some parts of the OROM authors' C code were confusing
   until I realised they were just the guts of a basic associative-array
-implementation. \ldots{} \ldots{}
+  implementation; when I switched to JavaScript, these lines vanished.
+  Perhaps another strength of JavaScript is its low concrete syntax cost
+  for \emph{instances} or \emph{literals} of associative arrays. Writing
+  or reading
 
-Inevitability / ubiquity of state descriptions that are at trees of
-strings with pointers
+  \begin{lstlisting}[language=JavaScript]
+  a = {
+    b1: { c1: z, c2: y },
+    b2: { c3: x, c4: w }
+  };
+  \end{lstlisting}
 
-Antranig ``natural co-ordinates''?
+  is more WYSIWYG\footnote{What You See Is What You Get} than the
+  imperative-style
 
-OROM/SVG as Boxer-like extensional functions
+  \begin{lstlisting}[language=JavaScript]
+  a = new Map();
+  a.set('b1', new Map());
+  a.get('b1').set('c1', z);
+  a.get('b1').set('c2', y);
+  a.set('b2', new Map());
+  a.get('b2').set('c3', x);
+  a.get('b2').set('c4', w);
+  \end{lstlisting}
+
+  The latter is unfortunately still required in JavaScript for extensional
+  functions with \emph{non-string} inputs.
+
+  The decision to switch from the flat tabular representation in OROM/HTML
+  to the nestable, Boxer-like structure in OROM/SVG permitted extensional
+  (tree) functions in its substrate. True, I could always encode these in
+  flat tables by having them reference each other. But this is like the
+  second listing above: hard to follow. In the case of distinguishing
+  between ordinary state mappings and the ``method dictionary'' of
+  vtables, I actually stored method mappings with a ``-'' character to
+  avoid doing this.
+
+  (Screenshot showing this)
+
+  In OROM/SVG, I can directly express the model I was thinking of: the
+  vtable box has a \texttt{methods} box and that's where the method boxes
+  go.
+
+  (Screenshot showing this)
 
   \hypertarget{externalisability}{%
-\paragraph{Externalisability}\label{externalisability}}
+  \subsubsection{Externalisability}\label{externalisability}}
+
+  This refers to exposing structured program state to the user. This data
+  can then be saved and used to restore the system at a later date, but it
+  can also be tweaked with corresponding changes reflected in the system.
+
+  Externalisability, or its weaker cousin \emph{persistence}, was
+  absolutely necessary to continue OROM/SVG development past a certain
+  point.
 
@@
   \hypertarget{polyfilling}{%
-\paragraph{Polyfilling}\label{polyfilling}}
+  \subsubsection{Polyfilling}\label{polyfilling}}
 
@@
 
+  objects in JS user data
+
   \hypertarget{the-orom-system-as-a-part-of-the-solution}{%
-\subsection{The OROM system as a part of the
+  \section{The OROM system as a part of the
   solution}\label{the-orom-system-as-a-part-of-the-solution}}
@@
   \hypertarget{tensions-between-philosophies}{%
-\subsubsection{Tensions between
+  \subsection{Tensions between
   philosophies}\label{tensions-between-philosophies}}
@@
   \hypertarget{stable-or-emergent-antranig-against-oop-dispatch}{%
-\paragraph{Stable or emergent? (Antranig against OOP
+  \subsubsection{Stable or emergent? (Antranig against OOP
   dispatch)}\label{stable-or-emergent-antranig-against-oop-dispatch}}
@@
   \hypertarget{turing-complete-antranig-against-this}{%
-\paragraph{Turing-complete? (Antranig against
+  \subsubsection{Turing-complete? (Antranig against
   this)}\label{turing-complete-antranig-against-this}}
@@
   \hypertarget{messaging-vs-read-write}{%
-\paragraph{Messaging vs Read-Write?}\label{messaging-vs-read-write}}
+  \subsubsection{Messaging vs Read-Write?}\label{messaging-vs-read-write}}
 

### 2020-01-21 convivial-2020/manuscript.tex
Polish images, rework ... ETC!

--- 2020-01-15 convivial-2020/tex/manuscript.tex
@@
-\documentclass[english,submission]{programming}
-\usepackage[backend=biber]{biblatex} % Use Biblatex
-\addbibresource{example.bib}
-
-\lstdefinelanguage{JavaScript}{
-  keywords={break, case, catch, continue, debugger, default, delete, do, else,
-  finally, for, function, if, in, instanceof, new, return, switch, this, throw,
-  try, typeof, var, void, while, with, let},
-  morecomment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  morestring=[b]',
-  morestring=[b]",
-  sensitive=true
-}
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\begin{document}
-
-\paperdetails{
-  perspective=engineering,
-  area={General-purpose programming}
-}
-
 \title{What does it take to create with domain-appropriate tools?}
@@
 
-\keywords{paper, showcase, lorem ipsum}
-
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10002944.10011122.10003459</concept_id>
-<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}`
-
 \maketitle
 
+\newcommand{\joel}[1]{}
+\newcommand{\svgel}[1]{\texttt{\textless{}#1\textgreater{}}}
+\newcommand{\OSFA}{One-Size-Fits-All}
+
 \begin{abstract}
-  This paper shows...
+There is a One-Size-Fits-Allquality to languages, APIs and even programming itself.
+Whether you're making a mobile game or a scientific simulation, you will be using a text-based language with similar devices for structuring your code.
+This is a source of artificial difficulty in creating, understanding, and modifying software systems.
+No matter the domain, the author's design needs encoding into a form that does not resemble it.
+
+This paper describes a vision where software can be built in a programming environment that is closer to the domain of the software itself.
+By doing so, users of the system can use familiar abstractions and tools for adapting it.
+A step towards this vision is presented: a Web version of a minimal OOP system, developed as an executable version of the diagrams of its design, in a substrate meant to facilitate this.
+The experience of creating such a substrate is analysed, and I suggest deficiencies in programming environments that stand in the way of making this practice commonplace, as well as ways to fill in these gaps.
 \end{abstract}
+
+\keywords{substrate, visual, context, domain, adaptation}
 
@@
 speaking in unqualified terms of ``better'' and ``worse'' as if there
-  really is a one-size-fits-all solution to software creation!
+really is a One-Size-Fits-Allsolution to software creation!
 
@@
 text-imperialism with anything-else-imperialism, which is one
-  interpretation of calls for alternatives. This dream goes beyond
-  familiar senses of what constitutes a ``craft'', as far as a strong
-  melding of tool and material.
-
-  Certainly in the real world there is only one ``material'' --- matter
-  --- and humanity as a whole does create ``higher-order'' tools. A
-  (future) software craft, however, could be expected to give this power
-  to \emph{individuals} instead of the community alone. The
-  one-size-fits-all approach is an understandable consequence of few
-  designers catering to many users --- perhaps due to the necessity to
-  specialise in topics, languages, libraries, and tools. Their complexity
-  is part of the general difficulty of software creation, exacerbated by
-  the one-size-fits-all approach, closing the circle.
+interpretation of calls for alternatives.
+
+This dream goes beyond the familiar sense of what constitutes a
+``craft'', as far as a strong melding of tool and material. Parallels
+can be drawn with industrialisation and a strong division of labour: the
+community as a whole produces its higher-order tools, but currently no
+single person can have the same autonomy. A (future) software craft
+could be expected to give this power to \emph{individuals}, instead of
+the community alone. Whenever there are many small specialities
+(e.g.~languages, tools, or subject areas) each serving many clients, the
+One-Size-Fits-Allstyle is the best one can hope for. Adaptation to
+individual preferences and idiosyncrasies is only feasible when those
+individuals can do it themselves.
 
@@
 different ``way of creating''. Existing systems seem to only have one of
-  these properties without the other. Smalltalk and LISP try to minimise
-  arbitrary commitments of language \emph{semantics} to be adaptable, but
+these properties without the other: Smalltalk and LISP try to minimise
+arbitrary commitments of language \emph{semantics} to this end, but
 their being textual languages is a fairly tough commitment to break out
@@
 
-  Again, if someone wants to type out pictures in ASCII, let them ---
-  whether they do it for a challenge, or even if they find that more
-  natural for themselves. But equally, if I want to do it another way,
-  then please give me that affordance. This is how I segue into the
-  software artefact for which I have been attempting to build a natural
-  representation. True, it is a programmer's artefact, but it is still
-  representative of what any normal person has to do, insofar as:
+If someone wants to type out pictures in ASCII, let them --- whether
+they do it for a challenge, or even if they find that more natural for
+themselves. But equally, if I want to do it another way, then please
+give me that affordance. This is how I segue into the software artefact
+for which I have been attempting to build a natural representation.
+True, it is a programmer's artefact, but it is still representative of
+what any normal person has to do, insofar as:
 
@@
 
-  In summary, an object is a key-value mappings of state which can change
-  as a result of messages received by it. A message is sent by first
-  \emph{bind}-ing its name to its implementation, i.e.~specific code which
-  is then run in the context of the receiver. ``Bind'' is accomplished via
-  another message send, this time to the receiver's \emph{vtable}, or
-  ``behaviour''. This is another object whose state defines a particular
-  way of associating ``message name'' to ``implementation code''. This
-  continues up the vtable-chain, and terminates at a base case. The system
-  as a whole is ``bootstrapped'' into existence by its initialisation
-  code. Each step of this code makes use of any parts of the system set up
-  by the previous steps.
+Essentially, an OROM object is a block of state which can change as a
+result of messages received by it. A message is sent by first
+\emph{bind}-ing its name to its \emph{implementation}: specific code,
+which is then run in the context of the receiver. ``Bind'' is
+accomplished via another message send, this time to the receiver's
+\emph{vtable}, or ``behaviour''. This is another object whose state
+defines a particular way of associating ``message name'' to
+``implementation code''. This continues up the vtable-chain, and
+terminates at a base case. The system as a whole is ``bootstrapped''
+into existence by its initialisation code. Each step of this code makes
+use of any parts of the system set up by the previous steps.
 
@@
 not necessarily representative of those who read it. The
-  ``one-size-fits-all'' approach is perhaps unavoidable for static media
-  like print.
+One-Size-Fits-Allapproach is perhaps unavoidable for static media like
+print.
 
@@
 
-  (OROM/HTML screenshot)
-
-  I was grateful for the browser's management of graphical layout,
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{../orom-html.png}
+  \caption{OROM/HTML: Obj-dicts are rendered as resizable tables, which reference
+           each other through numerical IDs. \label{fig:orom-html}}
+\end{figure}
+
+To emphasise the tendency of OROM objects to be visualised as key-value
+mappings, I will refer to them as \emph{obj-dicts}. Figure
+\ref{fig:orom-html} shows the OROM/HTML implementation, in which they
+take the form of HTML tables.
+
+Here, I was grateful for the browser's management of graphical layout,
 resizable text fields, and keeping the DOM tree synchronised with what
-  you actually see. This last property enabled me to make the decision to
+one sees. This last property enabled me to make the decision to
 \emph{directly} encode much of the system state in the DOM, achieving
-  basic liveness (the thing on the screen is the actual thing) for the
-  keys and values of obj-dicts. For each obj-dict I used a two-column
+basic liveness\footnote{``The thing on the screen is the actual thing''}
+for the keys and values of obj-dicts.
+
+For each obj-dict I used a two-column
 \texttt{\textless{}table\textgreater{}} within a
@@
 
-  \hypertarget{orom-as-files-and-directories}{%
-  \subsection{OROM as files and
-  directories}\label{orom-as-files-and-directories}}
-
-  \begin{quote}
-  I would probably put this after the SVG part or perhaps even later once
-  you discuss more general principles (like the idea of polyfilling to
-  make a domain substrate into a programmable system). Then you can give
-  this as another example.
-  \end{quote}
+\hypertarget{practicality-vs.-high-aesthetics}{%
+\subsection{Practicality vs.~High
+Aesthetics}\label{practicality-vs.-high-aesthetics}}
+
+There is the view that insists on provably-correct code written with
+strange Unicode symbols, and at the opposite end there is the attitude
+of ``get the job done'', high style be damned. Each has its merits, but
+for my task I quickly saw the latter route to be most suitable.
+
+My favourite example of this has been the ``encoding'' of obj-dicts as
+DOM trees containing particular arrangements of children. In OROM/HTML,
+I would gleefully get/set state by calling code like this:
+
+\begin{lstlisting}[language=JavaScript]
+Entity.prototype.addState = function(key) {
+  let tr = document.createElement('tr');
+  tr.setAttribute('class', key);
+
+  let td = document.createElement('td');
+  td.setAttribute('class', 'key');
+  td.textContent = key;
+  tr.appendChild(td);
+  ...
+\end{lstlisting}
+
+and in OROM/SVG the story is similar. It takes advantage of the powerful
+developer tools included in browsers (e.g.~the DOM inspector), and
+avoids the need to manually synchronise separate Model and View for such
+encoded state. With a little work, it also allowed the system to be
+saved and restored --- see Section \ref{persistence}. I consider this as
+practical, expedient, ``hijacking'' of these substrates because it
+either goes against their intended use (HTML tables for presentation
+rather than semantics) or are simply ``odd'' (using certain arrangements
+of shapes to ``encode'' machine-readable semantics).
+
+Contrast this with an approach that had to re-synthesise bits of the DOM
+whenever something changed. Admittedly this does still happen to a
+certain extent, but at least it was minimised where obvious.
+
+Another example of practicality was in the issue of positioning and
+sizing. In Section \ref{nut-cracking-with-sledgehammers} I mentioned
+that I was able to sidestep some layout work by offloading it to the
+user. While building the system I tended to want to draw boxes into
+existence, and for that my brain's existing aesthetic algorithms can
+decide where they go. There are areas where the machine must be able to
+figure this out itself --- I don't want to be constantly bothered by
+prompts to place boxes as they are allocated by some running code ---
+but for one-offs it is a very convenient way to do a little less
+yak-shearing.
 
@@
 
-  (Screenshot of OROM/SVG showing boxes, code and arrows)
-
-  In this version, obj-dicts are encoded as nested SVG
-  \texttt{\textless{}rect\textgreater{}}s and other elements, reminiscent
-  of diSessa's Boxer. This was a significant departure from the table
-  representation, and even though SVG supports (some) nested HTML via
-  \texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
-  the possibility of multiple levels of nesting.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{../orom-svg.png}
+  \caption{OROM/SVG: Obj-dicts are moveable and resizable nested boxes,
+           referencing each other via real arrows.\label{fig:orom-svg}}
+\end{figure}
+
+In this version, shown in Figure \ref{fig:orom-svg}, obj-dicts are
+encoded as nested SVG \texttt{\textless{}rect\textgreater{}}s and other
+elements, reminiscent of diSessa's Boxer. This was a significant
+departure from the table representation, and even though SVG supports
+(some) nested HTML via \texttt{\textless{}foreignObject\textgreater{}},
+I actually preferred the possibility of multiple levels of nesting.
 
@@
 
-  \begin{quote}
-  Why is OROM/SVG more challenging? I think it would be useful to sketch
-  here, because it reveals some useful principles. There is more
-  polyfilling (arrows and resizing is not in the SVG substrate itself),
-  but at least it's possible to add those (unlike in HTML)
-  \end{quote}
-
-  \hypertarget{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}{%
-  \section{Inevitable Requirements Of Most Software (and the work we
-  must do to meet
-  them)}\label{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}}
-
-  As programmers, we have a maxim along the lines of: when you find
-  yourself repeating the same thing over and over again, factor it into a
-  ``thing'' and let the computer do the duplication. Some consider the
-  ``design pattern'' to be simply what a ``language feature'' looks like
-  when it's not part of the syntax.
-
-  \begin{quote}
-  The jump from very general abstraction idea to design patterns and
-  language features in the above is not exactly clear - it would be good
-  to help the reader understand why are you talking about this here.
-  \end{quote}
-
-  As I developed OROM/SVG I found myself implementing many such patterns.
-  It seems that languages are often at the wrong level of abstraction for
-  the requirements of modern software, necessitating the same boilerplate
-  per project just to get up to basic functionality. This burden either
-  falls on the author, or on the wider community to build and maintain
-  higher-level frameworks, syntax extensions, etc.
-
-  By contrast, these languages seem much better adapted to e.g.~batch-mode
-  printf()-centric applications. That they fail for the \emph{common case}
-  should be concerning. I will now present what seem to be the inevitable
-  demands of this common case.
-
-  \begin{quote}
-  I'm not sure I understand what the above opening is trying to say - I
-  suppose it might be because it is quite high-level/general. What is this
-  section about more specifically? Is it about what certain kind of
-  software consists of and what we need to build that? I would also be a
-  bit careful about things like ``most modern software'' below. It would
-  be good to instead characterize what kind of software are you interested
-  in - end-user graphical systems? (a microservice or something like that
-  does not need SVG\ldots{})
-  \end{quote}
-
-  \hypertarget{as-a-mere-consumer}{%
-  \subsection{As A ``Mere Consumer''}\label{as-a-mere-consumer}}
-
-  I begin by describing expectations of software that even ``end-users''
-  hold, consigned as they are to more or less accept what we give them.
+\hypertarget{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}{%
+\section{Typical Requirements Of Common Software (and the work we must
+do to meet
+them)}\label{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}}
+
+The ``common case'' of software present in our daily lives shares
+certain properties, such as being graphical and interactive. Such are
+expectations that ``end-users'' hold, consigned as they are to merely
+\emph{consume} what programmers give them. But if we want to make
+``programming'' more like ``using'', such expectations on \emph{use} at
+least need to be acknowledged. In this section, I will present these and
+other seemingly \emph{inevitable} demands of normal software,
+exemplified by OROM/SVG. I will comment how our programming platforms
+measure up to the task of helping us meet them, including my chosen
+platform of JavaScript and Web technologies.
 
 \hypertarget{retained-mode-vector-graphics}{%
-  \subsubsection{Retained-Mode Vector
+\subsection{Retained-Mode Vector
 Graphics}\label{retained-mode-vector-graphics}}
@@
 vector graphics editor --- just one with no GUI. I will return to this
-  later.
+at the end (Section ??).
 
@@
 rendering loop. Alternatively, there is the W3C's chosen ontology of CSS
-  animations, but see section ??.
+animations, but see Section \ref{context-appropriate-ontologies}.
 
 \hypertarget{basic-assumptions-about-physical-objects}{%
-  \subsubsection{Basic Assumptions About Physical
+\subsection{Basic Assumptions About Physical
 Objects}\label{basic-assumptions-about-physical-objects}}
@@
 instead of conceiving physics as an inherently complex thing to be found
-  only in specialised simulations.
-
-  \begin{quote}
-  People might not find this easy to believe - but I think you can
-  convince them by saying that pretty much all mobile phone apps have some
-  sort of ``phone touch physics'' (with sliding, opening menus etc.).
-  \end{quote}
+only in specialised simulations. For example, pretty much all mobile
+apps have what could be called ``phone touch physics'' where menus and
+screens slide in and out.
 
@@
 The level of physics in software tends to not involve force, or mass, or
-  very much at all, merely position and space. One thing that all usable
-  software must do, for example, is not crush many visually complex
-  shapes, such as lines of text, into the same region, since it becomes
-  unreadable. Such concepts of ``solid objects do not intersect'' or
-  ``only things at different layers may overlap'' are basic rules
-  inherited from the real world of graphical presentation.
+very much at all, merely position and space; we could call it
+``geometric physics''. One thing that all usable software must do, for
+example, is not crush many visually complex shapes, such as lines of
+text, into the same region, since it becomes unreadable. Such concepts
+of ``solid objects do not intersect'' or ``only things at different
+layers may overlap'' are basic rules inherited from the real world of
+graphical presentation.
 
@@
 \hypertarget{translationally-rigid-bodies}{%
-  \subparagraph{Translationally Rigid
+\subsubsection{Translationally Rigid
 Bodies}\label{translationally-rigid-bodies}}
@@
 \hypertarget{maintaining-relationships-over-time}{%
-  \subsubsection{Maintaining Relationships Over
+\subsection{Maintaining Relationships Over
 Time}\label{maintaining-relationships-over-time}}
@@
 of live state, then this must form part of the standard boilerplate.
-
 Such was the case in OROM/SVG. I implemented the system in an OOP
@@
 current value and a list of subscribers, notifying them when it changes.
-  It can be seen that this supplanted JavaScript's own ``dumb'' state by
-  the frequency of the change() function throughout the code.
 
@@
 
-  \begin{verbatim}
+\begin{lstlisting}[language=JavaScript, numbers=none]
 subscribe(object.position, pointer.position);
-  \end{verbatim}
+\end{lstlisting}
 
@@
 \hypertarget{nut-cracking-with-sledgehammers}{%
-  \subsubsection{Nut-Cracking With
+\subsection{Nut-Cracking With
 Sledgehammers}\label{nut-cracking-with-sledgehammers}}
@@
 The vector graphics I use in OROM/SVG are just rects, lines, circles and
-  text; a fraction of the full capability of SVG. The ``basic physics'' I
-  use is dwarfed by fully general 2D or 3D physics engines. The only
-  layout algorithm I had the patience to implement was a simple way to
-  expand a list of boxes to fit in a new child at the bottom; the
+text; a fraction of the full capability of SVG. The ``geometric
+physics'' I use is dwarfed by fully general 2D or 3D physics engines.
+The only layout algorithm I had the patience to implement was a simple
+way to expand a list of boxes to fit in a new child at the bottom; the
 affordance to place and size boxes \emph{manually} staves off the rest
@@
 The inevitable requirements I suggest here, do \emph{not} necessitate
-  Fully General anything. In fact, such representations would make it more
-  cumbersome to express what I wanted in OROM. As the old wisdom goes,
-  there's no point expressing a simple regex search as an arbitrary Turing
-  machine; my boxes don't \emph{have} a moment of inertia or a mass and I
-  don't \emph{need} a linear optimisation solver for my space management
-  --- for the time being.
-
-  Unfortunately, there doesn't seem to be much interest on the
-  smaller-scale instances of these problems. I suggest the name
-  ``geometric physics'' for the subset of physics I have mentioned.
-
-  \begin{quote}
-  I think the above section talks about 3 things - (i) it describes some
-  of the things that you had to implement for OROM/SVG, (ii) it talks
-  about some general patterns that arise there (like shapes and physics)
-  and (iii) it mentions some implementation details. I found it a bit
-  confusing to have these all mixed together. It reads a bit as if you
-  were jumping from one random topic that you encountered when
-  implementing this - but that makes the section very specific to your
-  implementation. How can we make it clear that there are more general
-  principles here? I think it might be better if you started with some
-  larger example (hypothetical is fine!) to illustrate what kind of things
-  you'd like to build - e.g. imagine you're building PowerPoint in OROM -
-  what would be some of the things you'd need? (for both the end
-  application and the system that you're using to build it). This would
-  then nicely motivate the more specific topics (graphics, physics,
-  relationships).
-  \end{quote}
-
-  \hypertarget{as-a-developer}{%
-  \subsection{As A ``Developer''}\label{as-a-developer}}
-
-  I now turn to some things that were perhaps inevitable, in any case
-  useful, as the \emph{implementor} of OROM/SVG.
-
-  \hypertarget{practicality-vs.-high-aesthetics}{%
-  \subsubsection{Practicality vs.~High
-  Aesthetics}\label{practicality-vs.-high-aesthetics}}
-
-  There is the view that insists on provably-correct code written with
-  strange Unicode symbols, and at the opposite end there is the attitude
-  of ``get the job done'', high style be damned. Each has its merits, but
-  for my task I quickly saw the latter route to be most suitable.
-
-  My favourite example of this has been the ``encoding'' of obj-dicts as
-  DOM trees containing particular arrangements of children. In OROM/HTML,
-  I would gleefully get/set state by calling code like this:
-
-  \begin{verbatim}
-  Entity.prototype.addState = function(key) {
-    let tr = document.createElement('tr');
-    tr.setAttribute('class', key);
-
-    let td = document.createElement('td');
-    td.setAttribute('class', 'key');
-    td.textContent = key;
-    tr.appendChild(td);
-    ...
-  \end{verbatim}
-
-  and in OROM/SVG the story is similar. It takes advantage of the powerful
-  developer tools included in browsers (e.g.~the DOM inspector), and
-  avoids the need to manually synchronise separate Model and View for such
-  encoded state. With a little work, it also allowed the system a large
-  degree of externalisability --- see Section ??. I consider this as
-  practical, expedient, ``hijacking'' of these substrates because it
-  either goes against their intended use (HTML tables for presentation
-  rather than semantics) or are simply ``odd'' (using certain arrangements
-  of shapes to ``encode'' machine-readable semantics).
-
-  Contrast this with an approach that had to re-synthesise bits of the DOM
-  whenever something changed. Admittedly this does still happen to a
-  certain extent, but at least it was minimised where obvious.
-
-  Another example of practicality was in the issue of positioning and
-  sizing. In Section ?? I mentioned that I was able to sidestep some
-  layout work by offloading it to the user. While building the system I
-  tended to want to draw boxes into existence, and for that my brain's
-  existing aesthetic algorithms can decide where they go. There are areas
-  where the machine must be able to figure this out itself --- I don't
-  want to be constantly bothered by prompts to place boxes as they are
-  allocated by some running code --- but for one-offs it is a very
-  convenient way to do a little less yak-shearing.
-
-  \begin{quote}
-  I think you could adapt the above discussion a bit and focus less on
-  what you don't want (high modernism) and more on what you want - what is
-  that? I think it's simplifying the problem along a different axis - to a
-  certain domain with relatively simple interactions and aesthetics. (So
-  rather than trying to make minimal formal calculus that's provably
-  correct, you're trying to make minimal working interaction model that's
-  interesting)
-  \end{quote}
+Fully General anything. In fact, such representations would make it
+\emph{more} cumbersome to express what I wanted in OROM. As the old
+wisdom goes, there's no point expressing a simple regex search as an
+arbitrary Turing machine; my boxes don't \emph{have} a moment of inertia
+or a mass and I don't \emph{need} a linear optimisation solver for my
+space management --- for the time being. Under the theme of
+domain-appropriate tools, it is worth designing interfaces for
+smaller-scale instances of these areas and exploring their capabilities.
+
+\hypertarget{patterns-and-polyfilling}{%
+\section{Patterns and Polyfilling}\label{patterns-and-polyfilling}}
+
+The message of the previous section is that existing platforms are often
+at the wrong level of abstraction for the requirements of common
+software. I recognise that they are reasonably well-adapted to batch
+mode file I/O tasks, but that they fail for the \emph{common case} is a
+problem. It necessitates largely the same ``boilerplate'' setup per
+project just to get basic functionality:
+
+\begin{itemize}
+\tightlist
+\item
+  Here is how I shall describe shapes
+\item
+  Here is how I make these shapes move together
+\item
+  Here is how I maintain internal consistency as the user unpredictably
+  changes things
+\end{itemize}
+
+This burden either falls on the author, or on the wider community to
+build and maintain higher-level frameworks, syntax extensions, etc. In
+this paper, I refer to this process as ``polyfilling'', with an emphasis
+on the DIY, individual level.
+
+It is true, we already have a term for bringing into existence some
+software feature that isn't already there: programming. The difference
+is that polyfilling is about filling in \emph{boilerplate},
+i.e.~functionality \emph{that should have been there in the first place,
+but isn't}. This injects some subjectivity and value judgement into the
+term. But ultimately, all platforms are designed with certain features
+``out of the box'' and leave other features to be implemented ``if
+needed''. And when a problem domain ``requires batteries'' yet the tools
+available say ``batteries not included'', it is quite reasonable to want
+the batteries to be part of the platform.
+
+This section covers the boilerplate I found myself polyfilling in
+OROM/SVG.
+
+Is to bring things into the mindset / ontology of the specific software.
+For example, wrapping the official Web conventions for event handling in
+the Observable infrastructure, or wrapping DOM nodes with ``user data''
+objects. The latter case, takes advantage of JavaScript's ability to
+simply ``add on'' new fields to objects, even objects of Web APIs, to
+``annotate'' the DOM node with a JavaScript helper. Very advantageous to
+permit this in JavaScript. In another language like C, such would
+require saving the address in a lookup table elsewhere.
 
 \hypertarget{positioning-and-sizing}{%
-  \subsubsection{Positioning and Sizing}\label{positioning-and-sizing}}
+\subsection{Positioning and Sizing}\label{positioning-and-sizing}}
 
 The simple desire to move and resize boxes with the mouse motivated a
-  lot of the live-state and geometric physics ideas. This problem could be
+lot of the concepts in Sections
+\ref{basic-assumptions-about-physical-objects} and
+\ref{maintaining-relationships-over-time}. This problem could be
 considered a microcosm of OROM/SVG: what's a natural way I
 \emph{conceive} of this behaviour, and could I implement it that way?
-
-  \begin{quote}
-  The above point (moving motivated live state and physics ideas) would be
-  a lot more obvious if you followed the structure with motivating example
-  first followed by an analysis\ldots{}
-  \end{quote}
 
@@
 endpoint. Since they both feel the same deltas, the displacement vector
-  between them is preserved.
-
-  (Diagram of delta transmission)
+between them is preserved (Figure \ref{fig:delta-transmission}).
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{../delta-transmission.png}
+  \caption{Fully rigid rods transmit changes in one end to the other end. \label{fig:delta-transmission}}
+\end{figure}
 
@@
 
-  Resizing (of boxes at least) could be achieved through rods that stay
-  horizontal or vertical. In the language of ``small differences'' spoken
-  by my live-state infrastructure, this is expressed as a rod
-  ``transmitting deltas'' in the vertical and horizontal, ``absorbing''
-  the other component into itself. Mirroring a DOM rect to these rods is
-  as simple as subscribing its width and height to horizontal and vertical
-  rods' length Observables. This way, boxes can be resized from whatever
-  corner is convenient.
-
-  (Box demonstrating all three rod types)
-
-  I colour such ``half-rigid'' rods yellow and fully rigid rods red. The
-  ``absorb all endpoint changes'' case is coloured green.
-
-  \begin{quote}
-  I don't think you explained what `half-rigid' means yet.
-  \end{quote}
+Resizing of boxes could be achieved through rods that stay horizontal or
+vertical. In the language of ``small differences'' spoken by the
+live-state infrastructure, this is expressed as a rod ``transmitting
+deltas'' in the vertical and horizontal, ``absorbing'' the other
+component into itself (Figure \ref{fig:rods}). Mirroring a DOM rect to
+these rods is as simple as subscribing its width and height to
+horizontal and vertical rods' \texttt{length} Observables. This way,
+boxes can be resized from whatever corner is convenient.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.75\textwidth]{../rods.png}
+  \caption{The yellow rods on the box border are ``half-rigid'': they transmit horizontal or vertical changes, but not both.
+The green rod in the top-left absorbs all changes from an endpoint into itself, visualising the displacement vector of the child box from its parent.
+The remaining red rods are fully rigid.
+A line of CSS can show this ``scaffolding'' which is normally hidden for users.
+\label{fig:rods}}
+\end{figure}
 
@@
 
-  \begin{quote}
-  This section reads more as an implementation detail - even though it
-  makes some interesting points about the physics \& rods system. However,
-  it might better to have it in a section that's more about implementation
-  than about general principles. Or perhaps you could restructure Section
-  3 to have (i) example (ii) general lessons (like physics etc.) and (iii)
-  specific points about implementation (e.g.~your concrete physics system
-  with rods, what you had to `polyfill')
-  \end{quote}
-
 \hypertarget{visible-coordinate-systems}{%
-  \subsubsection{Visible Coordinate
+\subsection{Visible Coordinate
 Systems}\label{visible-coordinate-systems}}
@@
 First of all, SVG shapes (e.g.~\texttt{\textless{}rect\textgreater{}})
-  are strictly leaf nodes, so if I wish to nest boxes within boxes, the
-  visible ``box'' must be a mere \emph{accessory} to the nestable element,
-  in my case a \texttt{\textless{}g\textgreater{}} (group). This means
-  that instead of resizing the x, y, width, height attributes of the
-  \texttt{\textless{}rect\textgreater{}}, only its width and height change
-  along with the transform attribute of its \emph{parent}
+are strictly \emph{leaf} nodes of the DOM. So if I wish to nest boxes
+within boxes, the visible box \texttt{\textless{}rect\textgreater{}}
+must be a mere \emph{accessory} to the nestable element, in my case a
+\texttt{\textless{}g\textgreater{}} (group). This means that instead of
+resizing the \texttt{x}, \texttt{y}, \texttt{width}, \texttt{height}
+attributes of the \texttt{\textless{}rect\textgreater{}}, only its
+\texttt{width} and \texttt{height} change, along with the
+\texttt{transform} attribute of its \emph{parent}
 \texttt{\textless{}g\textgreater{}}. This was not too bad; just
@@
 my early decision to have Point objects all share the global co-ordinate
-  system. This way, connecting ``this point'' to the mouse pointer, etc.
-  is not infuriating to express. Still, it was necessary in the case of
-  certain elements --- especially those which must transcend the tree
-  structure altogether, like arrows between boxes --- to bite this bullet,
-  one way or another.
+system (this was to ensure that simple relations, such as a point
+following the mouse pointer, are not infuriating to express.) Still, it
+was necessary in the case of certain elements --- especially those which
+must transcend the tree structure altogether, like arrows between boxes
+--- to bite this bullet, one way or another.
 
@@
 to children. It is nicer to express the relation (as well as see it!)
-  this way --- see Figure ?? above.
+this way (Figure \ref{fig:rods}).
 
 \hypertarget{context-appropriate-ontologies}{%
-  \subsubsection{Context-appropriate
+\subsection{Context-appropriate
 ontologies}\label{context-appropriate-ontologies}}
@@
 Each API has its own conventions, including a way of naming and
-  structuring expressions --- an ontology. The one-size-fits-all approach
+structuring expressions --- an ontology. The One-Size-Fits-Allapproach
 is exemplified in such interfaces. For example, in SVG we express a
 rectangle by
-  \texttt{\textless{}rect\ x="10"\ y="10"\ width="600"\ height="400"\textgreater{}}.
+\texttt{\textless{}rect x="10" y="10" width="600" height="400"\textgreater{}}.
 The SVG specification \emph{defines} to the user that a rect \emph{is} a
@@
 SVG-approved parametrisation of a rectangle is far from the only
-  one(1){[}https://www.shift-society.org/salon/papers/2018/critiques/critique-semprola.pdf{]},
+one\href{https://www.shift-society.org/salon/papers/2018/critiques/critique-semprola.pdf}{1},
 and thus is, unsurprisingly, ill-fitted to some contexts.
@@
 class that can be attached manually to any SVG
-  \texttt{\textless{}rect\textgreater{}}. The SVG \texttt{x} and
-  \texttt{y} are subscribed to the top-left Point; \texttt{width} and
-  \texttt{height} are subscribed to Rod lengths.
+\texttt{\textless{}rect\textgreater{}}. The \texttt{x} and \texttt{y}
+are subscribed to the top-left Point\footnote{In the usual case where
+  the \texttt{\textless{}rect\textgreater{}} is part of a box, the
+  top-left instead controls the parent
+  \texttt{\textless{}g\textgreater{}}'s \texttt{transform} --- but the
+  idea is the same.}; \texttt{width} and \texttt{height} are subscribed
+to Rod lengths.
 
@@
 Conceptually, many ``events'' are in fact changes to the state of some
-  physical (or conceptual) device. Some of these genuinely are different.
-  But both keyboard keys and mouse buttons, for example, have two states
-  --- pressed, not pressed --- which ought to make them interchangeable to
-  an extent. This is essentially the logic behind ``key remapping'' in
-  many PC games. But the official ontology of the Web makes it nontrivial
-  to do this.
+physical (or conceptual) device. And both keyboard keys and mouse
+buttons, for example, have two states --- pressed, not pressed --- which
+ought to make them \emph{interchangeable} to this extent. This is the
+logic behind ``key mapping''\footnote{For example, the player can
+  re-assign the action ``shoot'' from its default mouse button to a
+  keyboard key, or another action from keyboard to mouse, as it suits
+  them.} in PC games. But the official ontology of the Web makes it
+nontrivial to do this.
 
@@
 changing, but instead as a sort of Cartesian product of event listeners.
-  Rather than, say, a \texttt{left\_mouse\_button} piece of live-state, we
-  get \texttt{onmousedown} and \texttt{onmouseup}. Thankfully we do not
+Rather than, say, a piece of live-state for the left mouse button (LMB),
+we get \texttt{onmousedown} and \texttt{onmouseup}. Thankfully we do not
 have \texttt{onAdown}/\texttt{onAup}, \texttt{onBdown/onBup}, \ldots{}
-  all the way to \texttt{onZdown}/\texttt{onZup}, just
-  \texttt{onkeydown}/\texttt{onkeyup}. But this is just one of the many
+all the way to \texttt{onZdown}/\texttt{onZup}, but only
+\texttt{onkeydown}/\texttt{onkeyup}. Yet this is just one of the many
 possible ways to slice this 3-dimensional\footnote{Device (mouse,
@@
 \texttt{is-considering-me?} Observable wrapping
-  \texttt{onmouseover}/\texttt{onmouseout}, and the LMB\footnote{left
-    mouse button} is reified into \texttt{left\_mouse\_button\_is\_down}
-  to be explicit. The aforementioned subscription is set up whenever
+\texttt{onmouseover}/\texttt{onmouseout}, and the LMB is reified into
+\texttt{left\_mouse\_button\_is\_down} to be explicit. The
+aforementioned subscription is set up whenever
 \texttt{is-considering-me?} and \texttt{left\_mouse\_button\_is\_down}
@@
 
+The way these things are connected to the browser's event listeners
+could be called ``device drivers''. This approach is described in more
+detail in \href{https://prog21.dadgum.com/66.html}{13}. It amounts to
+translating information from the ``unchangeable'' Web platform into the
+worldview of my substrate, as early as possible:
+
+\begin{lstlisting}
+svg.onmousedown = e =>
+  if (e.button === 0) change(left_mouse_button_is_down, true);
+
+svg.onmouseup = e =>
+  if (e.button === 0) change(left_mouse_button_is_down, false);
+
+svg.onmousemove = e => {
+  let r = svg.getBoundingClientRect();
+  let pos = vsub([e.clientX, e.clientY], [r.left, r.top]);
+  change(pointer.position, pos);
+};
+\end{lstlisting}
+
 It takes some frustration and experience to get used to the idea that
-  you have a right to polyfill in an alternative representation. Before I
+you have a right to polyfill in alternative representations. Before I
 came to this conclusion, I used to twist my head around translating my
@@
 more of it --- especially when you know you are new, surrounded by
-  veterans who see no problem, etc. But nowadays I take the position that
-  these are simply \emph{widespread} failings of our way of doing things,
-  instead of \emph{our} failure to adapt to the way software is.
+veterans who see no problem, and so on. Nowadays, I take the position
+that these are simply \emph{widespread} failings of our way of doing
+things, instead of \emph{our} failure to adapt to the way software is.
 
@@
 two answers: anticipate the possibilities ahead of time, or support
-  users adding their own. The latter seems to imply polyfilling. As for
-  the former: in general, anticipating the diversity of ways someone might
-  look at the world is doomed to fail. But in the case of fairly
-  \emph{formalised} concepts such as geometrical shapes or mathematics, I
-  could suggest that simple \emph{under-specification} is the root of the
-  problem in SVG. Instead of explaining in English that ``\texttt{x} and
-  \texttt{y} are the co-ordinates of the top left-hand corner\ldots{}'',
-  it might be better to make these relations machine-readable or
-  \emph{embodied} in the API. For the rect this might look like:
-
-  A \texttt{rect} is\ldots{} - a \texttt{polygon} (defined elsewhere) -
+users adding their own. It is unclear if we can do better on the latter
+than simply ``support polyfilling''. As for the former: in general,
+anticipating the diversity of ways someone might look at the world is
+doomed to fail. But in the case of fairly \emph{formalised} concepts
+such as geometrical shapes or mathematics, I could suggest that simple
+\emph{under-specification} is the root of the problem in SVG. Instead of
+explaining in English that ``\texttt{x} and \texttt{y} are the
+co-ordinates of the top left-hand corner\ldots{}'', it might be better
+to make these relations machine-readable or \emph{embodied} in the API.
+For the rect this might look like:
+
+\noindent A \texttt{rect} is\ldots{}
+
+\begin{itemize}
+\tightlist
+\item
+  a \texttt{polygon} (defined elsewhere)
+\item
   defined by 4 \emph{degrees of freedom} \texttt{x\ y\ width\ height}
-  (internal representation) - where there are 4 \emph{vertices}, all
-  \texttt{point}s + \texttt{{[}x,y{]}} ``top-left'' +
-  \texttt{{[}x+width,y{]}} ``top-right'' + \texttt{{[}x,y+height{]}}
-  ``bot-left'' + \texttt{{[}x+width,y+height{]}} ``bot-right''
+  (internal representation)
+\item
+  where there are 4 \emph{vertices}, all \texttt{point}s
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \texttt{{[}x,y{]}} ``top-left''
+  \item
+    \texttt{{[}x+width,y{]}} ``top-right''
+  \item
+    \texttt{{[}x,y+height{]}} ``bot-left''
+  \item
+    \texttt{{[}x+width,y+height{]}} ``bot-right''
+  \end{itemize}
+\end{itemize}
 
 The hope is that if we then specify enough information --- say, the
-  bot-left and top-right --- then the runtime has enough information to
-  derive its internal 4 degrees of freedom. In a crude sense I have
-  successfully anticipated the most obvious ontologies of a rectangle
-  here. However, I missed out the ``centre plus half-width and
-  half-height'' formulation, among many others.
-
-  Is this a futile effort even for precise mathematical knowledge
-  structures, or could enough formalisation of Euclidean geometry in the
-  Web platform put an end to this sort of polyfilling?\footnote{This
-    relates to my belief that type systems and other auto-reasoning tools
-    are trapped in a doomed quest for ``closed-form'' AI, representable as
-    a LaTeX formula.}
+bot-left and top-right --- then the runtime has all it needs to derive
+its internal 4 degrees of freedom.
+
+In a crude sense, I have successfully anticipated the most obvious
+ontologies of a rectangle here. However, I missed out the ``centre plus
+half-width and half-height'' formulation, among others. Is this a futile
+effort even for precise mathematical knowledge structures, or could
+enough formalisation of Euclidean geometry in the Web platform put an
+end to this sort of polyfilling?\footnote{This relates to my intuition
+  that type systems and other auto-reasoning tools are trapped in a
+  doomed quest for ``closed-form'' AI, representable as a
+  \LaTeX formula.}
 
 \hypertarget{extensional-functions}{%
-  \subsubsection{Extensional Functions}\label{extensional-functions}}
+\subsection{Extensional Functions}\label{extensional-functions}}
 
@@
 this as a mathematical ``function'' defined \emph{extensionally} by
-  listing its input/output mappings -- this opposed to an
-  \emph{intensional} definition such as
-  \texttt{x\ \textbackslash{}mapsto\ 2x+3}, or a computer program.
+listing its input/output mappings --- this opposed to an
+\emph{intensional} definition such as \(x \mapsto 2x+3\), or a computer
+program.
 
@@
 partition a system in this way enables what
-  (2){[}https://github.com/amb26/papers/blob/master/ppig-2016a/ppig2016a.pdf{]}
+\href{https://github.com/amb26/papers/blob/master/ppig-2016a/ppig2016a.pdf}{2}
 calls a ``natural co-ordinate system'' for a piece of software, crucial
@@
 such as C, do have static, \emph{compile-time} associative arrays
-  (\texttt{struct}s). In my experience this is not enough, and it's
-  necessary to bring in a library or clutter the code with a home-grown
-  approximation. Some parts of the OROM authors' C code were confusing
-  until I realised they were just the guts of a basic associative-array
-  implementation; when I switched to JavaScript, these lines vanished.
-  Perhaps another strength of JavaScript is its low concrete syntax cost
-  for \emph{instances} or \emph{literals} of associative arrays. Writing
-  or reading
-
-  \begin{lstlisting}[language=JavaScript]
+(\texttt{struct}s). In my experience this is usually not enough, and
+it's necessary to bring in a library or clutter the code with a
+home-grown approximation to dynamic ones. Some parts of the OROM
+authors' C code were confusing until I realised they were just the guts
+of a basic associative-array implementation; when I switched to
+JavaScript, these lines vanished.
+
+Perhaps another strength of JavaScript is its low \emph{concrete syntax
+cost} for \emph{instances} or \emph{literals} of associative arrays.
+Writing or reading
+
+\begin{lstlisting}
 a = {
@@
 
-  \begin{lstlisting}[language=JavaScript]
+\begin{lstlisting}
 a = new Map();
@@
 vtables, I actually stored method mappings with a ``-'' character to
-  avoid doing this.
-
-  (Screenshot showing this)
-
-  In OROM/SVG, I can directly express the model I was thinking of: the
-  vtable box has a \texttt{methods} box and that's where the method boxes
-  go.
-
-  (Screenshot showing this)
-
-  \hypertarget{externalisability}{%
-  \subsubsection{Externalisability}\label{externalisability}}
+avoid doing this (Figure \ref{fig:method-hyphens}). In OROM/SVG, I can
+directly express the model I was thinking of (Figure
+\ref{fig:method-boxes}).
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.5\textwidth]{../method-hyphens.png}
+  \caption{OROM/HTML: use of hyphenated \texttt{-lookup} to signify a \emph{method} called
+           \texttt{lookup} instead of a property like \texttt{vtable}. \label{fig:method-hyphens}}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.5\textwidth]{../method-boxes.png}
+  \caption{OROM/SVG: the \texttt{vtable} box has a \texttt{methods} box, and that's where the method boxes go. \label{fig:method-boxes}}
+\end{figure}
+
+\hypertarget{persistence}{%
+\subsection{Persistence}\label{persistence}}
 
@@
 
-  Externalisability, or its weaker cousin \emph{persistence}, was
-  absolutely necessary to continue OROM/SVG development past a certain
-  point.
-
-  Reference Antranig's stuff on this. How I dealt with event listeners etc
-
-  Necessity to development: frustrating to find bug, tear down, refresh,
-  lose work, build it up again!
-
-  Inspect Element: copy to save works, arbitrary edits \emph{should} work
-  (cuz it's on demand)
-
-  Work required to parse SVG trees into objects ``on demand''
-
-  Started heavyweight, ended lightweight (generate rods on demand, let SVG
-  move children)
-
-  \hypertarget{polyfilling}{%
-  \subsubsection{Polyfilling}\label{polyfilling}}
-
-  Very advantageous to permit this e.g.~JavaScript
-
-  Stupid lack of XYZ in JS or SVG due to come out in the next spec, etc
-
-  Syntax sugar (SVG tree ops, vector ops)
-
-  objects in JS user data
+Persistence was absolutely necessary to continue OROM/SVG development,
+past a certain point. This is because, upon discovering a bug and fixing
+it in the source code, the web page must be refreshed and started anew.
+In the beginning, when verifying that box drawing with the mouse is
+working correctly, this is not much of a problem: upon refresh, the
+blank initial state is restored and I could draw again. But as the
+substrate matured, and I began to implement parts of the target system
+(OROM) --- the cycle of finding a bug, tearing down the system,
+refreshing and losing work, and manually building it up again, proved
+frustrating. Because the system could only be patched externally and
+restarted, there needed to be some way to persist changes that
+\emph{weren't} part of the code, but the DOM instead.
+
+Normally, this can be as simple as autosave to the filesystem. But the
+Web platform is very wary of this\footnote{As anyone who learns WebGL
+  can attest to, when they discover they must run a local Web server to
+  provide image files for textures since any filesystem requests will be
+  rejected for security.}, so the solution I turned to was manually
+copying the markup in the browser's inspector (Figure
+\ref{fig:html-inspector}).
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.8\textwidth]{../html-inspector.png}
+  \caption{To save the state of the system, the SVG group corresponding to the \texttt{root} box is copied and pasted inside the HTML file's \texttt{\textless{}svg\textgreater{}} element.\label{fig:html-inspector}}
+\end{figure}
+
+This required a slight change towards an architecture where the all the
+data required to reconstruct the system's current state is contained in
+the inspector HTML (plus the OROM/SVG JavaScript source code.) Where
+previously ``boxes'' were created first as invisible JavaScript objects
+responsible for some SVG, now it was the other way round. When a rect is
+clicked, the system must look at some SVG and interpret it ``on demand''
+as a box (lazily initialising its helper object if so).
+
+Persistence seems to be a weaker cousin of \textbf{externalisability},
+defined in \href{antranig}{1}. As it stands, the system does not quite
+qualify as externalisable. When making changes to the HTML in the
+element inspector, the system's behaviour \emph{should} adjust to match,
+but this is not currently guaranteed in all cases.
 
@@
 
-  Meta-circularity as empowerment, not just ``cool''
-
-  The allure
+The OROM system was designed with a goal of eliminating the
+``artificial'' distinction between implementation language and end-user
+language, by means of a mostly self-defining, or ``meta-circular''
+object model. This general way of working on a piece of software ``by
+means of itself'' is easy to agree with, even if OOP is not to
+everyone's taste. It is an open question whether OROM's approach could
+be applied to other programming styles.
+
+The allure of meta-circularity here is not merely in being ``cool'', but
+that it paves the way to end-user empowerment. This is somewhat
+misleading because in OROM, these ``end-users'' are actually
+programmers. However, its purpose is to free a programmer's dependence
+on some distant and busy language designer, and it seems plausible that
+this is a stepping stone on the way to enabling programmers to
+\emph{create} a piece of software which non-programmers do not depend on
+them for.
+
+I emphasise that this is the \emph{allure} of OROM. However, there are
+some significant practical issues that must be overcome or clarified
+first.
+
+\hypertarget{minimal-descriptions}{%
+\subsection{Minimal descriptions}\label{minimal-descriptions}}
+
+The OROM paper was part of the STEPS project of Alan Kay's
+VPRI\footnote{Viewpoints Research Institute}. This project argued that
+the immense level of ``accidental complexity'' present in software
+implementation could be reduced, and Kay himself dreams of an end result
+analogous to a ``Maxwell's Equations'' of software. That is: the
+behaviour of electromagnetic fields can be represented in four short
+equations that fit ``on a T-shirt''. A self-hosted LISP interpreter fits
+on a page. Could we aim at a similar ``fundamental description of
+software'' that fits on something less than millions of lines of code?
+
+This argument as stated suffers by glossing over an important fact.
+Maxwell's equations certainly do fit onto a T-shirt, but most people
+will not be able to explain what they mean. What typically amounts to
+years of study is compressed into those mathematical symbols, and the
+learning material involved most certainly does \emph{not} fit on a
+T-shirt. The obvious \emph{reductio ad absurdudum} is where we
+encapsulate these equations under a single symbol, M. M is defined as
+``Maxwell's Equations are true''. Ta-da --- this fits on a coin, but
+good luck doing anything with it.\footnote{In fact, this is almost
+  achieved by the formalism of Geometric Calculus, which reduces them to
+  only one equation.}
+
+I say this not to dismiss the argument, but to highlight the actually
+hard part of getting a ``concise description'' of some system; defining
+complexity away into a symbol helps us no more than naming the solution
+to an equation ``x''. There is a connection with data compression: even
+if the data successfully compressed into a smaller file, the size of the
+compression \emph{program} should be added as well. What matters is to
+reduce the \emph{combined} size of notation and platform.
+
+We have a word ``re-factoring'', but what about the ``factoring'' in the
+first place? The previous discussion of ``ontologies''
+
+Further, there is perhaps a risk of optimising for \emph{formal} rather
+than \emph{practical} minimality (e.g.~the Turing machine). OROM's
+``minimality'' does not necessarily translate into ``simplicity''; it
+suffers from the same \emph{cognitive} complexity, or need for study, as
+Maxwell's Equations. I cannot stress the amount of effort I have put in
+to wrap my head around the self-referential ``vtable vtable'' and the
+task of self-implementation. An overall better system might be one which
+is easier to pick up or understand, even if the number of formal objects
+is not as minimal.
+
+\hypertarget{self-implementation}{%
+\subsection{Self-implementation}\label{self-implementation}}
+
+This is also cool, but hard.
 
 Other themes e.g.~homoiconicity of code and data
-
-  Problems: cognitive vs.~``official'' complexity
-
-  Maxwell's eqns metaphor: omits the years of study compressed into the
-  math symbols!
-
-  And yet, the allure of the Maxwell's Eqns of Software?
 
@@
 
-  \hypertarget{misc}{%
-  \section{Misc}\label{misc}}
-
-  A note about DIY and libraries
-
-  My other problem was one of cost-benefit. My aim was to \ldots{}
-
-\acks
-I want to thank ....
-
-\printbibliography
-\end{document}
+\hypertarget{conclusion-and-future-work}{%
+\section{Conclusion and future work}\label{conclusion-and-future-work}}
+
+Conclusion: what I have that Maxwell's E doesn't.

### 2020-01-21 convivial-2020/tex/manuscript.tex
Save recent work...
Polish images, rework ... ETC!

--- 2020-01-15 convivial-2020/tex/manuscript.tex
@@
 
+\lstdefinelanguage{JavaScript}{
+  keywords={break, case, catch, continue, debugger, default, delete, do, else,
+  finally, for, function, if, in, instanceof, new, return, switch, this, throw,
+  try, typeof, var, void, while, with, let, false, true},
+  morecomment=[l]{//},
+  morecomment=[s]{/*}{*/},
+  morestring=[b]',
+  morestring=[b]",
+  sensitive=true
+}
+\lstset{language=JavaScript, numbers=none}
+
+\providecommand{\tightlist}{}% Don't want Pandoc's tight lists
+  %\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
 \begin{document}
@@
 \paperdetails{
-  perspective=engineering,
+  perspective=art,
   area={General-purpose programming}
 }
-
-\title{What does it take to create with domain-appropriate tools? A case study on the ``OROM'' system.}
-\author{Joel Jakubovic}
-\affiliation{University of Kent, Canterbury}
-
-\keywords{paper, showcase, lorem ipsum}
 
@@
 
-\maketitle
-
-\begin{abstract}
-  This paper shows...
-\end{abstract}
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-As someone who can code, I have already passed the first and most
-important hurdle for making full use of the potential of my computer.
-However, even in this supposedly empowered state, I am still far away
-from feeling the relationship between myself and software as between
-artisan and material, free to shape it into any form with effort
-proportional to complexity.
-
-One would have thought that software-creation acts like hypothetical
-super-intelligent AI. That is: even though we start from a primitive
-base in the 50s (or even today), there would surely be a recursive
-process of self-improvement, building better software-creation tools
-with the existing ones, until an ``expressivity singularity'' where
-software becomes a workable material as described.
-
-However, this didn't happen. Or at least, it is happening glacially
-slowly. The brute fact is that whenever you want to create software, you
-go to a text editor and figure out how to translate your design into
-that. The text editors, being software, were written with the help of
-previous text editors, and so on. It's undeniable that text editors have
-improved, even if you think it peaked with Emacs. We just don't seem
-able to go beyond them.
-
-\begin{quote}
-You discuss what's wrong with text editor in the next few paragraphs,
-but I think it would be useful to give some brief summary here so that
-readers don't think that ``we need to go beyond text editors'' is
-unmotivated - I think the summary of the text below is that you want to
-create software using tools that are closer to the domain of the
-software (text is far from everything; visual editors would be closer to
-a more visual domain) This general formulation might also nicely frame
-some of the things you say next.
-\end{quote}
-
-Amdahl's Law generalises the following idea: even when you spend hours
-of effort doubling the performance of a component used 1\% of the time,
-your reward is a system overall improved by a mere 0.5\%.
-
-Text coding is certainly ubiquitous, the 99\% case in programming, so
-you might wonder where I'm going with this. Well, a small improvement to
-text editing, if adopted by everyone, certainly does have a massive
-\emph{intermediate} effect -- but this only \emph{matters} to the extent
-that text was helping the programmers in the first place. If my goal is
-to draw or animate pictures, or create a digital synth from a frequency
-spectrogram, then giving me the ability to auto-indent my SVG markup is
-rather underwhelming as a productivity increase, as it doesn't target
-the core of the enterprise that makes it so hard.
-
-My experience of coding, most of the projects requiring shapes (such as
-GUIs), leads me to conclude that no matter how much I improve my skill
-at a particular language, knowledge of libraries or even general coding
-ability, my predicament stays the same. Our basic method of creating
-software is optimised for an ever-diminishing proportion of the software
-we actually want to make; ill-optimised for the graphics, layout,
-interactivity and and basic physics -- more on this later -- that we
-usually require.
-
-Whenever I work on these I feel stuck in a box I know I can never escape
-from: that box is the text editor, a fixed conduit through which all
-\emph{fundamental} changes to my program must pass. It's not a part of
-the system I am building, so I can't even make use of features of the
-thing I'm developing, to make its own development easier.
-
-Surely the trick is to \emph{use} coding to build something \emph{better
-than it}. And then use that, to build something even better. But there
-is an enormous breadth and depth of philosophies here, along with many
-failed historical attempts to do better -- or at least, ones that failed
-to catch on. But even worse than this is that in my very \emph{language}
-here I am making the same mistake as the text editor -- speaking in
-unqualified terms of ``better'' and ``worse'' as if there really is a
-one-size-fits-all solution to software creation!
-
-Of course what we \emph{really} want is the ability for people to create
-\emph{in the way that they think is best} in their particular context --
-to equip them to feasibly create the tools that suit them for the thing
-they want to make. And second-order tools that suit them for making the
-first-order tools, etc. It would do no good to replace text-imperialism
-with anything-else-imperialism, which is one common interpretation of
-calls for alternatives. If someone wants to type out pictures in ASCII,
-let them -- whether they do it for a challenge, or even if they find
-that more natural for themselves. But equally, if I want to do it
-another way, then please give me that affordance.
-
-\begin{quote}
-Is programming like `craft'? I guess some artisans make their own tools,
-but most likely not in the higher-order sense you are discussing here.
-This suggests that programming is different in some sense - but in what
-sense?
-\end{quote}
-
-This is how I segue into the software artefact for which I have been
-attempting to build a natural representation. True, it is a programmer's
-artefact, but it is still representative of what any normal person has
-to do, insofar as:
-
-\begin{enumerate}
-\def\labelenumi{\alph{enumi})}
-\tightlist
-\item
-  Wanting to create a piece of software (for whatever reason)
-\item
-  Having in mind a natural way to represent it as it's being built.
-\end{enumerate}
-
-\begin{quote}
-I think it would be good to say (fairly generally) what kind of system
-do you imagine we should be using (and goals that your OROM/SVG is
-getting closer to). This also clarifies why nothing like that exists
-already. I suppose you want something that lets you program in a way
-that is ``close to the problem domain'' (i.e.~visually if the domain is
-visual) and something that has no arbitrary boundaries
-(i.e.~meta-circular?). Boxer is the former, but not the latter;
-Smalltalk/LISP is the latter but not the former.
-\end{quote}
-
-\hypertarget{the-orom-system}{%
-\section{The OROM system}\label{the-orom-system}}
-
-When I first read the paper Open, Reusable Object Models, I was hooked
-on its idea of a small but expressive starting system that could be
-self-improved into anything. It describes a late-bound, Smalltalk-style
-objects and messaging environment that the authors call ``Id'' -- but
-which I refer to as OROM, for pronunciation and Googleability.
-
-\begin{quote}
-How much does the reader need to know about OROM for the next
-paragraphs? I think they need to know there are objects with references
-that send messages to each other - this will be clear to those who know
-what `Smalltalk-style' means, but not others. It would be useful to give
-a very brief summary before getting to the next part of the story.
-\end{quote}
-
-The paper as a publication follows the usual patterns. It consists of
-mostly prose, several code listings, some diagrams, and full C sources
-for a sample implementation at the end. Insofar as this is normal, I
-cannot criticise it. But reading this paper, and trying to grasp their
-ideas, brought to my attention how inadequate this norm can sometimes
-be.
-
-\begin{quote}
-The above feels a bit more chatty than necessary - you could probably
-shorten the paragraph into just one sentence without losing much.
-\end{quote}
-
-In order to understand it in the first place, I repeatedly found myself
-drawing extra diagrams. For example, the first acts of the running
-system boil down to initialising the three or so objects. This consists
-of allocating memory, interpreting it as a C struct and then filling in
-fields in a mundane manner. I had great difficulty grasping the
-specifics in my head, but when I drew tables in the style of the paper's
-diagrams I readily saw what what going on.
-
-Still, areas like messaging semantics could be quite confusing. After
-all, should we expect to be able to predict the entire future evolution
-of a dynamical system from a static description of its initial state,
-such as source code? Is this not why debuggers exist? Having
-``de-compiled'' English text and C source code into object diagrams, it
-is a shame to have to compile it all back to struct member assignments.
-
-Worst of all, the reference system does not even have text I/O when it
-is run, let alone some sort of GUI. That is, the (un)intended user
-interface for this project is a C debugger! Faced with the necessity of
-adding \emph{some} UI, it seemed a waste of effort to end up with a
-system that must be continually polled for its current state at a
-terminal prompt. If I naturally think of this system as 2D tables, why
-can't that be how the running system looks? I do not have to keep
-polling my eyes for what state my diagrams are in.
-
-But further than that -- why can't the system be \emph{built} out of
-tables in the first place? Shouldn't this be the main takeaway from the
-amount of time we spend prototyping, explaining and designing software
-as diagrams on paper? Why must the ``natural representation'' be
-restricted to the finished product?
-
-Thus was my natural representation decided. My first attempt to make it
-a reality was a partial success: a webpage made of HTML tables, evolved
-via JavaScript.
-
-\hypertarget{orom-as-html-tables}{%
-\subsubsection{OROM as HTML tables}\label{orom-as-html-tables}}
-
-(OROM/HTML screenshot)
-
-(Describe basic encoding of obj-dicts as tables, use of css class, etc)
-
-Unfortunately, the choice of ordinary HTML as a substrate proved rather
-two-edged. I was grateful for the browser's management of graphical
-layout, resizable text fields, and keeping the DOM tree synchronised
-with what you actually see. This last property enabled me to make the
-decision to \emph{directly} encode much of the system state in the DOM,
-achieving basic liveness (the thing on the screen is the actual thing)
-for the keys and values of obj-dicts.
-
-On the other hand, the browser requires many features to do its job of
-rendering complex web pages. And sadly, as its client, I could only make
-use of those capabilities which the W3C had decided, at the time of
-authorship, were worth the effort exposing in JavaScript. For anything
-else, the browser is a black box, and this was very frustrating in the
-following case.
-
-\hypertarget{the-radical-concept-of-arrows-that-stay-on-the-shapes}{%
-\paragraph{The Radical Concept of Arrows That Stay On The
-Shapes}\label{the-radical-concept-of-arrows-that-stay-on-the-shapes}}
-
-A key aspect of the OROM system is that there is an object \emph{graph}.
-That is, obj-dicts can have entries pointing to other obj-dicts, without
-restriction to a tree structure. Drawing arrows between things to denote
-this is a no-brainer, and I wanted it in my substrate for OROM to match
-the diagrams.
-
-A small problem to surmount first: even though I could hijack the
-\texttt{\textless{}table\textgreater{}} for its display properties, what
-element could I hijack to make arrows between arbitrary points? Luckily,
-there was SVG at my disposal, which could be persuaded to display over
-the HTML. However, another key feature of my intended substrate was to
-be able to rearrange and resize the boxes as desired. So I would also
-need to detect changes to the position and size of an element.
-
-Bizarrely, there is no such facility provided for HTML elements. This,
-despite the fact the browser \emph{needs} this functionality hence it
-must reside \emph{somewhere} inside the black box.
-
-Reluctantly, I stuck with my plan B: each object has a numerical ID and
-pointers are just fields containing a number, followed using a deref()
-function.
-
-At this point it was starting to look like a mistake not to have done
-the whole thing in SVG. I would gain full graphical freedom, though also
-lose some benefits of the browser's managing it on my behalf. I already
-knew from experience the surprising complexity of a DIY approach to
-layout, model-view updates, and interactivity. But such an exercise
-would be an opportunity to carefully observe this tedium, and
-crystallise some of my intuitions about why it is so consistently
-frustrating.
-
-\hypertarget{orom-as-files-and-directories}{%
-\subsubsection{OROM as files and
-directories}\label{orom-as-files-and-directories}}
-
-\begin{quote}
-I would probably put this after the SVG part or perhaps even later once
-you discuss more general principles (like the idea of polyfilling to
-make a domain substrate into a programmable system). Then you can give
-this as another example.
-\end{quote}
-
-\hypertarget{orom-as-svg-trees}{%
-\subsubsection{OROM as SVG trees}\label{orom-as-svg-trees}}
-
-(Screenshot of OROM/SVG showing boxes, code and arrows)
-
-In this version, obj-dicts are encoded as nested SVG
-\texttt{\textless{}rect\textgreater{}}s and other elements, reminiscent
-of diSessa's Boxer. This was a significant departure from the table
-representation, and even though SVG supports (some) nested HTML via
-\texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
-the possibility of multiple levels of nesting.
-
-OROM/SVG more or less realises my desired substrate for implementing
-OROM. Needless to say, this version was far more challenging and took
-much longer to reach a satisfactory state. However, it is precisely this
-drudgery that brings me to a better understanding of this paper's
-question: \emph{what has it taken?} I shall discuss this in the form of
-broad patterns or themes that stand out to me from my development
-experience, hammered home by these OROM projects.
-
-\begin{quote}
-Why is OROM/SVG more challenging? I think it would be useful to sketch
-here, because it reveals some useful principles. There is more
-polyfilling (arrows and resizing is not in the SVG substrate itself),
-but at least it's possible to add those (unlike in HTML)
-\end{quote}
-
-\hypertarget{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}{%
-\subsection{Inevitable Requirements Of Most Software (and the work we
-must do to meet
-them)}\label{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}}
-
-As programmers, we have a maxim along the lines of: when you find
-yourself repeating the same thing over and over again, factor it into a
-``thing'' and let the computer do the duplication. Some consider the
-``design pattern'' to be simply what a ``language feature'' looks like
-when it's not part of the syntax.
-
-\begin{quote}
-The jump from very general abstraction idea to design patterns and
-language features in the above is not exactly clear - it would be good
-to help the reader understand why are you talking about this here.
-\end{quote}
-
-As I developed OROM/SVG I found myself implementing many such patterns.
-It seems that languages are often at the wrong level of abstraction for
-the requirements of modern software, necessitating the same boilerplate
-per project just to get up to basic functionality. This burden either
-falls on the author, or on the wider community to build and maintain
-higher-level frameworks, syntax extensions, etc.
-
-By contrast, these languages seem much better adapted to e.g.~batch-mode
-printf()-centric applications. That they fail for the \emph{common case}
-should be concerning. I will now present what seem to be the inevitable
-demands of this common case.
-
-\begin{quote}
-I'm not sure I understand what the above opening is trying to say - I
-suppose it might be because it is quite high-level/general. What is this
-section about more specifically? Is it about what certain kind of
-software consists of and what we need to build that? I would also be a
-bit careful about things like ``most modern software'' below. It would
-be good to instead characterize what kind of software are you interested
-in - end-user graphical systems? (a microservice or something like that
-does not need SVG\ldots{})
-\end{quote}
-
-\hypertarget{as-a-mere-consumer}{%
-\subsubsection{As A ``Mere Consumer''}\label{as-a-mere-consumer}}
-
-I begin by describing expectations of software that even ``end-users''
-hold, consigned as they are to more or less accept what we give them.
-
-\hypertarget{retained-mode-vector-graphics}{%
-\paragraph{Retained-Mode Vector
-Graphics}\label{retained-mode-vector-graphics}}
-
-Most software is designed for the subset of people who have a colour
-display they can perceive. So right away it is going to require ways to
-draw coloured shapes. There are usually libraries for this (though note:
-not part of the language), but some only provide \emph{immediate mode}:
-commands to instantaneously rasterise pixels to a buffer. This is not
-enough for modern software, as we often expect animation, or at least to
-see things change as we interact. Most often we wish to see \emph{small
-changes} to the \emph{same} shapes, rather than completely different
-shapes altogether; the reification that this requires to persist between
-frames, is known as \emph{retained mode.}
-
-On this requirement, SVG fits the bill very well. Although it is not
-part of the JavaScript language \emph{per se}, it is a standard and
-widely supported technology of the Web \emph{platform}. We can observe
-that anyone with a browser \emph{in principle} has access to a powerful
-vector graphics editor -- just one with no GUI. I will return to this
-later.
-
-The SVG tree keeps the nice properties of the DOM, such as updating the
-display when shape parameters are changed. This is well-adapted to
-``I/O-bound'' software like mine, where things change only in response
-to user input. If I wanted animation, this boils down to a regular
-``advance simulation'' signal, and would require setting up some
-rendering loop. Alternatively, there is the W3C's chosen ontology of CSS
-animations, but see section ??.
-
-\hypertarget{basic-assumptions-about-physical-objects}{%
-\paragraph{Basic Assumptions About Physical
-Objects}\label{basic-assumptions-about-physical-objects}}
-
-In any software making use of vector graphics, there is usually some
-level of ``physics'' expected by users. This need not be nearly as
-exhaustive as the word ``physics'' might imply, as in e.g.~physics
-engines for games; I feel it is important to recognise it for what it is
-instead of conceiving physics as an inherently complex thing to be found
-only in specialised simulations.
-
-\begin{quote}
-People might not find this easy to believe - but I think you can
-convince them by saying that pretty much all mobile phone apps have some
-sort of ``phone touch physics'' (with sliding, opening menus etc.).
-\end{quote}
-
-All humans learn a basic set of expectations about the things they see
-around them. Some of these, such as ``things fall down'', are not
-generally appropriate to software UIs -- perhaps because the screen has
-a role in our lives a more of a table work surface rather than a
-vertical wall, even if it is vertical in real life.
-
-The level of physics in software tends to not involve force, or mass, or
-very much at all, merely position and space. One thing that all usable
-software must do, for example, is not crush many visually complex
-shapes, such as lines of text, into the same region, since it becomes
-unreadable. Such concepts of ``solid objects do not intersect'' or
-``only things at different layers may overlap'' are basic rules
-inherited from the real world of graphical presentation.
-
-I feel the need to point this out, because by default the computer does
-not know even the most obvious things about how space works, so we must
-laboriously algorithmise this intuitive concept. This is not only true
-in the case of 2-dimensional visual domains, but even in the
-1-dimensional case of memory allocation. The physics of 1D memory are
-something like this:
-
-\begin{itemize}
-\tightlist
-\item
-  This number range 0000-FFFF is like a space, numbers = points
-\item
-  Every point has at most one owner block
-\item
-  These blocks are contiguous, finite ranges (i.e.~1D boxes)
-\end{itemize}
-
-Hence the boilerplate involved to realise this in any domain with
-something resembling space. Far from being a niche topic in games and
-graphics, spatial partitioning algorithms and data structures have
-surprising relevance to more ordinary software. Both memory allocation
-and graphical layout are essential to today's; shame that only one of
-those has been recognised as such and entered the runtime of modern
-languages.
-
-\hypertarget{translationally-rigid-bodies}{%
-\subparagraph{Translationally Rigid
-Bodies}\label{translationally-rigid-bodies}}
-
-When you have both a screen and a pointing device (e.g.~touch or mouse),
-immediately it becomes worth having ways to move things around in at
-least a minimally realistic way. We can debate the appropriateness of
-Direct Manipulation (DM) for various situations. But it does make a lot
-of sense in simple cases, such moving around subdivisions of space
-(e.g.~windows) or elements of a graphical design.
-
-In OROM, the obvious candidate for this is the obj-dicts, plus all
-nested boxes in OROM/SVG. If I move the top-level rect, then I expect
-its children to move with it. This is simply the translational physics
-of rigid bodies: this set of points all move together. Of course, proper
-rigid bodies might also rotate and have mass, but this is usually
-undesirable for UI elements.
-
-Translational rigidity can be expressed as the points X and Y always
-having the same displacement from each other. Or, when one point is
-moved, the rest also move by the same delta. This is a problem of
-preserving the relationship over time, which was a significant area of
-OROM/SVG.
-
-\hypertarget{maintaining-relationships-over-time}{%
-\paragraph{Maintaining Relationships Over
-Time}\label{maintaining-relationships-over-time}}
-
-The model of state-mutation present in most imperative languages is what
-I call ``dumb'' state. The language provides an affordance to change any
-part of the state to a new value, but nothing else.
-
-What more could there be? Well, in \emph{every} software system there
-are certain rules, or ``invariants'' of internal consistency, such as
-``translational rigidity'' above. Often, changes to any part of the
-system are permissible, but only if connected or dependent parts of the
-state change in response.
-
-The job of keeping track of who depends on whom can fall either on the
-programmer, or the computer. If the programmer has to do this, they can
-only go so far managing and simulating in their head. As systems grow
-more complex, it is only natural to try and make the computer more
-intelligent to do this work. What I am building up to is that whenever
-we (or I) consider constraints, or ``reactive'' programming, or the
-``Observer'' pattern as things you only wheel out on special occasions,
-we only deceive ourselves into doing the same work less explicitly. It
-seems that such ``live state'' should be the expected common case for
-software development.
-
-If a platform does not provide a means to causally link and unlink bits
-of live state, then this must form part of the standard boilerplate.
-
-Such was the case in OROM/SVG. I implemented the system in an OOP
-fashion, and the Observable class is the most widely used. It wraps a
-current value and a list of subscribers, notifying them when it changes.
-It can be seen that this supplanted JavaScript's own ``dumb'' state by
-the frequency of the change() function throughout the code.
-
-Getting an object to follow the mouse pointer, for example when
-dragging, is conceptually very simple: an ``always equal'' relation. In
-OROM/SVG founded on live-state, this can be expressed in much the same
-way:
-
-\begin{verbatim}
-subscribe(object.position, pointer.position);
-\end{verbatim}
-
-(By default, an Observable A responds to a change from Observable B by
-adopting B's new value.)
-
-\hypertarget{nut-cracking-with-sledgehammers}{%
-\paragraph{Nut-Cracking With
-Sledgehammers}\label{nut-cracking-with-sledgehammers}}
-
-Speaking about vector graphics, physics, layout and constraint
-maintenance might give the impression of high \emph{conceptual}
-complexity at the heart of even simple software. This is not quite true,
-which makes it all the worse that there is yet still immense
-\emph{incidental} or implementation complexity.
-
-We are conditioned to only think of these in their most general forms.
-The vector graphics I use in OROM/SVG are just rects, lines, circles and
-text; a fraction of the full capability of SVG. The ``basic physics'' I
-use is dwarfed by fully general 2D or 3D physics engines. The only
-layout algorithm I had the patience to implement was a simple way to
-expand a list of boxes to fit in a new child at the bottom; the
-affordance to place and size boxes \emph{manually} staves off the rest
-for the time being! Yet search for material on layout algorithms and it
-can seem like Fully General Linear Inequality Solvers like Cassowary are
-all there is.
-
-The inevitable requirements I suggest here, do \emph{not} necessitate
-Fully General anything. In fact, such representations would make it more
-cumbersome to express what I wanted in OROM. As the old wisdom goes,
-there's no point expressing a simple regex search as an arbitrary Turing
-machine; my boxes don't \emph{have} a moment of inertia or a mass and I
-don't \emph{need} a linear optimisation solver for my space management
--- for the time being.
-
-Unfortunately, there doesn't seem to be much interest on the
-smaller-scale instances of these problems. I suggest the name
-``geometric physics'' for the subset of physics I have mentioned.
-
-\begin{quote}
-I think the above section talks about 3 things - (i) it describes some
-of the things that you had to implement for OROM/SVG, (ii) it talks
-about some general patterns that arise there (like shapes and physics)
-and (iii) it mentions some implementation details. I found it a bit
-confusing to have these all mixed together. It reads a bit as if you
-were jumping from one random topic that you encountered when
-implementing this - but that makes the section very specific to your
-implementation. How can we make it clear that there are more general
-principles here? I think it might be better if you started with some
-larger example (hypothetical is fine!) to illustrate what kind of things
-you'd like to build - e.g. imagine you're building PowerPoint in OROM -
-what would be some of the things you'd need? (for both the end
-application and the system that you're using to build it). This would
-then nicely motivate the more specific topics (graphics, physics,
-relationships).
-\end{quote}
-
-\hypertarget{as-a-developer}{%
-\subsubsection{As A ``Developer''}\label{as-a-developer}}
-
-I now turn to some things that were perhaps inevitable, in any case
-useful, as the \emph{implementor} of OROM/SVG.
-
-\hypertarget{practicality-vs.-high-aesthetics}{%
-\paragraph{Practicality vs.~High
-Aesthetics}\label{practicality-vs.-high-aesthetics}}
-
-There is the view that insists on provably-correct code written with
-strange Unicode symbols, and at the opposite end there is the attitude
-of ``get the job done'', high style be damned. Each has its merits, but
-for my task I quickly saw the latter route to be most suitable.
-
-My favourite example of this has been the ``encoding'' of obj-dicts as
-DOM trees containing particular arrangements of children. In OROM/HTML,
-I would gleefully get/set state by calling code like this:
-
-(Listing involving CSS, element querying / munging)
-
-and in OROM/SVG the story is similar. It has the advantage of the
-powerful developer tools included in browsers, and avoids the need to
-manually synchronise separate Model and View for such encoded state.
-With a little work, it also allowed the system a large degree of
-externalisability -- see Section ??.
-
-Contrast this with an approach that had to re-synthesise bits of the DOM
-whenever something changed. Admittedly this does still happen to a
-certain extent, but at least it was minimised where obvious.
-
-Another example of practicality was in the issue of positioning and
-sizing. In Section ?? I mentioned that I was able to sidestep some
-layout work by offloading it to the user. While building the system I
-tended to want to draw boxes into existence, and for that my brain's
-existing aesthetic algorithms can decide where they go. There are areas
-where the machine must be able to figure this out itself -- I don't want
-to be constantly bothered by prompts to place boxes as they are
-allocated by some running code -- but for one-offs it is a very
-convenient way to do a little less yak-shearing.
-
-\begin{quote}
-I think you could adapt the above discussion a bit and focus less on
-what you don't want (high modernism) and more on what you want - what is
-that? I think it's simplifying the problem along a different axis - to a
-certain domain with relatively simple interactions and aesthetics. (So
-rather than trying to make minimal formal calculus that's provably
-correct, you're trying to make minimal working interaction model that's
-interesting)
-\end{quote}
-
-\hypertarget{positioning-and-sizing}{%
-\paragraph{Positioning and Sizing}\label{positioning-and-sizing}}
-
-The simple desire to move and resize boxes with the mouse motivated a
-lot of the live-state and geometric physics ideas. This problem could be
-considered a microcosm of OROM/SVG: what's a natural way I
-\emph{conceive} of this behaviour, and could I implement it that way?
-
-\begin{quote}
-The above point (moving motivated live state and physics ideas) would be
-a lot more obvious if you followed the structure with motivating example
-first followed by an analysis\ldots{}
-\end{quote}
-
-It starts with a consideration of translational rigidity. In its most
-primitive form, this is a relation between two points. Thus it is
-natural to draw a line or ``rod'' between them. It seems that this rod
-transmits changes in one of its endpoints directly to the other
-endpoint. Since they both feel the same deltas, the displacement vector
-between them is preserved.
-
-(Diagram of delta transmission)
-
-I like to see what I'm doing, so I wanted these rods to be visible and
-thus somehow \emph{present in the SVG}. This was not hard; in SVG I draw
-a \texttt{\textless{}line\textgreater{}}, though the background
-machinery of the Rod class does all the work. There is also a Point
-class that is used wherever manipulable points (SVG circles) are
-intended.
-
-Resizing (of boxes at least) could be achieved through rods that stay
-horizontal or vertical. In the language of ``small differences'' spoken
-by my live-state infrastructure, this is expressed as a rod
-``transmitting deltas'' in the vertical and horizontal, ``absorbing''
-the other component into itself. Mirroring a DOM rect to these rods is
-as simple as subscribing its width and height to horizontal and vertical
-rods' length Observables. This way, boxes can be resized from whatever
-corner is convenient.
-
-(Box demonstrating all three rod types)
-
-I colour such ``half-rigid'' rods yellow and fully rigid rods red. The
-``absorb all endpoint changes'' case is coloured green.
-
-\begin{quote}
-I don't think you explained what `half-rigid' means yet.
-\end{quote}
-
-Unfortunately, with these rods came possibly the most frustrating
-technical challenges of the entire system. Initially I hoped to move
-boxes as rigid bodies by temporarily making their border rods rigid.
-However, the four border rods form a cyclic graph, as rods are not
-directed. This, coupled with the unintended depth-first semantics of
-Observable notification (a result of JavaScript function calls in a
-loop), led to duplicate deltas applied twice and other nightmares. This
-is the tip of an entire research iceberg stretching from Functional
-Reactive Programming to internet routing and distributed algorithms.
-
-I did manage to surmount this through kludging and compromise. But I do
-not know whether these problems are merely a consequence of some design
-decision I could change to escape it, or if they are intrinsic to my
-(modest) UI goal.
-
-\begin{quote}
-This section reads more as an implementation detail - even though it
-makes some interesting points about the physics \& rods system. However,
-it might better to have it in a section that's more about implementation
-than about general principles. Or perhaps you could restructure Section
-3 to have (i) example (ii) general lessons (like physics etc.) and (iii)
-specific points about implementation (e.g.~your concrete physics system
-with rods, what you had to `polyfill')
-\end{quote}
-
-\hypertarget{visible-coordinate-systems}{%
-\paragraph{Visible Coordinate
-Systems}\label{visible-coordinate-systems}}
-
-Rigidity in a flat world of sibling shapes is somewhat straightforward.
-However, rigidity in the SVG world is more involved.
-
-First of all, SVG shapes (e.g.~\texttt{\textless{}rect\textgreater{}})
-are strictly leaf nodes, so if I wish to nest boxes within boxes, the
-visible ``box'' must be a mere \emph{accessory} to the nestable element,
-in my case a \texttt{\textless{}g\textgreater{}} (group). This means
-that instead of resizing the x, y, width, height attributes of the
-\texttt{\textless{}rect\textgreater{}}, only its width and height change
-along with the transform attribute of its \emph{parent}
-\texttt{\textless{}g\textgreater{}}. This was not too bad; just
-subscribe this attribute, instead of the
-\texttt{\textless{}rect\textgreater{}} position, to the top-left Point
-handle.
-
-All child elements of a node transform with it, so already SVG has baked
-in a basic facility for translational rigidity. This is only available
-as a tree hierarchy, but it is still useful. However, it conflicts with
-my early decision to have Point objects all share the global co-ordinate
-system. This way, connecting ``this point'' to the mouse pointer, etc.
-is not infuriating to express. Still, it was necessary in the case of
-certain elements -- especially those which must transcend the tree
-structure altogether, like arrows between boxes -- to bite this bullet,
-one way or another.
-
-\begin{quote}
-Maybe mention that this is caused by the mismatch between the substrate
-and the programming system (substrate is tree, but programming is
-graphs)?
-\end{quote}
-
-Again, I return to how we tend to work things out in the freedom of
-paper. Co-ordinate systems, here merely positionally displaced, have
-their origins here and there and have vectors between them. The rods
-thus far let me visually express relations between global Points; now
-was a question of expressing one global Point as a displacement from
-another (the \texttt{\textless{}g\textgreater{}} transform). New rod
-Observables p2\_from\_p1 and p1\_from\_p2 do the vector subtraction,
-which can then be propagated as local co-ordinates to children. It is
-nicer to express the relation (as well as see it!) this way -- see
-Figure ?? above.
-
-\hypertarget{context-appropriate-ontologies}{%
-\paragraph{Context-appropriate
-ontologies}\label{context-appropriate-ontologies}}
-
-\ldots{}
-
-Talk about: fixed ontologies (a rect IS x,y,w,h, and nothing else!)
-imposed on programmer
-
-Example from 3.1.1. mention: CSS animations?
-
-``One size fits all'' rears its ugly head again.
-
-I want: top-left, top-right, bot-left, bot-right \ldots{}
-
-Instead of explaining ``the x y is the top left blah blah'' in English
-in the W3C spec, they should make these relationships \emph{machine
-readable} with information like:
-
-A Rect is
-
-\begin{itemize}
-\tightlist
-\item
-  a Polygon
-\item
-  defined by 4 degrees of freedom x y width height
-\item
-  where the four vertices are
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \texttt{{[}x,y{]}} ``top-left''
-  \item
-    \texttt{{[}x+width,y{]}} ``top-right''
-  \item
-    etc
-  \end{itemize}
-\end{itemize}
-
-Reference to paper related to Antranig involving knitting\ldots{}?
-
-\ldots{}
-
-\hypertarget{extensional-functions}{%
-\paragraph{Extensional Functions}\label{extensional-functions}}
-
-aka dicts, Maps, std::vector, \ldots{}
-
-\ldots{} \ldots{} Some parts of the OROM authors' C code were confusing
-until I realised they were just the guts of a basic associative-array
-implementation. \ldots{} \ldots{}
-
-Inevitability / ubiquity of state descriptions that are at trees of
-strings with pointers
-
-Antranig ``natural co-ordinates''?
-
-OROM/SVG as Boxer-like extensional functions
-
-\hypertarget{externalisability}{%
-\paragraph{Externalisability}\label{externalisability}}
-
-Reference Antranig's stuff on this. How I dealt with event listeners etc
-
-Necessity to development: frustrating to find bug, tear down, refresh,
-lose work, build it up again!
-
-Inspect Element: copy to save works, arbitrary edits \emph{should} work
-(cuz it's on demand)
-
-Work required to parse SVG trees into objects ``on demand''
-
-Started heavyweight, ended lightweight (generate rods on demand, let SVG
-move children)
-
-\hypertarget{polyfilling}{%
-\paragraph{Polyfilling}\label{polyfilling}}
-
-Very advantageous to permit this e.g.~JavaScript
-
-Stupid lack of XYZ in JS or SVG due to come out in the next spec, etc
-
-Syntax sugar (SVG tree ops, vector ops)
-
-\hypertarget{the-orom-system-as-a-part-of-the-solution}{%
-\subsection{The OROM system as a part of the
-solution}\label{the-orom-system-as-a-part-of-the-solution}}
-
-Meta-circularity as empowerment, not just ``cool''
-
-The allure
-
-Other themes e.g.~homoiconicity of code and data
-
-Problems: cognitive vs.~``official'' complexity
-
-Maxwell's eqns metaphor: omits the years of study compressed into the
-math symbols!
-
-And yet, the allure of the Maxwell's Eqns of Software?
-
-``Taming'' SVG - ability to use SVG in a vaguely sane way (vector
-graphics editor)
-
-Similarly for e.g.~3D OpenGL, Web Audio, sockets \ldots{}
-
-\hypertarget{tensions-between-philosophies}{%
-\subsubsection{Tensions between
-philosophies}\label{tensions-between-philosophies}}
-
-\hypertarget{stable-or-emergent-antranig-against-oop-dispatch}{%
-\paragraph{Stable or emergent? (Antranig against OOP
-dispatch)}\label{stable-or-emergent-antranig-against-oop-dispatch}}
-
-\hypertarget{turing-complete-antranig-against-this}{%
-\paragraph{Turing-complete? (Antranig against
-this)}\label{turing-complete-antranig-against-this}}
-
-\hypertarget{messaging-vs-read-write}{%
-\paragraph{Messaging vs Read-Write?}\label{messaging-vs-read-write}}
-
-\hypertarget{misc}{%
-\section{Misc}\label{misc}}
-
-A note about DIY and libraries
-
-My other problem was one of cost-benefit. My aim was to \ldots{}
+\input{../manuscript.tex}
 

### 2020-01-22 convivial-2020/manuscript.tex
DONE

@@
 \subtitle{A case study on the ``OROM'' system}
-\author{Joel Jakubovic}
-\affiliation{University of Kent, Canterbury}
+\author[ ]{Joel Jakubovic}
+\affiliation[ ]{University of Kent, Canterbury}
 
@@
 \newcommand{\svgel}[1]{\texttt{\textless{}#1\textgreater{}}}
-\newcommand{\OSFA}{One-Size-Fits-All}
+\newcommand{\OSFA}{One-Size-Fits-All }
 
 \begin{abstract}
-There is a One-Size-Fits-Allquality to languages, APIs and even programming itself.
+There is a One-Size-Fits-All quality to languages, APIs and even programming itself.
 Whether you're making a mobile game or a scientific simulation, you will be using a text-based language with similar devices for structuring your code.
@@
 One would have thought that software-creation acts like hypothetical
-super-intelligent AI. That is: even though we start from a primitive
-base in the 50s (or even today), there would surely be a recursive
-process of self-improvement, building better software-creation tools
-with the existing ones, until an ``expressivity singularity'' where
-software becomes a workable material as described.
+super-intelligent Artificial Intelligence (AI). That is: even though we
+start from a primitive base in the 50s (or even today), there would
+surely be a recursive process of self-improvement, building better
+software-creation tools with the existing ones, until an ``expressivity
+singularity'' where software becomes a workable material as described.
 
@@
 failed historical attempts to do better --- or at least, ones that
-failed to catch on. But even worse than this is that in my very
+failed to catch on. And even worse than this is that in my very
 \emph{language} here I am making the same mistake as the text editor ---
 speaking in unqualified terms of ``better'' and ``worse'' as if there
-really is a One-Size-Fits-Allsolution to software creation!
+really is a One-Size-Fits-All solution to software creation!
 
@@
 thing they want to make. And second-order tools that suit them for
-making the first-order tools, etc. It would do no good to replace
+making the first-order tools, and so on. It would do no good to replace
 text-imperialism with anything-else-imperialism, which is one
@@
 (e.g.~languages, tools, or subject areas) each serving many clients, the
-One-Size-Fits-Allstyle is the best one can hope for. Adaptation to
+One-Size-Fits-All style is the best one can hope for. Adaptation to
 individual preferences and idiosyncrasies is only feasible when those
@@
 
-When I first read the paper Open, Reusable Object Models, I was hooked
-on its idea of a small but expressive starting system that could be
-self-improved into anything. It describes a late-bound\footnote{Fewer
-  commitments; more things determined by runtime conditions},
-Smalltalk-style\footnote{OOP with more emphasis on object instances and
-  messaging (methods) as in a distributed system; less emphasis on class
-  hierarchies implementing traditional data structures} objects and
-messaging environment that the authors call ``Id'' --- but which I refer
-to as OROM, for pronunciation and Googleability.
+When I first read the paper ``Open, Reusable Object Models''
+\cite{OROM}, I was hooked on its idea of a small but expressive starting
+system that could be self-improved into anything. It describes a
+late-bound\footnote{Fewer commitments; more things determined by runtime
+  conditions}, Smalltalk-style\footnote{OOP with more emphasis on object
+  instances and messaging (methods) as in a distributed system; less
+  emphasis on class hierarchies implementing traditional data structures}
+objects and messaging environment that the authors call ``Id'' --- but I
+shall refer to it as OROM, for pronunciation and Googleability.
 
@@
 not necessarily representative of those who read it. The
-One-Size-Fits-Allapproach is perhaps unavoidable for static media like
+One-Size-Fits-All approach is perhaps unavoidable for static media like
 print.
@@
 a reality was a partial success: a webpage made of HTML tables, evolved
-via JavaScript.
+via JavaScript (JS).
 
@@
 mappings, I will refer to them as \emph{obj-dicts}. Figure
-\ref{fig:orom-html} shows the OROM/HTML implementation, in which they
-take the form of HTML tables.
+\ref{fig:orom-html} shows the OROM/HTML implementation \cite{orom-html},
+in which they take the form of HTML tables.
 
@@
 
-For each obj-dict I used a two-column
-\texttt{\textless{}table\textgreater{}} within a
-\texttt{\textless{}div\textgreater{}}, matching the diagrams, and gave
-the rows CSS class names matching the keys for easy lookup. Sending
-messages relied on the JavaScript console, but existing values in text
-boxes (such as method implementations) could be edited directly. Also
-thanks to the browser, I could easily make text boxes and
-\texttt{\textless{}div\textgreater{}}s resizable as necessary.
-
-However, the choice of ordinary HTML as a substrate proved rather
+I used a two-column \texttt{\textless{}table\textgreater{}} within a
+\texttt{\textless{}div\textgreater{}} for each obj-dict (matching the
+diagrams) and gave the rows CSS class names matching the keys (for easy
+lookup). Sending messages relied on the JS console, but existing values
+in text boxes (such as method implementations) could be edited directly.
+
+This choice of ordinary HTML as a substrate, however, proved rather
 two-edged. The browser requires many features to do its job of rendering
@@
 That is, obj-dicts can have entries pointing to other obj-dicts, without
-restriction to a tree structure. Drawing arrows between things to denote
-this is a no-brainer, and I wanted it in my substrate for OROM to match
-the diagrams.
+restriction to a tree structure. Drawing arrows to denote this is a
+no-brainer (and easy on paper), so I wanted it in my substrate for OROM.
 
@@
 element could I hijack to make arrows between arbitrary points? Luckily,
-there was SVG at my disposal, which could be persuaded to display over
-the HTML. However, another key feature of my intended substrate was to
-be able to rearrange and resize the boxes as desired. So I would also
-need to detect changes to the position and size of an element.
+there were Scalable Vector Graphics (SVG) at my disposal, which could be
+persuaded to display \texttt{\textless{}line\textgreater{}}s over the
+\texttt{\textless{}div\textgreater{}}s. But another key feature of my
+intended substrate was to be able to rearrange and resize the boxes. So
+I would also need to detect changes to the position and size of an
+element.
 
@@
 
-\hypertarget{practicality-vs.-high-aesthetics}{%
-\subsection{Practicality vs.~High
-Aesthetics}\label{practicality-vs.-high-aesthetics}}
-
-There is the view that insists on provably-correct code written with
-strange Unicode symbols, and at the opposite end there is the attitude
-of ``get the job done'', high style be damned. Each has its merits, but
-for my task I quickly saw the latter route to be most suitable.
-
-My favourite example of this has been the ``encoding'' of obj-dicts as
-DOM trees containing particular arrangements of children. In OROM/HTML,
-I would gleefully get/set state by calling code like this:
-
-\begin{lstlisting}[language=JavaScript]
-Entity.prototype.addState = function(key) {
-  let tr = document.createElement('tr');
-  tr.setAttribute('class', key);
-
-  let td = document.createElement('td');
-  td.setAttribute('class', 'key');
-  td.textContent = key;
-  tr.appendChild(td);
-  ...
-\end{lstlisting}
-
-and in OROM/SVG the story is similar. It takes advantage of the powerful
-developer tools included in browsers (e.g.~the DOM inspector), and
-avoids the need to manually synchronise separate Model and View for such
-encoded state. With a little work, it also allowed the system to be
-saved and restored --- see Section \ref{persistence}. I consider this as
-practical, expedient, ``hijacking'' of these substrates because it
-either goes against their intended use (HTML tables for presentation
-rather than semantics) or are simply ``odd'' (using certain arrangements
-of shapes to ``encode'' machine-readable semantics).
-
-Contrast this with an approach that had to re-synthesise bits of the DOM
-whenever something changed. Admittedly this does still happen to a
-certain extent, but at least it was minimised where obvious.
-
-Another example of practicality was in the issue of positioning and
-sizing. In Section \ref{nut-cracking-with-sledgehammers} I mentioned
-that I was able to sidestep some layout work by offloading it to the
-user. While building the system I tended to want to draw boxes into
-existence, and for that my brain's existing aesthetic algorithms can
-decide where they go. There are areas where the machine must be able to
-figure this out itself --- I don't want to be constantly bothered by
-prompts to place boxes as they are allocated by some running code ---
-but for one-offs it is a very convenient way to do a little less
-yak-shearing.
-
 \hypertarget{orom-as-svg-trees}{%
@@
   \caption{OROM/SVG: Obj-dicts are moveable and resizable nested boxes,
-           referencing each other via real arrows.\label{fig:orom-svg}}
+           referencing each other via \emph{real arrows}.\label{fig:orom-svg}}
 \end{figure}
 
-In this version, shown in Figure \ref{fig:orom-svg}, obj-dicts are
-encoded as nested SVG \texttt{\textless{}rect\textgreater{}}s and other
-elements, reminiscent of diSessa's Boxer. This was a significant
-departure from the table representation, and even though SVG supports
-(some) nested HTML via \texttt{\textless{}foreignObject\textgreater{}},
-I actually preferred the possibility of multiple levels of nesting.
-
-OROM/SVG more or less realises my desired substrate for implementing
-OROM. Needless to say, this version was far more challenging and took
-much longer to reach a satisfactory state. However, it is precisely this
-drudgery that brings me to a better understanding of this paper's
-question: \emph{what has it taken?} I shall discuss this in the form of
-broad patterns or themes that stand out to me from my development
-experience, hammered home by these OROM projects.
+In OROM/SVG, shown in Figure \ref{fig:orom-svg}, obj-dicts are encoded
+as nested SVG \texttt{\textless{}rect\textgreater{}}s and other
+elements, reminiscent of diSessa's Boxer\cite{boxer}. This was a
+significant departure from the table representation, and even though SVG
+supports (some) nested HTML via
+\texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
+the possibility of multiple levels of nesting.
+
+Still, this version was far more challenging and took much longer to
+reach a satisfactory state. Yet it is precisely this drudgery that
+brings me to a better understanding of this paper's question: \emph{what
+has it taken?} I shall discuss this in the form of broad patterns or
+themes that stand out to me from my development experience, hammered
+home by these OROM projects.
 
@@
 ``programming'' more like ``using'', such expectations on \emph{use} at
-least need to be acknowledged. In this section, I will present these and
-other seemingly \emph{inevitable} demands of normal software,
-exemplified by OROM/SVG. I will comment how our programming platforms
-measure up to the task of helping us meet them, including my chosen
-platform of JavaScript and Web technologies.
+least need to be acknowledged.
+
+In this section, I will present these and other seemingly
+\emph{inevitable} demands of normal software, exemplified by OROM/SVG. I
+will comment on how our programming platforms measure up to the task,
+including my chosen platform of JavaScript and Web technologies.
 
@@
 that anyone with a browser \emph{in principle} has access to a powerful
-vector graphics editor --- just one with no GUI. I will return to this
-at the end (Section ??).
-
-The SVG tree keeps the nice properties of the DOM, such as updating the
+vector graphics editor --- just one with no GUI.
+
+The SVG tree has the nice properties of the DOM, such as updating the
 display when shape parameters are changed. This is well-adapted to
 ``I/O-bound'' software like mine, where things change only in response
-to user input. If I wanted animation, this boils down to a regular
+to user input. If I wanted animation, this would boil down to a regular
 ``advance simulation'' signal, and would require setting up some
@@
 a role in our lives a more of a table work surface rather than a
-vertical wall, even if it is vertical in real life.
-
-The level of physics in software tends to not involve force, or mass, or
-very much at all, merely position and space; we could call it
-``geometric physics''. One thing that all usable software must do, for
-example, is not crush many visually complex shapes, such as lines of
-text, into the same region, since it becomes unreadable. Such concepts
-of ``solid objects do not intersect'' or ``only things at different
-layers may overlap'' are basic rules inherited from the real world of
-graphical presentation.
+vertical wall, even if it is vertical in real life. The level of physics
+in software tends to not involve force, or mass, or very much at all,
+merely position and space; we could call it ``geometric physics''.
+
+One thing that all usable software must do, for example, is avoid
+crushing many visually complex shapes, such as lines of text, into the
+same (unreadable) region. Such concepts of ``solid objects do not
+intersect'' or ``only things at different layers may overlap'' are basic
+rules inherited from the real world of graphical presentation.
 
@@
 \item
-  This number range 0000-FFFF is like a space, numbers = points
+  This number range \texttt{0000}--\texttt{FFFF} is like a space,
+  addresses = points
 \item
@@
 and graphical layout are essential to today's; shame that only one of
-those has been recognised as such and entered the runtime of modern
+those has been recognised as such --- and entered the runtime of modern
 languages.
@@
 least a minimally realistic way. We can debate the appropriateness of
-Direct Manipulation (DM) for various situations. But it does make a lot
-of sense in simple cases, such moving around subdivisions of space
+Direct Manipulation for various situations. But it does make a lot of
+sense in simple cases, such moving around subdivisions of space
 (e.g.~windows) or elements of a graphical design.
@@
 its children to move with it. This is simply the translational physics
-of rigid bodies: this set of points all move together. Of course, proper
-rigid bodies might also rotate and have mass, but this is usually
-undesirable for UI elements.
+of \textbf{rigid bodies}: ``this set of points all move together''. Of
+course, proper rigid bodies might also rotate and have mass, but this is
+usually undesirable for UI elements.
 
@@
 moved, the rest also move by the same delta. This is a problem of
-preserving the relationship over time, which was a significant area of
-OROM/SVG.
+\textbf{preserving the relationship over time}, which was a significant
+area of OROM/SVG.
 
@@
 What more could there be? Well, in \emph{every} software system there
-are certain rules, or ``invariants'' of internal consistency, such as
-``translational rigidity'' above. Often, changes to any part of the
-system are permissible, but only if connected or dependent parts of the
-state change in response.
+are certain rules, or ``invariants'' of \textbf{internal consistency},
+such as ``translational rigidity'' above. Often, changes to any part of
+the system are permissible, but only if connected or dependent parts of
+the state change in response.
 
@@
 intelligent to do this work. What I am building up to is that whenever
-we (or I) consider constraints, or ``reactive'' programming, or the
-``Observer'' pattern as things you only wheel out on special occasions,
-we only deceive ourselves into doing the same work less explicitly. It
-seems that such ``live state'' should be the expected common case for
-software development.
+we (or I) consider constraints, ``reactive'' programming, or the
+``Observer'' pattern as things you only wheel out \emph{on special
+occasions}, we only deceive ourselves into doing the \emph{same work}
+less explicitly. It seems that such ``live state'' should be the
+expected common case for software development.
 
@@
 
-Getting an object to follow the mouse pointer, for example when
-dragging, is conceptually very simple: an ``always equal'' relation. In
+Getting an object to follow the mouse pointer (e.g.~when dragging) is
+\emph{conceptually} very simple: an ``always equal'' relation. In
 OROM/SVG founded on live-state, this can be expressed in much the same
@@
 
-(By default, an Observable A responds to a change from Observable B by
-adopting B's new value.)
+By default, an Observable A responds to a change from Observable B by
+adopting B's new value. So by subscribing the object's position to the
+pointer's, pointer movements copy the new position to the object.
 
@@
 which makes it all the worse that there is yet still immense
-\emph{incidental} or implementation complexity.
+\emph{implementation} complexity.
 
 We are conditioned to only think of these in their most general forms.
-The vector graphics I use in OROM/SVG are just rects, lines, circles and
-text; a fraction of the full capability of SVG. The ``geometric
-physics'' I use is dwarfed by fully general 2D or 3D physics engines.
-The only layout algorithm I had the patience to implement was a simple
-way to expand a list of boxes to fit in a new child at the bottom; the
-affordance to place and size boxes \emph{manually} staves off the rest
-for the time being! Yet search for material on layout algorithms and it
-can seem like Fully General Linear Inequality Solvers like Cassowary are
-all there is.
+But the ``vector graphics'' I use in OROM/SVG are just rects, lines,
+circles and text; a fraction of the full capability of SVG. The
+``geometric physics'' I use is dwarfed by fully general 2D or 3D physics
+engines. The only layout algorithm I had the patience to implement was a
+simple way to expand a list of boxes to fit in a new child at the
+bottom. The affordance to place and size boxes \emph{manually} is a
+convenient substitute, when required infrequently. Yet search for
+material on layout algorithms, and it can seem like Fully General Linear
+Inequality Solvers like Cassowary \cite{cassowary} are all there is.
 
 The inevitable requirements I suggest here, do \emph{not} necessitate
-Fully General anything. In fact, such representations would make it
+Fully General anything. In fact, such generality might make it
 \emph{more} cumbersome to express what I wanted in OROM. As the old
@@
 space management --- for the time being. Under the theme of
-domain-appropriate tools, it is worth designing interfaces for
-smaller-scale instances of these areas and exploring their capabilities.
+domain-appropriate tools, I think it is worth designing interfaces for
+\emph{smaller-scale} instances of these areas and exploring what they
+can express.
 
@@
 mode file I/O tasks, but that they fail for the \emph{common case} is a
-problem. It necessitates largely the same ``boilerplate'' setup per
-project just to get basic functionality:
+problem. There is largely the same setup per project just to get basic
+functionality:
 
@@
 term. But ultimately, all platforms are designed with certain features
-``out of the box'' and leave other features to be implemented ``if
-needed''. And when a problem domain ``requires batteries'' yet the tools
-available say ``batteries not included'', it is quite reasonable to want
-the batteries to be part of the platform.
-
-This section covers the boilerplate I found myself polyfilling in
-OROM/SVG.
-
-Is to bring things into the mindset / ontology of the specific software.
-For example, wrapping the official Web conventions for event handling in
-the Observable infrastructure, or wrapping DOM nodes with ``user data''
-objects. The latter case, takes advantage of JavaScript's ability to
-simply ``add on'' new fields to objects, even objects of Web APIs, to
-``annotate'' the DOM node with a JavaScript helper. Very advantageous to
-permit this in JavaScript. In another language like C, such would
-require saving the address in a lookup table elsewhere.
+``out of the box'' and leave other features for us to implement ``if
+needed''.
+
+When the \emph{normal} problem domain ``requires batteries'', yet the
+tools available say ``batteries not included'', it is quite reasonable
+to want the batteries to be part of the platform. This section details
+the ``batteries not included'' I found myself polyfilling in OROM/SVG.
+
+\hypertarget{wrapping-and-user-data}{%
+\subsection{Wrapping and ``user data''}\label{wrapping-and-user-data}}
+
+The classic example of polyfilling is finding that JavaScript arrays do
+not support some familiar operation. For example, this used to be the
+case with the \texttt{map} function. Yet JavaScript lets you
+\emph{directly augment} the \texttt{Array} class with the new operation,
+even though \texttt{Array} is ``part of the language'':
+
+\begin{lstlisting}
+Array.prototype.map = function() { ... }
+\end{lstlisting}
+
+One can do this to all objects, even those of Web APIs. For example, I
+use it to connect SVG nodes to the OROM system: when a
+\texttt{\textless{}rect\textgreater{}} is created, it is ``annotated''
+with a \texttt{userData} pointer to the box that it is part of. Then,
+when clicked, the box can be found immediately.
+
+This could be described as the box ``wrapping'' the
+\texttt{\textless{}rect\textgreater{}}, but risks impliying that the
+\texttt{\textless{}rect\textgreater{}} merely needs to \emph{live
+inside}, or \emph{be referenced by}, the box. In this case, the
+\emph{reverse} association is also needed, because clicks can only enter
+the system \emph{through} SVG in the first place. What is significant
+here is that JavaScript lets us add such a link to an entity that
+\emph{we did not design}.
+
+This term, ``user data'', often crops up in libraries with the same
+idea: to associate arbitrary data with an API object. But in languages
+that do not permit such ``annotation'', this has to be explicitly
+\emph{designed in} ahead of time. If it wasn't, then it has to be done
+in a roundabout way --- for example, a separate \texttt{userData} lookup
+table, indexed by the memory address of the API object.
 
@@
   \centering
-  \includegraphics[width=0.75\textwidth]{../rods.png}
+  \includegraphics[width=\textwidth]{../rods.png}
   \caption{The yellow rods on the box border are ``half-rigid'': they transmit horizontal or vertical changes, but not both.
@@
 
-I did manage to surmount this through kludging and compromise. But I do
-not know whether these problems are merely a consequence of some design
-decision I could change to escape it, or if they are intrinsic to my
-(modest) UI goal.
+I had to shelve this investigation in the interest of continuing with
+the rest of the system, but I did manage to surmount this through
+kludging and compromise. So I do not know whether these problems are
+merely a consequence of some design decision I could change to escape
+it, or if they are intrinsic to my (modest) UI goal.
 
@@
 in a basic facility for translational rigidity. This is only available
-as a tree hierarchy, but it is still useful. However, it conflicts with
-my early decision to have Point objects all share the global co-ordinate
-system (this was to ensure that simple relations, such as a point
-following the mouse pointer, are not infuriating to express.) Still, it
-was necessary in the case of certain elements --- especially those which
-must transcend the tree structure altogether, like arrows between boxes
---- to bite this bullet, one way or another.
-
-\begin{quote}
-Maybe mention that this is caused by the mismatch between the substrate
-and the programming system (substrate is tree, but programming is
-graphs)?
-\end{quote}
+as a tree hierarchy\footnote{This highlights the mismatch between the
+  tree-based DOM and any system that is graph-structured.}, but it is
+still useful. However, it conflicts with my early decision to have Point
+objects all share the global co-ordinate system (this was to ensure that
+simple relations, such as a point following the mouse pointer, are not
+infuriating to express.) Still, it was necessary in the case of certain
+elements --- especially those which must transcend the tree structure
+altogether, like arrows between boxes --- to bite this bullet, one way
+or another.
 
@@
 vector subtraction, which can then be propagated as local co-ordinates
-to children. It is nicer to express the relation (as well as see it!)
+to children. It is nice to express the relation (as well as see it!)
 this way (Figure \ref{fig:rods}).
@@
 Each API has its own conventions, including a way of naming and
-structuring expressions --- an ontology. The One-Size-Fits-Allapproach
-is exemplified in such interfaces. For example, in SVG we express a
-rectangle by
+structuring expressions --- an ontology \cite{crit-semprola}. The
+One-Size-Fits-All approach is exemplified in such interfaces. For
+example, in SVG we express a rectangle by
 \texttt{\textless{}rect x="10" y="10" width="600" height="400"\textgreater{}}.
-The SVG specification \emph{defines} to the user that a rect \emph{is} a
-top-left corner, a width, and a height --- and that's it. However, this
-SVG-approved parametrisation of a rectangle is far from the only
-one\href{https://www.shift-society.org/salon/papers/2018/critiques/critique-semprola.pdf}{1},
-and thus is, unsurprisingly, ill-fitted to some contexts.
+The SVG specification \emph{defines} to the user that a rect simply
+\textbf{is} a top-left corner, a width, and a height --- and that's it.
+However, this SVG-approved parametrisation of a rectangle is far from
+the only one, and thus is, unsurprisingly, ill-fitted to some contexts.
 
@@
 Conceptually, many ``events'' are in fact changes to the state of some
-physical (or conceptual) device. And both keyboard keys and mouse
-buttons, for example, have two states --- pressed, not pressed --- which
-ought to make them \emph{interchangeable} to this extent. This is the
-logic behind ``key mapping''\footnote{For example, the player can
+physical device. And both keyboard keys and mouse buttons, for example,
+have two states --- pressed, not pressed --- which ought to make them
+\emph{interchangeable} to some extent. Indeed, this is why there is such
+a thing as ``key mapping''\footnote{For example, the player can
   re-assign the action ``shoot'' from its default mouse button to a
@@
 
-To start with, the situation is modelled not as a time-varying value
-changing, but instead as a sort of Cartesian product of event listeners.
-Rather than, say, a piece of live-state for the left mouse button (LMB),
-we get \texttt{onmousedown} and \texttt{onmouseup}. Thankfully we do not
-have \texttt{onAdown}/\texttt{onAup}, \texttt{onBdown/onBup}, \ldots{}
-all the way to \texttt{onZdown}/\texttt{onZup}, but only
+To start with, the situation is modelled not as the \emph{changing} of
+some time-varying property, but instead as a sort of Cartesian product
+of event listeners. Rather than, say, a piece of live-state for the left
+mouse button (LMB), we get \texttt{onmousedown} and \texttt{onmouseup}.
+Thankfully we do not have \texttt{onAdown}/\texttt{onAup},
+\texttt{onBdown/onBup}, \ldots{} all the way to
+\texttt{onZdown}/\texttt{onZup}, but only
 \texttt{onkeydown}/\texttt{onkeyup}. Yet this is just one of the many
@@
 form ``subscribe to pointer \emph{only when} pointer is-considering-me
-\emph{and} LMB is down'', but I could live with this as a future
-polyfill in JavaScript.
+\emph{and} LMB is down'', but I could live with this notation as a
+future polyfill in JavaScript.
 
 The way these things are connected to the browser's event listeners
-could be called ``device drivers''. This approach is described in more
-detail in \href{https://prog21.dadgum.com/66.html}{13}. It amounts to
-translating information from the ``unchangeable'' Web platform into the
-worldview of my substrate, as early as possible:
+could be called ``device drivers'' --- an approach described in
+\cite{prog21-bbox}. It amounts to translating information from the Web's
+ontology into that of my substrate, as early as possible:
 
@@
 
-What would it look like to support multiple ontologies? There are the
-two answers: anticipate the possibilities ahead of time, or support
-users adding their own. It is unclear if we can do better on the latter
-than simply ``support polyfilling''. As for the former: in general,
-anticipating the diversity of ways someone might look at the world is
-doomed to fail. But in the case of fairly \emph{formalised} concepts
-such as geometrical shapes or mathematics, I could suggest that simple
-\emph{under-specification} is the root of the problem in SVG. Instead of
-explaining in English that ``\texttt{x} and \texttt{y} are the
-co-ordinates of the top left-hand corner\ldots{}'', it might be better
-to make these relations machine-readable or \emph{embodied} in the API.
-For the rect this might look like:
+What would it look like to support multiple ontologies? There are
+perhaps two answers: anticipate the possibilities ahead of time, or
+support users adding their own. It is unclear if we can do better on the
+latter than simply ``support polyfilling''. As for the former: in
+general, anticipating the diversity of ways someone might look at the
+world is doomed to fail. But in the case of fairly \emph{formalised}
+concepts such as geometrical shapes or mathematics, I could suggest that
+simple \emph{under-specification} is the root of the problem in SVG.
+Instead of the specification explaining in English that ``\texttt{x} and
+\texttt{y} are the co-ordinates of the top left-hand corner\ldots{}'',
+it might be better to make these relations machine-readable or
+\emph{embodied} in the API. For the rect this might look like:
 
@@
 \item
-  a \texttt{polygon} (defined elsewhere)
+  a \texttt{polygon} (to be defined elsewhere)
 \item
@@
   \item
-    \texttt{{[}x,y{]}} ``top-left''
-  \item
-    \texttt{{[}x+width,y{]}} ``top-right''
-  \item
-    \texttt{{[}x,y+height{]}} ``bot-left''
-  \item
-    \texttt{{[}x+width,y+height{]}} ``bot-right''
+    \texttt{{[}x,y{]}} called \texttt{top-left}
+  \item
+    \texttt{{[}x+width,y{]}} called \texttt{top-right}
+  \item
+    \texttt{{[}x,y+height{]}} called \texttt{bot-left}
+  \item
+    \texttt{{[}x+width,y+height{]}} called \texttt{bot-right}
   \end{itemize}
@@
 effort even for precise mathematical knowledge structures, or could
-enough formalisation of Euclidean geometry in the Web platform put an
-end to this sort of polyfilling?\footnote{This relates to my intuition
-  that type systems and other auto-reasoning tools are trapped in a
-  doomed quest for ``closed-form'' AI, representable as a
-  \LaTeX formula.}
+enough formalisation\footnote{A related question is whether type systems
+  and other auto-reasoning formalisms are trapped in a doomed quest for
+  ``closed-form'' AI, representable as a \LaTeX~formula.} of Euclidean
+geometry in the Web platform put an end to this sort of polyfilling?
 
@@
 \texttt{window.my\_obj.component}. The common case is an association of
-a textual (string) name, to an arbitrary value. I find it useful to see
-this as a mathematical ``function'' defined \emph{extensionally} by
-listing its input/output mappings --- this opposed to an
-\emph{intensional} definition such as \(x \mapsto 2x+3\), or a computer
-program.
+a textual name to an arbitrary value. I find it useful to see this as a
+mathematical ``function'' defined \emph{extensionally} by listing its
+input/output mappings --- this opposed to an \emph{intensional}
+definition such as \(x \mapsto 2x+3\), or a computer program.
 
@@
 equation
-\texttt{root\ (bicycle)\ (wheels)\ (spokes)\ (colour)\ =\ root\ (silver)}.
+
+\begin{lstlisting}
+root (bicycle) (wheels) (spokes) (colour) = root (silver)
+\end{lstlisting}
+
 That is, whatever object is the output of \texttt{silver} in the
@@
 function on the left) points to the same object. The ability to
-partition a system in this way enables what
-\href{https://github.com/amb26/papers/blob/master/ppig-2016a/ppig2016a.pdf}{2}
-calls a ``natural co-ordinate system'' for a piece of software, crucial
-for understanding and adaptability by others (so long as it is
-externalised).
+partition a system in this way enables what \cite{externalise} calls a
+``natural co-ordinate system'' for a piece of software, crucial for
+understanding and adaptability by others.
 
@@
 
-The latter is unfortunately still required in JavaScript for extensional
-functions with \emph{non-string} inputs.
+The latter style is unfortunately still required in JavaScript for
+extensional functions with \emph{non-string} inputs.
 
@@
 directly express the model I was thinking of (Figure
-\ref{fig:method-boxes}).
+\ref{fig:method-boxes}). So we can see this as polyfilling a mistake in
+the design of my substrate, rather than the Web platform underneath it.
 
@@
 frustrating. Because the system could only be patched externally and
-restarted, there needed to be some way to persist changes that
-\emph{weren't} part of the code, but the DOM instead.
+restarted, there needed to be some way to persist changes which live in
+the DOM, rather than the source files.
 
@@
 data required to reconstruct the system's current state is contained in
-the inspector HTML (plus the OROM/SVG JavaScript source code.) Where
-previously ``boxes'' were created first as invisible JavaScript objects
-responsible for some SVG, now it was the other way round. When a rect is
-clicked, the system must look at some SVG and interpret it ``on demand''
-as a box (lazily initialising its helper object if so).
+the inspector HTML (as well as the OROM/SVG JavaScript source code, of
+course.) Where previously ``boxes'' were created first as invisible JS
+objects responsible for some SVG, now it was the other way round. When a
+rect is clicked, the system must look at some SVG and interpret it ``on
+demand'' as a box (lazily initialising its helper object if so).
 
 Persistence seems to be a weaker cousin of \textbf{externalisability},
-defined in \href{antranig}{1}. As it stands, the system does not quite
+defined in \cite{externalise}. As it stands, the system does not quite
 qualify as externalisable. When making changes to the HTML in the
-element inspector, the system's behaviour \emph{should} adjust to match,
-but this is not currently guaranteed in all cases.
+element inspector, the system's behaviour \emph{ought} to adjust to
+match, but this is not currently guaranteed. Depending on the ability to
+listen for changes in node attributes or children, this may remain the
+case.
 
@@
 programmers. However, its purpose is to free a programmer's dependence
-on some distant and busy language designer, and it seems plausible that
-this is a stepping stone on the way to enabling programmers to
-\emph{create} a piece of software which non-programmers do not depend on
-them for.
+on some distant and busy\footnote{Not to mention unelected? ``Take Back
+  Control!'' (jokes aside, in this context it makes a lot of sense.)}
+language designer. It seems plausible that this is a step on the way to
+enabling \emph{creation} of a piece of software which non-programmers do
+not depend on us for.
 
@@
 behaviour of electromagnetic fields can be represented in four short
-equations that fit ``on a T-shirt''. A self-hosted LISP interpreter fits
-on a page. Could we aim at a similar ``fundamental description of
-software'' that fits on something less than millions of lines of code?
+equations that fit ``on a T-shirt''.
+
+A self-hosted LISP interpreter fits on a page. Could we aim at a similar
+``fundamental description of software'' that fits on something less than
+millions of lines of code?
 
@@
 
-We have a word ``re-factoring'', but what about the ``factoring'' in the
-first place? The previous discussion of ``ontologies''
-
 Further, there is perhaps a risk of optimising for \emph{formal} rather
 than \emph{practical} minimality (e.g.~the Turing machine). OROM's
-``minimality'' does not necessarily translate into ``simplicity''; it
+``minimality'' does not necessarily translate into ``simplicity''. It
 suffers from the same \emph{cognitive} complexity, or need for study, as
-Maxwell's Equations. I cannot stress the amount of effort I have put in
-to wrap my head around the self-referential ``vtable vtable'' and the
-task of self-implementation. An overall better system might be one which
-is easier to pick up or understand, even if the number of formal objects
-is not as minimal.
+Maxwell's Equations; I cannot stress enough the amount of effort I have
+put in to wrap my head around the self-referential ``vtable vtable'' and
+the task of self-implementation. An overall better system might be one
+which is easier to pick up or understand, even if the number of formal
+objects is not as minimal.
+
+Much to their credit, in the conclusion of \cite{OROM}, the authors note
+that ``it is not necessarily a friendly model for hand-written code'',
+suggesting its use as a compilation target. In a similar vein, it could
+function as the kernel of a much more familiar system (on the surface)
+--- it is, after all, supposed to be a vehicle that other things sit on,
+rather than the final user interface itself.
 
@@
 
-This is also cool, but hard.
-
-Other themes e.g.~homoiconicity of code and data
-
-``Taming'' SVG - ability to use SVG in a vaguely sane way (vector
-graphics editor)
-
-Similarly for e.g.~3D OpenGL, Web Audio, sockets \ldots{}
-
-\hypertarget{tensions-between-philosophies}{%
-\subsection{Tensions between
-philosophies}\label{tensions-between-philosophies}}
-
-\hypertarget{stable-or-emergent-antranig-against-oop-dispatch}{%
-\subsubsection{Stable or emergent? (Antranig against OOP
-dispatch)}\label{stable-or-emergent-antranig-against-oop-dispatch}}
-
-\hypertarget{turing-complete-antranig-against-this}{%
-\subsubsection{Turing-complete? (Antranig against
-this)}\label{turing-complete-antranig-against-this}}
-
-\hypertarget{messaging-vs-read-write}{%
-\subsubsection{Messaging vs Read-Write?}\label{messaging-vs-read-write}}
+The related paper \cite{COLAs} expands on OROM, giving it a
+``structural'' role complemented by a LISP-like programming language.
+Section 6.1 sketches out the intended ``bootstrapping'' process, ending
+with a self-sufficent, self-hosting version of the system.
+
+However, this task again suffers from the same cognitive complexity as
+any other self-referential circle. This is even more so, having made the
+complexity of graphics and interaction somehow still ``part of the
+system''. Figuring out exactly how the authors' bootstrapping process of
+their wholly \emph{language-based} system maps on to my task is full of
+unknowns at this point.
+
+Finally, in \cite{crit-semprola}, there is a warning against ``obsession
+with completely homogenous systems'' written in themselves. If this
+approach is doomed then obviously I want to take a different one, but
+the argument against it seems to hinge on what constitutes ``failure''.
+It considers Smalltalk and LISP as ``unsuitable'', yet the authors of
+OROM clearly think the opposite. The disagreement probably hinges on
+``unsuitable for \emph{what}?'' and it would be clearer if the following
+two questions were answered:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Is there a difference in (final, intermediate) goals between the two
+  views?
+\item
+  Could systems like OROM and COLAs plausibly succeed at their own
+  stated goals?
+\end{enumerate}
 
@@
 
-Conclusion: what I have that Maxwell's E doesn't.
+It takes a lot of work to create in a substrate that feels suited to the
+domain. Much of this work involves filling in the machinery necessary to
+an awful lot of software, rather than the ways one's domain
+\emph{differs from} this common case. I have explicitly highlighted the
+obstacles one is likely to meet when starting from an ordinary
+programming platform. I also showed how I overcame them with solutions
+intended to be domain-appropriate to their use, within the limits of the
+implementation language.
+
+OROM/SVG more or less realises my desired substrate for implementing
+OROM. However, there is one major area I failed to make
+domain-appropriate. Despite representing the ``data'' parts of the
+system as I wanted, the \emph{computational} parts of OROM were just
+transplanted into the text boxes as source code. I will reiterate that
+this is \emph{sometimes} suitable, but not always, and it would be worth
+exploring alternative ways to express some of it in the substrate I
+have. Having JavaScript code that the user can modify also seems to mess
+up the browser's debugger when stepping through it.
+
+Still, in this mostly-suitable substrate for OROM, I can continue with
+my project of seeing whether the \emph{allure} of \cite{COLAs} can be
+saved from the text-based ``hidden world'' limitation that pervades it.
+The obvious next step is attempting ``self-implementation''. This is
+desirable is because I still have not escaped my text editor. Any
+changes to my OROM/SVG substrate (of nested box drawing) require going
+back to the script files; I still cannot take advantage of the system I
+have developed, to ease its own development. In the words of
+\cite{COLAs}, I wish to make it so that the original JavaScript files
+can be ``jettisoned without remorse''.
+
+In Section \ref{retained-mode-vector-graphics}, I touched on how
+everyone with a browser has access to a powerful vector graphics editor
+(SVG) locked behind a completely inappropriate UI (the JS Console). It
+is similar for 3D graphics (WebGL), and sound and music (Web Audio).
+Similar observations are made about ``native'' OS apps in
+\cite{prog21-dyn}. But even the interactive JS Console is a step above
+batch-mode compilation, which is the interface that unlocks your
+operating system's range of functionalities. So I single out the far
+more ubiquitous Web platform as having more ``wasted potential''.
+
+I hope to use OROM/SVG to ``tame'' SVG and other JavaScript APIs, with
+some minimal on-demand visualisation in the large portion of the screen
+adjacent to the JS console. This is an attempt to generalise the ``rect
+controls'', which allow the obvious geometric properties of a
+\texttt{\textless{}rect\textgreater{}} to be directly manipulated
+(Section \ref{context-appropriate-ontologies}). Giving OROM access to
+Web technologies in this way is essential for evolving it into a
+\emph{fully-featured} self-changeable software environment, which could
+allow as much domain-specifc representation as its user wishes.

### 2020-01-22 convivial-2020/tex/manuscript.tex
DONE

@@
 \documentclass[english,submission]{programming}
-\usepackage[backend=biber]{biblatex} % Use Biblatex
-\addbibresource{example.bib}
+\usepackage[
+backend=biber,
+style=alphabetic
+]{biblatex} % Use Biblatex
+\addbibresource{../refs.bib}
 
@@
   perspective=art,
-  area={General-purpose programming}
+  area={Domain-specific programming, visual programming, self-sustaining systems}
 }
@@
    <concept>
-<concept_id>10002944.10011122.10003459</concept_id>
-<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
+       <concept_id>10011007.10011006.10011050.10011017</concept_id>
+       <concept_desc>Software and its engineering~Domain specific languages</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011050.10011058</concept_id>
+       <concept_desc>Software and its engineering~Visual languages</concept_desc>
        <concept_significance>300</concept_significance>
@@
 
-\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}`
+\ccsdesc[500]{Software and its engineering~Domain specific languages}
+\ccsdesc[300]{Software and its engineering~Visual languages}
 
@@
 
-\acks
-I want to thank ....
-
 \printbibliography

### 2020-05-13 convivial-2020/demo-talk/drivers.js
CCS2020 talk slides, notes, demo, recording

@@ -1,0 +1,26 @@
+svg.onmousedown = e => {
+  if (e.button === 0) change(left_mouse_button_is_down, true);
+};
+
+svg.onmouseup = e => {
+  if (e.button === 0) change(left_mouse_button_is_down, false);
+};
+
+// mousemove => pointer position changed
+svg.onmousemove = e => {
+  let r = svg.getBoundingClientRect();
+  let pos = vsub([e.clientX, e.clientY], [r.left, r.top]);
+  change(pointer.position, pos);
+};
+
+// mouseover => considering a new object
+svg.onmouseover = e => {
+  if (svg_userData(e.target) !== undefined)
+    change(pointer.is_considering_whom, obj);
+};
+
+// mouseout => no longer considering the object
+svg.onmouseout = e => {
+  if (svg_userData(e.target) !== undefined)
+    change(pointer.is_considering_whom), undefined);
+};

### 2020-05-13 convivial-2020/demo-talk/orom-svg.js

@@ -1,0 +1,1331 @@
+document.documentElement.style.height = '99%';
+
+body = document.body;
+body.style.margin = '0px';
+body.style.minHeight = '100%';
+
+log = (x, str) => {
+  if (str === undefined) console.log(x);
+  else console.log(str, x);
+  return x;
+};
+
+attr_single = (elem, key, val_or_func) => {
+  let old;
+  if (key === 'textContent') old = elem.textContent;
+  else old = elem.getAttribute(key);
+
+  let value = typeof(val_or_func) === 'function' ? val_or_func(old) : val_or_func;
+  if (key === 'textContent') elem.textContent = value;
+  else if (value !== undefined) elem.setAttribute(key, value);
+
+  return old;
+};
+
+// e.g. attr(rect, {stroke_width: 5, stroke: 'red'})
+//      attr(rect, 'stroke', 'red')
+//      attr(rect, 'height', h => h+32)
+//      attr(rect, {fill: 'orange', height: h => h+32})
+attr = (elem, key_or_dict, val_or_nothing) => {
+  if (typeof(key_or_dict) === 'string') {
+    let key = key_or_dict;
+    let value = val_or_nothing;
+    return attr_single(elem, key, value);
+  } else {
+    let dict = key_or_dict;
+    for (let [k,v_or_f] of Object.entries(dict)) {
+      let key = k.replace('_','-');
+      attr_single(elem, key, v_or_f);
+    }
+  }
+}
+
+nums = (arr) => arr.map(x => +x);
+attrs = (el, ...keys) => keys.map(k => attr(el, k));
+props = (o,  ...keys) => keys.map(k => o[k]);
+
+svg_parent = body; // Default parent for new SVG elements
+
+create_element = (tag, attrs, parent, namespace) => {
+  let elem = document.createElementNS(namespace, tag);
+  if (attrs !== undefined) attr(elem, attrs);
+  if (parent === undefined) parent = svg_parent;
+  parent.appendChild(elem);
+  return elem;
+};
+
+// e.g. rect = svgel('rect', {x: 5, y: 5, width: 5, height: 5}, svg)
+svgel = (tag, attrs, parent) => create_element(tag, attrs, parent, 'http://www.w3.org/2000/svg');
+
+htmlel = (tag, attrs, parent) => create_element(tag, attrs, parent, 'http://www.w3.org/1999/xhtml');
+
+state = function(o, k, v) {
+  let old = o[k];
[…]

### 2020-05-14 convivial-2020/manuscript.tex
More amendments

@@
+\newcommand{\OROM}{Id}
 \title{What does it take to create with domain-appropriate tools?}
-\subtitle{A case study on the ``OROM'' system}
+\subtitle{Reflections on implementing the ``Id{}'' system}
 \author[ ]{Joel Jakubovic}
@@
 \newcommand{\svgel}[1]{\texttt{\textless{}#1\textgreater{}}}
-\newcommand{\OSFA}{One-Size-Fits-All }
+\newcommand{\OSFA}{One-Size-Fits-All}
+\newcommand{\xywh}{\texttt{x},\texttt{y},\texttt{width},\texttt{height}}
 
 \begin{abstract}
-There is a One-Size-Fits-All quality to languages, APIs and even programming itself.
+There is a One-Size-Fits-All{} quality to languages, APIs and even programming itself.
 Whether you're making a mobile game or a scientific simulation, you will be using a text-based language with similar devices for structuring your code.
@@
 your reward is a system overall improved by a mere 0.5\%. Now, text
-coding is certainly ubiquitous, the 99\% case in programming, so you
-might wonder where I'm going with this. Well, a small improvement to
-text editing, if adopted by everyone, certainly does have a massive
-\emph{intermediate} effect --- but this only \emph{matters} to the
-extent that text was helping the programmers in the first place. If my
-goal is to draw or animate pictures, or create a digital synth from a
-frequency spectrogram, then giving me the ability to auto-indent my SVG
-markup is rather underwhelming as a productivity increase, as it doesn't
-target the core of the enterprise that makes it so hard.
-
-My experience of coding, most of the projects requiring shapes (such as
-GUIs), leads me to conclude that no matter how much I improve my skill
-at a particular language, knowledge of libraries or even general coding
-ability, my predicament stays the same. Our basic method of creating
-software is optimised for an ever-diminishing proportion of the software
-we actually want to make; ill-optimised for the graphics, layout,
-interactivity and and basic physics --- more on this later --- that we
-usually require.
-
-Whenever I work on these I feel stuck in a box I know I can never escape
+coding is certainly ubiquitous, the 99\% case in programming. A small
+improvement to text editing, if adopted by everyone, certainly does have
+a massive \emph{intermediate} effect---but this only \emph{matters} to
+the extent that text was helping us in the first place. If my goal is to
+draw or animate pictures, or create a digital synth from a frequency
+spectrogram, then giving me the ability to auto-indent my SVG markup is
+rather underwhelming as a productivity increase, as it doesn't target
+the core of the enterprise that makes it so hard.
+
+As a programmer, I often feel stuck in a box I know I can never escape
 from: that box is the text editor, a fixed conduit through which all
@@
 than it}. And then use that, to build something even better. But there
-is an enormous breadth and depth of philosophies here, along with many
-failed historical attempts to do better --- or at least, ones that
-failed to catch on. And even worse than this is that in my very
-\emph{language} here I am making the same mistake as the text editor ---
-speaking in unqualified terms of ``better'' and ``worse'' as if there
-really is a One-Size-Fits-All solution to software creation!
+is an enormous breadth and depth of philosophies here, along with all
+sorts of concrete systems that failed to catch on. Even worse than this,
+is that in my very \emph{language} here I am making the same mistake as
+the text editor---speaking in unqualified terms of ``better'' and
+``worse'' as if there really is a One-Size-Fits-All{} solution to
+software creation!
 
 Of course what we \emph{really} want is the ability for people to create
-\emph{in the way that they think is best} in their particular context
---- to equip them to feasibly create the tools that suit them for the
-thing they want to make. And second-order tools that suit them for
-making the first-order tools, and so on. It would do no good to replace
-text-imperialism with anything-else-imperialism, which is one
-interpretation of calls for alternatives.
+\emph{in the way that they think is best}\footnote{To be clear: if
+  someone \emph{wants} to type out pictures in ASCII, let them---whether
+  they do it for a challenge, or even if they find that more natural for
+  themselves. But equally, if I want to do it another way, I should have
+  that affordance.} in their particular context---to equip them to
+feasibly create the tools that suit them for the thing they want to
+make. And second-order tools that suit them for making the first-order
+tools, and so on. It would do no good to replace text-imperialism with
+anything-\emph{else}-imperialism, which is one interpretation of calls
+for alternatives.
 
@@
 (e.g.~languages, tools, or subject areas) each serving many clients, the
-One-Size-Fits-All style is the best one can hope for. Adaptation to
+One-Size-Fits-All{} style is the best one can hope for. Adaptation to
 individual preferences and idiosyncrasies is only feasible when those
@@
 of. And it is not so hard to make a specific, \emph{hard-baked} visual
-or alternative programming tool --- but it is hard to make it
+or alternative programming tool---but it is hard to make it
 re-programmable \emph{without} having to go back to \emph{its} textual
@@
 
-If someone wants to type out pictures in ASCII, let them --- whether
-they do it for a challenge, or even if they find that more natural for
-themselves. But equally, if I want to do it another way, then please
-give me that affordance. This is how I segue into the software artefact
-for which I have been attempting to build a natural representation.
-True, it is a programmer's artefact, but it is still representative of
-what any normal person has to do, insofar as:
+All these considerations attract me to the design of a particular system
+called ``Id'', as a potential way out of this mess \cite{OROM}. True, it
+is a programmer's artefact, but it is still representative of what any
+normal person has to do, insofar as:
 
@@
 
-\hypertarget{the-orom-system}{%
-\section{The OROM system}\label{the-orom-system}}
+What follows in this paper is an account of how I went about building
+Id, atop a preliminary ``box'' substrate.
+Section\textasciitilde{}\ref{the-system} gives an overview of Id.
+Section\textasciitilde{}\ref{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}
+discusses the work involved in obtaining said domain-appropriate
+substrate, and Section\textasciitilde{}\ref{patterns-and-polyfilling}
+continues with an interpretation of the labour costs of doing this. In
+Section\textasciitilde{}\ref{the-system-as-a-part-of-the-solution}, we
+return to the Id system and its role in supporting flexible software.
+Section\textasciitilde{}\ref{conclusions-and-future-work} addresses the
+question in the title, and concludes with next steps for the project.
+
+\hypertarget{the-system}{%
+\section{\texorpdfstring{The Id{}
+system}{The  system}}\label{the-system}}
 
@@
   conditions}, Smalltalk-style\footnote{OOP with more emphasis on object
-  instances and messaging (methods) as in a distributed system; less
-  emphasis on class hierarchies implementing traditional data structures}
-objects and messaging environment that the authors call ``Id'' --- but I
-shall refer to it as OROM, for pronunciation and Googleability.
-
-Essentially, an OROM object is a block of state which can change as a
-result of messages received by it. A message is sent by first
-\emph{bind}-ing its name to its \emph{implementation}: specific code,
-which is then run in the context of the receiver. ``Bind'' is
-accomplished via another message send, this time to the receiver's
-\emph{vtable}, or ``behaviour''. This is another object whose state
-defines a particular way of associating ``message name'' to
-``implementation code''. This continues up the vtable-chain, and
-terminates at a base case. The system as a whole is ``bootstrapped''
-into existence by its initialisation code. Each step of this code makes
-use of any parts of the system set up by the previous steps.
+  instances and messaging, as in a distributed system; less emphasis on
+  class hierarchies implementing traditional data structures} objects
+and messaging environment that the authors call ``Id''.
+
+An Id{} \emph{object} is a block of state which can change as a result
+of messages received by it. Messaging (analogous to \emph{method
+invocation} in mainstream OOP) works as follows:
+
+\begin{itemize}
+\tightlist
+\item
+  A message is sent by first \emph{bind}-ing its name to its
+  \emph{implementation}: specific code, which is then run in the context
+  of the receiver \(R\).
+\item
+  This ``bind'' step is accomplished by sending a further message; this
+  time, to the receiver's \emph{vtable} \(V(R)\). A vtable is another
+  object that maps ``message name'' to ``implementation code''---it's
+  analogous to a ``class'' in mainstream OOP.
+\item
+  This initial ``bind'' message triggers a similar ``bind'' to
+  \emph{its} vtable \(V(V(R))\), and so on: recursing up the
+  vtable-chain, and terminating at a base case.
+\item
+  The higher levels of the vtable-chain mean that different kinds of
+  vtables can be supported\footnote{As well as different kinds of
+    ``kinds of vtables'', and so on.}, each implementing the ``bind''
+  operation in its own way.
+\end{itemize}
+
+Inkeeping with its aims of minimality and self-describability, Id as a
+whole is ``bootstrapped'' into existence by its initialisation code.
+Each step of this code makes use of any parts of the system set up by
+the previous steps.
 
@@
 preferred these to have been in the paper in the first place, but I am
-not necessarily representative of those who read it. The
-One-Size-Fits-All approach is perhaps unavoidable for static media like
-print.
-
-Other areas like messaging semantics could still be quite confusing.
-After all, should we expect to be able to predict the entire future
-evolution of a dynamical system from a static description of its initial
-state, such as source code? Is this not why debuggers exist? Having
+not necessarily representative of those who read it\footnote{Besides
+  that, the One-Size-Fits-All{} approach is perhaps unavoidable for
+  static print media.}.
+
+At any rate, in order to solidify my understanding, there was no better
+way than to run the thing and explore it. And yet, having already
 ``de-compiled'' English text and C source code into object diagrams, it
 is a shame to have to compile it all back to struct member assignments.
-
-Worst of all, the reference system does not even have text I/O when it
-is run, let alone some sort of GUI. That is, the (un)intended user
-interface for this project is a C debugger! Faced with the necessity of
-adding \emph{some} UI, it seemed a waste of effort to end up with a
-system that must be continually polled for its current state at a
-terminal prompt. If I naturally think of this system as 2D tables, why
-can't that be how the running system looks? I do not have to keep
-polling my eyes for what state my diagrams are in.
-
-But further than that --- why can't the system be \emph{built} out of
-tables in the first place? Shouldn't this be the main takeaway from the
-amount of time we spend prototyping, explaining and designing software
-as diagrams on paper? Why must the ``natural representation'' be
-restricted to the finished product?
+Worse, the reference system does not even have text I/O when it is run,
+let alone some sort of GUI\footnote{That is, the (un)intended user
+  interface for Id ends up being a C debugger!}. Faced with the
+necessity of adding \emph{some} UI, it seemed a waste of effort to end
+up with a system that must be continually polled for its current state
+at a terminal prompt.
+
+If I naturally think of this system as 2D tables, why can't that be how
+the running system looks? I do not have to keep polling my eyes for what
+state my diagrams are in. But further than that---why can't the system
+be \emph{built} out of tables in the first place? Shouldn't this be the
+main takeaway from the amount of time we spend prototyping, explaining
+and designing software as diagrams on paper? Why must the ``natural
+representation'' be restricted to the finished product?
 
@@
 
-\hypertarget{orom-as-html-tables}{%
-\subsection{OROM as HTML tables}\label{orom-as-html-tables}}
+\hypertarget{as-html-tables}{%
+\subsection{\texorpdfstring{Id{} as HTML
+tables}{ as HTML tables}}\label{as-html-tables}}
 
@@
   \includegraphics[width=\textwidth]{../orom-html.png}
-  \caption{OROM/HTML: Obj-dicts are rendered as resizable tables, which reference
+  \caption{Id{}/HTML: Obj-dicts are rendered as resizable tables, which reference
            each other through numerical IDs. \label{fig:orom-html}}
@@
 
-To emphasise the tendency of OROM objects to be visualised as key-value
+To emphasise the tendency of Id{} objects to be visualised as key-value
 mappings, I will refer to them as \emph{obj-dicts}. Figure
-\ref{fig:orom-html} shows the OROM/HTML implementation \cite{orom-html},
+\ref{fig:orom-html} shows the Id{}/HTML implementation \cite{orom-html},
 in which they take the form of HTML tables.
@@
 \emph{directly} encode much of the system state in the DOM, achieving
-basic liveness\footnote{``The thing on the screen is the actual thing''}
-for the keys and values of obj-dicts.
+basic liveness (``The thing on the screen is the actual thing'') for the
+keys and values of obj-dicts.
 
@@
 
-A key aspect of the OROM system is that there is an object \emph{graph}.
+A key aspect of the Id{} system is that there is an object \emph{graph}.
 That is, obj-dicts can have entries pointing to other obj-dicts, without
 restriction to a tree structure. Drawing arrows to denote this is a
-no-brainer (and easy on paper), so I wanted it in my substrate for OROM.
-
-A small problem to surmount first: even though I could hijack the
-\texttt{\textless{}table\textgreater{}} for its display properties, what
-element could I hijack to make arrows between arbitrary points? Luckily,
-there were Scalable Vector Graphics (SVG) at my disposal, which could be
-persuaded to display \texttt{\textless{}line\textgreater{}}s over the
+no-brainer (and easy on paper), so I wanted it in my substrate for Id{}.
+
+Luckily, Scalable Vector Graphics (SVG) was at my disposal, which could
+be persuaded to display \texttt{\textless{}line\textgreater{}}s over the
 \texttt{\textless{}div\textgreater{}}s. But another key feature of my
@@
 
-\hypertarget{orom-as-svg-trees}{%
-\subsection{OROM as SVG trees}\label{orom-as-svg-trees}}
+\hypertarget{as-svg-trees}{%
+\subsection{\texorpdfstring{Id{} as SVG
+trees}{ as SVG trees}}\label{as-svg-trees}}
 
@@
   \includegraphics[width=\textwidth]{../orom-svg.png}
-  \caption{OROM/SVG: Obj-dicts are moveable and resizable nested boxes,
+  \caption{Id{}/SVG: Obj-dicts are moveable and resizable nested boxes,
            referencing each other via \emph{real arrows}.\label{fig:orom-svg}}
@@
 
-In OROM/SVG, shown in Figure \ref{fig:orom-svg}, obj-dicts are encoded
+In Id{}/SVG, shown in Figure \ref{fig:orom-svg}, obj-dicts are encoded
 as nested SVG \texttt{\textless{}rect\textgreater{}}s and other
-elements, reminiscent of diSessa's Boxer\cite{boxer}. This was a
+elements, reminiscent of diSessa's Boxer \cite{boxer}. This was a
 significant departure from the table representation, and even though SVG
@@
 themes that stand out to me from my development experience, hammered
-home by these OROM projects.
+home by these Id{} implementations.
 
@@
 In this section, I will present these and other seemingly
-\emph{inevitable} demands of normal software, exemplified by OROM/SVG. I
+\emph{inevitable} demands of normal software, exemplified by Id{}/SVG. I
 will comment on how our programming platforms measure up to the task,
@@
 that anyone with a browser \emph{in principle} has access to a powerful
-vector graphics editor --- just one with no GUI.
+vector graphics editor---just one with no GUI.
 
@@
 around them. Some of these, such as ``things fall down'', are not
-generally appropriate to software UIs --- perhaps because the screen has
-a role in our lives a more of a table work surface rather than a
-vertical wall, even if it is vertical in real life. The level of physics
-in software tends to not involve force, or mass, or very much at all,
+generally appropriate to software UIs---perhaps because the screen has a
+role in our lives a more of a table work surface rather than a vertical
+wall, even if it is vertical in real life. The level of physics in
+software tends to not involve force, or mass, or very much at all,
 merely position and space; we could call it ``geometric physics''.
@@
 \item
-  This number range \texttt{0000}--\texttt{FFFF} is like a space,
-  addresses = points
-\item
-  Every point has at most one owner block
-\item
-  These blocks are contiguous, finite ranges (i.e.~1D boxes)
+  This number range \texttt{0000}--\texttt{FFFF} is like a space
+  (addresses = points)
+\item
+  Every point has at most one owner block (no overlap)
+\item
+  These blocks are contiguous, finite ranges (1D boxes)
 \end{itemize}
 
-Hence the boilerplate involved to realise this in any domain with
-something resembling space. Far from being a niche topic in games and
-graphics, spatial partitioning algorithms and data structures have
-surprising relevance to more ordinary software. Both memory allocation
-and graphical layout are essential to today's; shame that only one of
-those has been recognised as such --- and entered the runtime of modern
+Far from being a niche topic in games and graphics, spatial partitioning
+algorithms and data structures have surprising relevance to more
+ordinary software. Both memory allocation and graphical layout are
+essential to today's; shame that only one of those has been recognised
+as such---and made part of the standard libraries of programming
 languages.
@@
 
-In OROM, the obvious candidate for this is the obj-dicts, plus all
-nested boxes in OROM/SVG. If I move the top-level rect, then I expect
+In Id{}, the obvious candidate for this is the obj-dicts, plus all
+nested boxes in Id{}/SVG. If I move the top-level rect, then I expect
 its children to move with it. This is simply the translational physics
@@
 \textbf{preserving the relationship over time}, which was a significant
-area of OROM/SVG.
+area of Id{}/SVG.
 
@@
 of live state, then this must form part of the standard boilerplate.
-Such was the case in OROM/SVG. I implemented the system in an OOP
-fashion, and the Observable class is the most widely used. It wraps a
-current value and a list of subscribers, notifying them when it changes.
+Such was the case in Id{}/SVG: the polyfilled ``Observable'' class is
+the most widely used. It wraps a current value and a list of
+subscribers, notifying them when it changes.
 
@@
 \emph{conceptually} very simple: an ``always equal'' relation. In
-OROM/SVG founded on live-state, this can be expressed in much the same
+Id{}/SVG founded on live-state, this can be expressed in much the same
 way:
@@
 We are conditioned to only think of these in their most general forms.
-But the ``vector graphics'' I use in OROM/SVG are just rects, lines,
+But the ``vector graphics'' I use in Id{}/SVG are just rects, lines,
 circles and text; a fraction of the full capability of SVG. The
@@
 Fully General anything. In fact, such generality might make it
-\emph{more} cumbersome to express what I wanted in OROM. As the old
+\emph{more} cumbersome to express what I wanted in Id{}. As the old
 wisdom goes, there's no point expressing a simple regex search as an
@@
 or a mass and I don't \emph{need} a linear optimisation solver for my
-space management --- for the time being. Under the theme of
+space management---for the time being. Under the theme of
 domain-appropriate tools, I think it is worth designing interfaces for
 \emph{smaller-scale} instances of these areas and exploring what they
-can express.
+are still capable of expressing.
 
@@
 tools available say ``batteries not included'', it is quite reasonable
-to want the batteries to be part of the platform. This section details
-the ``batteries not included'' I found myself polyfilling in OROM/SVG.
+to want the batteries to be part of the platform.
+Section\textasciitilde{}\ref{what-did-it-take-and-why} will expand on
+this further, but here I will list the ``batteries not included'' I
+found myself polyfilling in Id{}/SVG.
 
@@
 One can do this to all objects, even those of Web APIs. For example, I
-use it to connect SVG nodes to the OROM system: when a
+use it to connect SVG nodes to the Id{} system: when a
 \texttt{\textless{}rect\textgreater{}} is created, it is ``annotated''
@@
 This could be described as the box ``wrapping'' the
-\texttt{\textless{}rect\textgreater{}}, but risks impliying that the
+\texttt{\textless{}rect\textgreater{}}, but risks implying that the
 \texttt{\textless{}rect\textgreater{}} merely needs to \emph{live
 inside}, or \emph{be referenced by}, the box. In this case, the
-\emph{reverse} association is also needed, because clicks can only enter
-the system \emph{through} SVG in the first place. What is significant
-here is that JavaScript lets us add such a link to an entity that
-\emph{we did not design}.
+\emph{reverse} association is also needed (i.e.~mutual pointers); clicks
+can only enter the system \emph{through} SVG in the first place. What is
+significant here is that JavaScript lets us add such a link to an entity
+that \emph{we did not design}.
 
@@
 idea: to associate arbitrary data with an API object. But in languages
-that do not permit such ``annotation'', this has to be explicitly
-\emph{designed in} ahead of time. If it wasn't, then it has to be done
-in a roundabout way --- for example, a separate \texttt{userData} lookup
-table, indexed by the memory address of the API object.
+that do not permit such \emph{unanticipated} ``annotation'', this has to
+be explicitly \emph{designed in} ahead of time. If it wasn't, then it
+has to be done in a roundabout way---for example, a separate
+\texttt{userData} lookup table, indexed by the memory address of the API
+object.
 
@@
 \ref{maintaining-relationships-over-time}. This problem could be
-considered a microcosm of OROM/SVG: what's a natural way I
+considered a microcosm of Id{}/SVG: what's a natural way I
 \emph{conceive} of this behaviour, and could I implement it that way?
@@
 I like to see what I'm doing, so I wanted these rods to be visible and
-thus somehow \emph{present in the SVG}. This was not hard; in SVG I draw
-a \texttt{\textless{}line\textgreater{}}, though the background
-machinery of the Rod class does all the work. There is also a Point
-class that is used wherever manipulable points (SVG circles) are
-intended.
+thus somehow \emph{present in the SVG}. For this, I have a Point class
+that is used wherever manipulable points (SVG circles) are needed, and a
+Rod class that draws a \texttt{\textless{}line\textgreater{}} and
+transmits endpoint deltas across its length.
 
@@
 The remaining red rods are fully rigid.
-A line of CSS can show this ``scaffolding'' which is normally hidden for users.
+A line of CSS can reveal this ``scaffolding'' which is normally hidden for users.
 \label{fig:rods}}
@@
 infuriating to express.) Still, it was necessary in the case of certain
-elements --- especially those which must transcend the tree structure
-altogether, like arrows between boxes --- to bite this bullet, one way
-or another.
+elements---especially those which must transcend the tree structure
+altogether, like arrows between boxes---to bite this bullet, one way or
+another.
 
@@
 Each API has its own conventions, including a way of naming and
-structuring expressions --- an ontology \cite{crit-semprola}. The
-One-Size-Fits-All approach is exemplified in such interfaces. For
+structuring expressions---an ontology \cite{crit-semprola}. The
+One-Size-Fits-All{} approach is exemplified in such interfaces. For
 example, in SVG we express a rectangle by
@@
 The SVG specification \emph{defines} to the user that a rect simply
-\textbf{is} a top-left corner, a width, and a height --- and that's it.
+\textbf{is} a top-left corner, a width, and a height---and that's it.
 However, this SVG-approved parametrisation of a rectangle is far from
@@
 For example, I find it natural to resize boxes by dragging any of their
-four corners, so I wanted this in OROM/SVG. In this context, a
+four corners, so I wanted this in Id{}/SVG. In this context, a
 ``rectangle'' is \emph{seen as} four points: top-left, top-right,
@@
   top-left instead controls the parent
-  \texttt{\textless{}g\textgreater{}}'s \texttt{transform} --- but the
+  \texttt{\textless{}g\textgreater{}}'s \texttt{transform}---but the
   idea is the same.}; \texttt{width} and \texttt{height} are subscribed
@@
 physical device. And both keyboard keys and mouse buttons, for example,
-have two states --- pressed, not pressed --- which ought to make them
+have two states---pressed, not pressed---which ought to make them
 \emph{interchangeable} to some extent. Indeed, this is why there is such
@@
 some time-varying property, but instead as a sort of Cartesian product
-of event listeners. Rather than, say, a piece of live-state for the left
+of subroutines. Rather than, say, a piece of live-state for the left
 mouse button (LMB), we get \texttt{onmousedown} and \texttt{onmouseup}.
@@
 
-In OROM/SVG I did not quite want to re-map keys, but I did often want to
+In Id{}/SVG I did not quite want to re-map keys, but I did often want to
 have things follow the mouse when dragged. In order to do this, I
@@
 The way these things are connected to the browser's event listeners
-could be called ``device drivers'' --- an approach described in
+could be called ``device drivers''---an approach described in
 \cite{prog21-bbox}. It amounts to translating information from the Web's
@@
 came to this conclusion, I used to twist my head around translating my
-intention into the \texttt{x,y,width,height} parameters, and
+intention into the
+\texttt{x},\texttt{y},\texttt{width},\texttt{height}{} parameters, and
 un-translating when reading back the code I had produced. Once you get
@@
 things (i.e.~learning to code), it makes sense to simply expect to see
-more of it --- especially when you know you are new, surrounded by
+more of it---especially when you know you are new, surrounded by
 veterans who see no problem, and so on. Nowadays, I take the position
@@
 
-What would it look like to support multiple ontologies? There are
-perhaps two answers: anticipate the possibilities ahead of time, or
-support users adding their own. It is unclear if we can do better on the
-latter than simply ``support polyfilling''. As for the former: in
-general, anticipating the diversity of ways someone might look at the
-world is doomed to fail. But in the case of fairly \emph{formalised}
-concepts such as geometrical shapes or mathematics, I could suggest that
-simple \emph{under-specification} is the root of the problem in SVG.
-Instead of the specification explaining in English that ``\texttt{x} and
-\texttt{y} are the co-ordinates of the top left-hand corner\ldots{}'',
-it might be better to make these relations machine-readable or
-\emph{embodied} in the API. For the rect this might look like:
-
-\noindent A \texttt{rect} is\ldots{}
+What would it look like to support multiple ontologies? Na"ively, there
+are two answers: anticipate the possibilities ahead of time, or support
+users adding their own. It is unclear if we can do better on the latter
+than simply ``support polyfilling''. As for the former, anticipating the
+diversity of ways someone might look at the world is generally doomed to
+fail. But in the special case of fairly \emph{formalised} concepts such
+as geometrical shapes or mathematics, I could suggest that simple
+\emph{under-specification} is the root of the problem in SVG. Instead of
+the specification explaining in English that ``\texttt{x} and \texttt{y}
+are the co-ordinates of the top left-hand corner\ldots{}'', it might be
+better to make these relations machine-readable or \emph{embodied} in
+the API. For the rect this might look like:
 
@@
 \tightlist
+\item
+  A \texttt{rect} is\ldots{}
 \item
@@
 
-The hope is that if we then specify enough information --- say, the
-bot-left and top-right --- then the runtime has all it needs to derive
-its internal 4 degrees of freedom.
+The hope is that if we then specify enough information---say, the
+bot-left and top-right---then the runtime has all it needs to derive its
+internal 4 degrees of freedom.
 
@@
 
+There is, in fact, a part of the SVG specification \cite{svg-rect} that
+comes \emph{frustratingly close} to the above. It spells out how the
+\texttt{\textless{}rect\textgreater{}} parameters ``reduce'' to more
+primitive line-drawing steps, including expressions for the four corners
+in terms of \texttt{x},\texttt{y},\texttt{width},\texttt{height}{}. The
+problem is, this reduction is still described in \emph{natural language}
+for a human implementor, missing the potential of such a
+formulation\footnote{The point here is that a machine-readable
+  description could easily remain human-readable, but a natural-language
+  description is not easily machine-readable.}.
+
+It is worth admitting that there is is a significant shortcoming of my
+polyfill here, and perhaps throughout. It has taken the form of
+\emph{overriding} the default ontology, \emph{replacing} it with the one
+I preferred. From my perspective, this is fine. But if another person
+were to join in the development of Id{}/SVG, they would be stuck in the
+same situation I was in, unless they fully agreed with my ontologies.
+
+The desideratum of ``ontology co-existence'' is very much an open
+research problem. \cite{crit-semprola} gives a taste of the thorny
+complications unaddressed by a ``Semiotic Programming'' solution
+\cite{semprola}, while \cite{kell-c} points out how the idea of
+``linking'' is unconcerned with whatever language compiled to the object
+code; language-agnosticism is one salient example of ontology
+co-existence. This is especially clear if we agree with the latter
+author's view that ``which language'' ought to be an
+\emph{implementation detail}, that languages ought to be treatable as
+\emph{different views} onto the same thing.
+
+These considerations did not occur to me because this project is very
+idiosyncratic to my interests, at least in its initial stage. I always
+worked on the assumption that I was the only developer. However, a
+future Id{} system in a mature state should by no means be bound to this
+fact. It is also worth considering \emph{myself in the future} as a
+different agent, with different ontology preferences. In that case, the
+method here will need to be re-examined.
+
 \hypertarget{extensional-functions}{%
@@
 mathematical ``function'' defined \emph{extensionally} by listing its
-input/output mappings --- this opposed to an \emph{intensional}
-definition such as \(x \mapsto 2x+3\), or a computer program.
+input/output mappings---this opposed to an \emph{intensional} definition
+such as \(x \mapsto 2x+3\), or a computer program.
 
@@
 it's necessary to bring in a library or clutter the code with a
-home-grown approximation to dynamic ones. Some parts of the OROM
+home-grown approximation to dynamic ones. Some parts of the Id{}
 authors' C code were confusing until I realised they were just the guts
@@
 
-is more WYSIWYG\footnote{What You See Is What You Get} than the
-imperative-style
+is more ``What You See Is What You Get'' than the imperative-style
 
@@
 
-The decision to switch from the flat tabular representation in OROM/HTML
-to the nestable, Boxer-like structure in OROM/SVG permitted extensional
-(tree) functions in its substrate. True, I could always encode these in
-flat tables by having them reference each other. But this is like the
-second listing above: hard to follow. In the case of distinguishing
-between ordinary state mappings and the ``method dictionary'' of
-vtables, I actually stored method mappings with a ``-'' character to
-avoid doing this (Figure \ref{fig:method-hyphens}). In OROM/SVG, I can
-directly express the model I was thinking of (Figure
-\ref{fig:method-boxes}). So we can see this as polyfilling a mistake in
-the design of my substrate, rather than the Web platform underneath it.
+My original flat, tabular representation in Id{}/HTML did support a
+single level of such ``extensional functions'', but no more. In other
+words, the output of such a mapping could not be a further mapping
+itself. This was problematic because vtables, conceptually, have two
+levels of such nesting: the top-level fields, and the ``methods
+dictionary'' which contains its own input-output pairs. To distinguish
+between these, I actually stored method mappings with an ad-hoc ``-''
+character (Figure\textasciitilde{}\ref{fig:method-hyphens}).
+
+By contrast, after switching to the fully nestable box substrate in
+Id{}/SVG, I could directly express the model I was thinking of (Figure
+\ref{fig:method-boxes}). If I had stayed in HTML, there is a good chance
+I would have had to polyfill in extensional functions---whether through
+nested \texttt{\textless{}table\textgreater{}} elements, additional
+``-'' characters or something else.
 
@@
   \includegraphics[width=0.5\textwidth]{../method-hyphens.png}
-  \caption{OROM/HTML: use of hyphenated \texttt{-lookup} to signify a \emph{method} called
+  \caption{Id{}/HTML: use of hyphenated \texttt{-lookup} to signify a \emph{method} called
            \texttt{lookup} instead of a property like \texttt{vtable}. \label{fig:method-hyphens}}
@@
   \includegraphics[width=0.5\textwidth]{../method-boxes.png}
-  \caption{OROM/SVG: the \texttt{vtable} box has a \texttt{methods} box, and that's where the method boxes go. \label{fig:method-boxes}}
+  \caption{Id{}/SVG: the \texttt{vtable} box has a \texttt{methods} box, and that's where the method boxes go. \label{fig:method-boxes}}
 \end{figure}
@@
 
-Persistence was absolutely necessary to continue OROM/SVG development,
+Persistence was absolutely necessary to continue Id{}/SVG development,
 past a certain point. This is because, upon discovering a bug and fixing
@@
 substrate matured, and I began to implement parts of the target system
-(OROM) --- the cycle of finding a bug, tearing down the system,
-refreshing and losing work, and manually building it up again, proved
-frustrating. Because the system could only be patched externally and
-restarted, there needed to be some way to persist changes which live in
-the DOM, rather than the source files.
+(Id{})---the cycle of finding a bug, tearing down the system, refreshing
+and losing work, and manually building it up again, proved frustrating.
+Because the system could only be patched externally and restarted, there
+needed to be some way to persist changes which live in the DOM, rather
+than the source files.
 
@@
 data required to reconstruct the system's current state is contained in
-the inspector HTML (as well as the OROM/SVG JavaScript source code, of
-course.) Where previously ``boxes'' were created first as invisible JS
-objects responsible for some SVG, now it was the other way round. When a
-rect is clicked, the system must look at some SVG and interpret it ``on
-demand'' as a box (lazily initialising its helper object if so).
+the inspector HTML, as opposed to hidden JS properties. Where previously
+``boxes'' were created first as invisible JS objects responsible for
+some SVG, now it was the other way round. When a rect is clicked, the
+system must look at some SVG and interpret it ``on demand'' as a box
+(although this helper object, once created, can be cached in the SVG
+nodes).
 
@@
 
-\hypertarget{the-orom-system-as-a-part-of-the-solution}{%
-\section{The OROM system as a part of the
-solution}\label{the-orom-system-as-a-part-of-the-solution}}
-
-The OROM system was designed with a goal of eliminating the
+\hypertarget{the-system-as-a-part-of-the-solution}{%
+\section{\texorpdfstring{The Id{} system as a part of the
+solution}{The  system as a part of the solution}}\label{the-system-as-a-part-of-the-solution}}
+
+The Id{} system was designed with a goal of eliminating the
 ``artificial'' distinction between implementation language and end-user
@@
 means of itself'' is easy to agree with, even if OOP is not to
-everyone's taste. It is an open question whether OROM's approach could
+everyone's taste. It is an open question whether Id{}'s approach could
 be applied to other programming styles.
@@
 that it paves the way to end-user empowerment. This is somewhat
-misleading because in OROM, these ``end-users'' are actually
+misleading because in Id{}, these ``end-users'' are actually
 programmers. However, its purpose is to free a programmer's dependence
-on some distant and busy\footnote{Not to mention unelected? ``Take Back
-  Control!'' (jokes aside, in this context it makes a lot of sense.)}
-language designer. It seems plausible that this is a step on the way to
-enabling \emph{creation} of a piece of software which non-programmers do
-not depend on us for.
-
-I emphasise that this is the \emph{allure} of OROM. However, there are
+on some distant and busy\footnote{Not to mention \emph{unelected?}
+  ``Take Back Control!'' (jokes aside, in this context it makes a lot of
+  sense.)} language designer. It seems plausible that this is a step on
+the way to enabling \emph{creation} of a piece of software which
+non-programmers do not depend on us for.
+
+I emphasise that this is the \emph{allure} of Id{}. However, there are
 some significant practical issues that must be overcome or clarified
@@
 
-The OROM paper was part of the STEPS project of Alan Kay's
+The Id{} paper was part of the STEPS project of Alan Kay's
 VPRI\footnote{Viewpoints Research Institute}. This project argued that
@@
 learning material involved most certainly does \emph{not} fit on a
-T-shirt. The obvious \emph{reductio ad absurdudum} is where we
-encapsulate these equations under a single symbol, M. M is defined as
-``Maxwell's Equations are true''. Ta-da --- this fits on a coin, but
-good luck doing anything with it.\footnote{In fact, this is almost
-  achieved by the formalism of Geometric Calculus, which reduces them to
-  only one equation.}
-
-I say this not to dismiss the argument, but to highlight the actually
-hard part of getting a ``concise description'' of some system; defining
-complexity away into a symbol helps us no more than naming the solution
-to an equation ``x''. There is a connection with data compression: even
-if the data successfully compressed into a smaller file, the size of the
-compression \emph{program} should be added as well. What matters is to
-reduce the \emph{combined} size of notation and platform.
+T-shirt. The obvious \emph{reductio ad absurdum} is where we encapsulate
+these equations under a single symbol, \(\mathcal{M}\). \(\mathcal{M}\)
+is defined as ``Maxwell's Equations are true''. Ta-da---this fits on a
+coin, but good luck doing anything with it.\footnote{In fact, this is
+  almost achieved by the formalism of Geometric Calculus, which reduces
+  them to only one equation.}
+
+I say this not to dismiss the argument, but to highlight the
+\emph{actually hard part} of getting a ``concise description'' of some
+system; defining complexity away into a symbol helps us no more than
+naming the solution to an equation ``\(x\)''. There is a connection with
+data compression: even if the data is successfully compressed into a
+smaller file, the size of the compression \emph{program} should be added
+as well. What matters is to reduce the \emph{combined} size of notation
+and substrate.
 
 Further, there is perhaps a risk of optimising for \emph{formal} rather
-than \emph{practical} minimality (e.g.~the Turing machine). OROM's
+than \emph{practical} minimality (e.g.~the Turing machine). Id{}'s
 ``minimality'' does not necessarily translate into ``simplicity''. It
@@
 
-Much to their credit, in the conclusion of \cite{OROM}, the authors note
-that ``it is not necessarily a friendly model for hand-written code'',
-suggesting its use as a compilation target. In a similar vein, it could
-function as the kernel of a much more familiar system (on the surface)
---- it is, after all, supposed to be a vehicle that other things sit on,
-rather than the final user interface itself.
+In the conclusion of \cite{OROM}, the authors note that ``it is not
+necessarily a friendly model for hand-written code'', suggesting its use
+as a compilation target. In a similar vein, it could function as the
+kernel of a much more familiar system (on the surface)---it is, after
+all, supposed to be a vehicle that other things sit on, rather than the
+final user interface itself. In fact, this is precisely the approach set
+out in the related paper \cite{COLAs}.
 
@@
 
-The related paper \cite{COLAs} expands on OROM, giving it a
-``structural'' role complemented by a LISP-like programming language.
-Section 6.1 sketches out the intended ``bootstrapping'' process, ending
-with a self-sufficent, self-hosting version of the system.
+Said related ``COLAs'' paper expands on Id{}, giving it a ``structural''
+role complemented by a LISP-like programming language. Section 6.1
+sketches out the intended ``bootstrapping'' process, ending with a
+self-sufficent, self-hosting version of the system.
 
 However, this task again suffers from the same cognitive complexity as
-any other self-referential circle. This is even more so, having made the
-complexity of graphics and interaction somehow still ``part of the
-system''. Figuring out exactly how the authors' bootstrapping process of
-their wholly \emph{language-based} system maps on to my task is full of
-unknowns at this point.
+any other self-referential circle. This is even more so for me, having
+made the complexity of graphics and interaction somehow still ``part of
+the system''. Figuring out exactly how the authors' bootstrapping
+process of their wholly \emph{language-based} system maps on to my task
+is, as far as I know, uncharted territory.
 
@@
 It considers Smalltalk and LISP as ``unsuitable'', yet the authors of
-OROM clearly think the opposite. The disagreement probably hinges on
+Id{} clearly think the opposite. The disagreement probably hinges on
 ``unsuitable for \emph{what}?'' and it would be clearer if the following
@@
 \item
-  Could systems like OROM and COLAs plausibly succeed at their own
+  Could systems like Id{} and COLAs plausibly succeed at their own
   stated goals?
@@
 
-\hypertarget{conclusion-and-future-work}{%
-\section{Conclusion and future work}\label{conclusion-and-future-work}}
+\hypertarget{conclusions-and-future-work}{%
+\section{Conclusions and future
+work}\label{conclusions-and-future-work}}
+
+\hypertarget{what-did-it-take-and-why}{%
+\subsection{What did it take, and why?}\label{what-did-it-take-and-why}}
 
@@
 an awful lot of software, rather than the ways one's domain
-\emph{differs from} this common case. I have explicitly highlighted the
-obstacles one is likely to meet when starting from an ordinary
-programming platform. I also showed how I overcame them with solutions
-intended to be domain-appropriate to their use, within the limits of the
-implementation language.
-
-OROM/SVG more or less realises my desired substrate for implementing
-OROM. However, there is one major area I failed to make
+\emph{differs from} this common case.
+
+In an ideal world, there would be no accidental complexity, no
+boilerplate, no polyfilling. But this can't be the same as saying there
+would be no \emph{work} to do---that sounds like a world where
+\emph{essential} complexity has also been removed.
+
+Instead, we can note that software ecosystems typically centre around a
+``core'', such as a language, with a surrounding ``periphery'', such as
+libraries not bundled with the core language. The core (language syntax,
+semantics, standard libraries) acts as a common platform uniting the
+community; the periphery consists of functionality that is not
+``built-in'' to this core, both realised as software (e.g.~libraries),
+and that which is as-of-yet unrealised or nonexistent.
+
+There is a constant process of negotiation determining what should be
+part of the shared core versus the responsibility of the periphery. But
+at any given time, every community draws the line somewhere.
+
+To a first approximation, what is ``built-in'' encompasses the features
+and concerns that are \emph{common} to the entire community. Features
+that are of minority or niche use belong outside; it is one's own
+responsibility to implement them, or to integrate with an existing
+implementation.
+
+Perhaps in an ideal world, someone could realise their vision in
+software by focusing all their effort on what makes their idea
+\emph{unique}, or what is \emph{idiosyncratic} to it. In other words,
+they would only be responsible for that part of their idea which:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  It would be unrealistic to expect the wider community to have
+  \emph{thought of already}, and
+\item
+  It would be unfair to expect the wider community to \emph{build on his
+  or her behalf.}
+\end{enumerate}
+
+If the core Web platform (particularly JavaScript) is meant \emph{for}
+supporting ``web apps'', and this is indeed how it is used, then it
+seems it has not yet fully absorbed the functionality common to this
+usage. At present, this is simply a surprise, or a puzzle, and it might
+simply be a matter of time before it catches up.
+
+Nevertheless, I wanted to explicitly highlight the obstacles one is
+likely to meet when starting from this platform. They include not only
+those specific to the Web, but also issues that occur far more widely
+(e.g.~monopoly ontologies.) I showed how, in the short-term, we can
+overcome these with solutions intended to be domain-appropriate to their
+use (within the syntactic limits of JavaScript).
+
+\hypertarget{what-next}{%
+\subsection{What next?}\label{what-next}}
+
+Id{}/SVG more or less realises my desired substrate for implementing
+Id{}. However, there is one major area I failed to make
 domain-appropriate. Despite representing the ``data'' parts of the
-system as I wanted, the \emph{computational} parts of OROM were just
+system as I wanted, the \emph{computational} parts of Id{} were just
 transplanted into the text boxes as source code. I will reiterate that
@@
 
-Still, in this mostly-suitable substrate for OROM, I can continue with
+Still, in this mostly-suitable substrate for Id{}, I can continue with
 my project of seeing whether the \emph{allure} of \cite{COLAs} can be
@@
 desirable is because I still have not escaped my text editor. Any
-changes to my OROM/SVG substrate (of nested box drawing) require going
+changes to my Id{}/SVG substrate (of nested box drawing) require going
 back to the script files; I still cannot take advantage of the system I
@@
 Similar observations are made about ``native'' OS apps in
-\cite{prog21-dyn}. But even the interactive JS Console is a step above
-batch-mode compilation, which is the interface that unlocks your
-operating system's range of functionalities. So I single out the far
-more ubiquitous Web platform as having more ``wasted potential''.
-
-I hope to use OROM/SVG to ``tame'' SVG and other JavaScript APIs, with
+\cite{prog21-dyn}. Unfortunately for that domain, the interface that
+unlocks your operating system's range of functionalities---batch-mode
+compilation---is quite far from the affordances of the (interactive) JS
+Console.
+
+I hope to use Id{}/SVG to ``tame'' SVG and other JavaScript APIs, with
 some minimal on-demand visualisation in the large portion of the screen
-adjacent to the JS console. This is an attempt to generalise the ``rect
+next to the JS console. This is an attempt to generalise the ``rect
 controls'', which allow the obvious geometric properties of a
 \texttt{\textless{}rect\textgreater{}} to be directly manipulated
-(Section \ref{context-appropriate-ontologies}). Giving OROM access to
+(Section \ref{context-appropriate-ontologies}). Giving Id{} access to
 Web technologies in this way is essential for evolving it into a

### 2020-05-14 convivial-2020/manuscript.tex
No more Pandoc tilde BS, further rework

@@
 several diagrams. But to understand it, I repeatedly found myself
-drawing \emph{extra} diagrams. For example, the first acts of the
-running system boil down to initialising the three or so objects. This
-consists of allocating memory, interpreting it as a C struct and then
-filling in fields in a mundane manner. I had great difficulty following
-the specifics in my head, but when I drew tables in the style of their
-diagrams I readily saw what what going on. I personally would have
-preferred these to have been in the paper in the first place, but I am
-not necessarily representative of those who read it\footnote{Besides
-  that, the One-Size-Fits-All{} approach is perhaps unavoidable for
-  static print media.}.
+drawing \emph{extra} diagrams.
+
+For example, the first acts of the running system boil down to
+initialising the three or so objects. This consists of allocating
+memory, interpreting it as a C struct and then filling in fields in a
+mundane manner. I had great difficulty following the specifics in my
+head, but when I drew tables in the style of their diagrams I readily
+saw what what going on. I personally would have preferred these to have
+been in the paper in the first place, but I am not necessarily
+representative of those who read it\footnote{Besides that, the
+  One-Size-Fits-All{} approach is perhaps unavoidable for static print
+  media.}.
 
@@
   \centering
-  \includegraphics[width=\textwidth]{../orom-html.png}
-  \caption{Id{}/HTML: Obj-dicts are rendered as resizable tables, which reference
+  \includegraphics[width=\textwidth]{../orom-html-cropped.png}
+  \caption{Part of Id{}/HTML: Obj-dicts are rendered as resizable tables, which reference
            each other through numerical IDs. \label{fig:orom-html}}
@@
 To emphasise the tendency of Id{} objects to be visualised as key-value
-mappings, I will refer to them as \emph{obj-dicts}. Figure
-\ref{fig:orom-html} shows the Id{}/HTML implementation \cite{orom-html},
-in which they take the form of HTML tables.
+mappings, I will refer to them as \emph{obj-dicts}.
+Figure\textasciitilde{}\ref{fig:orom-html} shows the Id{}/HTML
+implementation \cite{orom-html}, in which they take the form of HTML
+tables.
 
@@
   \centering
-  \includegraphics[width=\textwidth]{../orom-svg.png}
-  \caption{Id{}/SVG: Obj-dicts are moveable and resizable nested boxes,
+  \includegraphics[width=\textwidth]{../orom-svg-cropped.png}
+  \caption{Part of Id{}/SVG: Obj-dicts are moveable and resizable nested boxes,
            referencing each other via \emph{real arrows}.\label{fig:orom-svg}}
@@
 
-In Id{}/SVG, shown in Figure \ref{fig:orom-svg}, obj-dicts are encoded
-as nested SVG \texttt{\textless{}rect\textgreater{}}s and other
-elements, reminiscent of diSessa's Boxer \cite{boxer}. This was a
-significant departure from the table representation, and even though SVG
-supports (some) nested HTML via
-\texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
-the possibility of multiple levels of nesting.
+In Id{}/SVG, shown in Figure\textasciitilde{}\ref{fig:orom-svg},
+obj-dicts are encoded as nested SVG
+\texttt{\textless{}rect\textgreater{}}s and other elements, reminiscent
+of diSessa's Boxer \cite{boxer}. This was a significant departure from
+the table representation, and even though SVG supports (some) nested
+HTML via \texttt{\textless{}foreignObject\textgreater{}}, I actually
+preferred the possibility of multiple levels of nesting.
 
@@
 display they can perceive. So right away it is going to require ways to
-draw coloured shapes. There are usually libraries for this (though note:
-not part of the language), but some only provide \emph{immediate mode}:
-commands to instantaneously rasterise pixels to a buffer. This is not
-enough for modern software, as we often expect animation, or at least to
-see things change as we interact. Most often we wish to see \emph{small
-changes} to the \emph{same} shapes, rather than completely different
-shapes altogether; the reification that this requires to persist between
-frames, is known as \emph{retained mode.}
+draw coloured shapes. There are usually libraries for this (though they
+aren't always \emph{standard} libraries), but some only provide
+\emph{immediate mode}: commands to instantaneously rasterise pixels to a
+buffer. This is not enough for modern software, as we often expect
+animation, or at least to see things change as we interact. Most often
+we wish to see \emph{small changes} to the \emph{same} shapes, rather
+than completely different shapes altogether; the reification that this
+requires to persist between frames, is known as \emph{retained mode.}
 
@@
 rendering loop. Alternatively, there is the W3C's chosen ontology of CSS
-animations, but see Section \ref{context-appropriate-ontologies}.
+animations, but see
+Section\textasciitilde{}\ref{context-appropriate-ontologies}.
 
@@
 endpoint. Since they both feel the same deltas, the displacement vector
-between them is preserved (Figure \ref{fig:delta-transmission}).
+between them is preserved
+(Figure\textasciitilde{}\ref{fig:delta-transmission}).
 
@@
 deltas'' in the vertical and horizontal, ``absorbing'' the other
-component into itself (Figure \ref{fig:rods}). Mirroring a DOM rect to
-these rods is as simple as subscribing its width and height to
-horizontal and vertical rods' \texttt{length} Observables. This way,
-boxes can be resized from whatever corner is convenient.
+component into itself (Figure\textasciitilde{}\ref{fig:rods}). Mirroring
+a DOM rect to these rods is as simple as subscribing its width and
+height to horizontal and vertical rods' \texttt{length} Observables.
+This way, boxes can be resized from whatever corner is convenient.
 
@@
 to children. It is nice to express the relation (as well as see it!)
-this way (Figure \ref{fig:rods}).
+this way (Figure\textasciitilde{}\ref{fig:rods}).
 
@@
 Each API has its own conventions, including a way of naming and
-structuring expressions---an ontology \cite{crit-semprola}. The
-One-Size-Fits-All{} approach is exemplified in such interfaces. For
-example, in SVG we express a rectangle by
+structuring expressions: an \emph{ontology} \cite{crit-semprola}. The
+One-Size-Fits-All{} approach is exemplified in such interfaces.
+
+In SVG, we express a rectangle by
 \texttt{\textless{}rect x="10" y="10" width="600" height="400"\textgreater{}}.
@@
 
-For example, I find it natural to resize boxes by dragging any of their
-four corners, so I wanted this in Id{}/SVG. In this context, a
-``rectangle'' is \emph{seen as} four points: top-left, top-right,
-bottom-left, bottom-right. Obviously this is not a \emph{minimal}
-description, since given e.g.~the top-left and bottom-right, the other
-two points can be inferred. But one way or another, to be able to drag
-any of them, all four points must be present at \emph{some} level. This
-alternative ontology was polyfilled in the form of a ``rect controls''
-class that can be attached manually to any SVG
+I find it natural to resize boxes by dragging any of their four corners,
+so I wanted this in Id{}/SVG. In this context, a ``rectangle'' is
+\emph{seen as} four points: top-left, top-right, bottom-left,
+bottom-right. Obviously this is not a \emph{minimal} description, since
+given e.g.~the top-left and bottom-right, the other two points can be
+inferred. But one way or another, to be able to drag any of them, all
+four points must be present at \emph{some} level. This alternative
+ontology was polyfilled in the form of a ``rect controls'' class that
+can be attached manually to any SVG
 \texttt{\textless{}rect\textgreater{}}. The \texttt{x} and \texttt{y}
@@
 
-What would it look like to support multiple ontologies? Na"ively, there
+What would it look like to support multiple ontologies? Naïvely, there
 are two answers: anticipate the possibilities ahead of time, or support
@@
 By contrast, after switching to the fully nestable box substrate in
-Id{}/SVG, I could directly express the model I was thinking of (Figure
-\ref{fig:method-boxes}). If I had stayed in HTML, there is a good chance
-I would have had to polyfill in extensional functions---whether through
-nested \texttt{\textless{}table\textgreater{}} elements, additional
-``-'' characters or something else.
+Id{}/SVG, I could directly express the model I was thinking in. Quite
+simply, the \texttt{vtable} box has a \texttt{methods} box, and that's
+where the method boxes go (Figure\textasciitilde{}\ref{fig:orom-svg}).
+If I had stayed in HTML, there is a good risk I would have had to
+polyfill in extensional functions---whether through nested
+\texttt{\textless{}table\textgreater{}} elements, additional ``-''
+characters, or something else.
 
@@
            \texttt{lookup} instead of a property like \texttt{vtable}. \label{fig:method-hyphens}}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.5\textwidth]{../method-boxes.png}
-  \caption{Id{}/SVG: the \texttt{vtable} box has a \texttt{methods} box, and that's where the method boxes go. \label{fig:method-boxes}}
 \end{figure}
@@
   rejected for security.}, so the solution I turned to was manually
-copying the markup in the browser's inspector (Figure
-\ref{fig:html-inspector}).
+copying the markup in the browser's inspector
+(Figure\textasciitilde{}\ref{fig:html-inspector}).
 
@@
 T-shirt. The obvious \emph{reductio ad absurdum} is where we encapsulate
-these equations under a single symbol, \(\mathcal{M}\). \(\mathcal{M}\)
-is defined as ``Maxwell's Equations are true''. Ta-da---this fits on a
-coin, but good luck doing anything with it.\footnote{In fact, this is
-  almost achieved by the formalism of Geometric Calculus, which reduces
-  them to only one equation.}
+these equations under a single symbol\footnote{In fact, this is almost
+  achieved by the formalism of Geometric Calculus, which reduces them to
+  only one equation.}, \(\mathcal{M}\). \(\mathcal{M}\) is defined as
+``Maxwell's Equations are true''. Ta-da---this fits on a coin, but good
+luck doing anything with it.
 
@@
 Said related ``COLAs'' paper expands on Id{}, giving it a ``structural''
-role complemented by a LISP-like programming language. Section 6.1
-sketches out the intended ``bootstrapping'' process, ending with a
-self-sufficent, self-hosting version of the system.
+role complemented by a LISP-like programming language.
+Section\textasciitilde{}6.1 sketches out the intended ``bootstrapping''
+process, ending with a self-sufficent, self-hosting version of the
+system.
 
@@
 
-In Section \ref{retained-mode-vector-graphics}, I touched on how
-everyone with a browser has access to a powerful vector graphics editor
-(SVG) locked behind a completely inappropriate UI (the JS Console). It
-is similar for 3D graphics (WebGL), and sound and music (Web Audio).
-Similar observations are made about ``native'' OS apps in
+In Section\textasciitilde{}\ref{retained-mode-vector-graphics}, I
+touched on how everyone with a browser has access to a powerful vector
+graphics editor (SVG) locked behind a completely inappropriate UI (the
+JS Console). It is similar for 3D graphics (WebGL), and sound and music
+(Web Audio). Similar observations are made about ``native'' OS apps in
 \cite{prog21-dyn}. Unfortunately for that domain, the interface that
@@
 
-I hope to use Id{}/SVG to ``tame'' SVG and other JavaScript APIs, with
+I intend to use Id{}/SVG to ``tame'' SVG and other JavaScript APIs, with
 some minimal on-demand visualisation in the large portion of the screen
@@
 \texttt{\textless{}rect\textgreater{}} to be directly manipulated
-(Section \ref{context-appropriate-ontologies}). Giving Id{} access to
-Web technologies in this way is essential for evolving it into a
-\emph{fully-featured} self-changeable software environment, which could
-allow as much domain-specifc representation as its user wishes.
+(Section\textasciitilde{}\ref{context-appropriate-ontologies}). Giving
+Id{} access to Web technologies in this way is essential for evolving it
+into a \emph{fully-featured} self-changeable software environment, which
+could allow as much domain-specifc representation as its user wishes.

### 2020-05-15 convivial-2020/acm-style/ccs20-acm.tex
All day --> ACM style. I hate BibTeX. heck maybe i just hate latex altogether

--- 2020-01-22 convivial-2020/tex/manuscript.tex
@@
-\documentclass[english,submission]{programming}
-\usepackage[
-backend=biber,
-style=alphabetic
-]{biblatex} % Use Biblatex
-\addbibresource{../refs.bib}
+\documentclass[sigconf,screen]{acmart}
+\usepackage{listings}
+
+\citestyle{acmauthoryear}
 
@@
 
+\setcopyright{acmlicensed}
+\acmPrice{15.00}
+\acmDOI{10.1145/3397537.3397549}
+\acmYear{2020}
+\copyrightyear{2020}
+\acmSubmissionID{prog20ccs-id8-p}
+\acmISBN{978-1-4503-7507-8/20/03}
+\acmConference[<Programming'20> Companion]{Companion Proceedings of the 4th International Conference on the Art, Science, and Engineering of Programming}{March 23--26, 2020}{Porto, Portugal}
+\acmBooktitle{Companion Proceedings of the 4th International Conference on the Art, Science, and Engineering of Programming (<Programming'20> Companion), March 23--26, 2020, Porto, Portugal}
+
 \begin{document}
-
-\paperdetails{
-  perspective=art,
-  area={Domain-specific programming, visual programming, self-sustaining systems}
-}
+% The following ONLY works after the \begin{document}!!
+\renewcommand{\textasciitilde}{~}% Don't want Pandoc's BS with the ~ space
 
@@
 
-\input{../manuscript.tex}
+\input{manuscript-acm.tex}
 
-\printbibliography
+\bibliography{../refs.bib}
+
 \end{document}

### 2020-05-15 convivial-2020/acm-style/manuscript-acm.tex
All day --> ACM style. I hate BibTeX. heck maybe i just hate latex altogether

--- 2020-05-14 convivial-2020/manuscript.tex
@@
 
-\maketitle
-
 \newcommand{\joel}[1]{}
@@
 \end{abstract}
+
+\maketitle
 
@@
 
-\hypertarget{the-system}{%
-\section{\texorpdfstring{The Id{}
-system}{The  system}}\label{the-system}}
+\hypertarget{the-id-system}{%
+\section{The Id system}\label{the-id-system}}
 
@@
 
-\hypertarget{as-html-tables}{%
-\subsection{\texorpdfstring{Id{} as HTML
-tables}{ as HTML tables}}\label{as-html-tables}}
+\hypertarget{id-as-html-tables}{%
+\subsection{Id as HTML tables}\label{id-as-html-tables}}
 
@@
   \centering
-  \includegraphics[width=\textwidth]{../orom-html-cropped.png}
+  \includegraphics[width=0.5\textwidth]{orom-html.png}
   \caption{Part of Id{}/HTML: Obj-dicts are rendered as resizable tables, which reference
@@
 
-\hypertarget{as-svg-trees}{%
-\subsection{\texorpdfstring{Id{} as SVG
-trees}{ as SVG trees}}\label{as-svg-trees}}
+\hypertarget{id-as-svg-trees}{%
+\subsection{Id as SVG trees}\label{id-as-svg-trees}}
 
@@
   \centering
-  \includegraphics[width=\textwidth]{../orom-svg-cropped.png}
+  \includegraphics[width=0.5\textwidth]{orom-svg.png}
   \caption{Part of Id{}/SVG: Obj-dicts are moveable and resizable nested boxes,
@@
 between these, I actually stored method mappings with an ad-hoc ``-''
-character (Figure\textasciitilde{}\ref{fig:method-hyphens}).
+character (Figure\textasciitilde{}\ref{fig:orom-html}).
 
@@
 characters, or something else.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.5\textwidth]{../method-hyphens.png}
-  \caption{Id{}/HTML: use of hyphenated \texttt{-lookup} to signify a \emph{method} called
-           \texttt{lookup} instead of a property like \texttt{vtable}. \label{fig:method-hyphens}}
-\end{figure}
 
@@
 
-\hypertarget{the-system-as-a-part-of-the-solution}{%
-\section{\texorpdfstring{The Id{} system as a part of the
-solution}{The  system as a part of the solution}}\label{the-system-as-a-part-of-the-solution}}
+\hypertarget{the-id-system-as-a-part-of-the-solution}{%
+\section{The Id system as a part of the
+solution}\label{the-id-system-as-a-part-of-the-solution}}
 

### 2020-05-20 convivial-2020/acm-style/manuscript-acm.tex
Style corrections; fix random vspace; change title because QUESTION MARK >.>

@@
 \newcommand{\OROM}{Id}
-\title{What does it take to create with domain-appropriate tools?}
+\title{What it takes to create with domain-appropriate tools}
 \subtitle{Reflections on implementing the ``Id{}'' system}
@@
 of effort doubling the performance of a component used 1\% of the time,
-your reward is a system overall improved by a mere 0.5\%. Now, text
+your reward is a system overall improved by a mere 0.47\%. Now, text
 coding is certainly ubiquitous, the 99\% case in programming. A small
@@
 Id, atop a preliminary ``box'' substrate.
-Section\textasciitilde{}\ref{the-system} gives an overview of Id.
-Section\textasciitilde{}\ref{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}
+Section\textasciitilde{}\ref{the-id-system} gives an overview of Id.
+Section\textasciitilde{}\ref{typical-requirements-of-common-software}
 discusses the work involved in obtaining said domain-appropriate
@@
 continues with an interpretation of the labour costs of doing this. In
-Section\textasciitilde{}\ref{the-system-as-a-part-of-the-solution}, we
-return to the Id system and its role in supporting flexible software.
+Section\textasciitilde{}\ref{the-id-system-as-a-part-of-the-solution},
+we return to the Id system and its role in supporting flexible software.
 Section\textasciitilde{}\ref{conclusions-and-future-work} addresses the
-question in the title, and concludes with next steps for the project.
+question in the title\footnote{The original title of this paper was
+  ``What does it take to create with domain-appropriate tools?'', but it
+  had to be changed for technical reasons.}, and concludes with next
+steps for the project.
 
@@
 system that could be self-improved into anything. It describes a
-late-bound\footnote{Fewer commitments; more things determined by runtime
-  conditions}, Smalltalk-style\footnote{OOP with more emphasis on object
-  instances and messaging, as in a distributed system; less emphasis on
-  class hierarchies implementing traditional data structures} objects
-and messaging environment that the authors call ``Id''.
+late-bound,\footnote{Fewer commitments; more things determined by
+  runtime conditions.} Smalltalk-style\footnote{OOP with more emphasis
+  on object instances and messaging, as in a distributed system; less
+  emphasis on class hierarchies implementing traditional data
+  structures.} objects and messaging environment that the authors call
+``Id''.
 
@@
   The higher levels of the vtable-chain mean that different kinds of
-  vtables can be supported\footnote{As well as different kinds of
-    ``kinds of vtables'', and so on.}, each implementing the ``bind''
+  vtables can be supported,\footnote{As well as different kinds of
+    ``kinds of vtables'', and so on.} each implementing the ``bind''
   operation in its own way.
 \end{itemize}
+
+\vspace{-0.1em}
 
@@
 been in the paper in the first place, but I am not necessarily
-representative of those who read it\footnote{Besides that, the
+representative of those who read it.\footnote{Besides that, the
   One-Size-Fits-All{} approach is perhaps unavoidable for static print
-  media.}.
+  media.}
 
@@
 Worse, the reference system does not even have text I/O when it is run,
-let alone some sort of GUI\footnote{That is, the (un)intended user
-  interface for Id ends up being a C debugger!}. Faced with the
-necessity of adding \emph{some} UI, it seemed a waste of effort to end
-up with a system that must be continually polled for its current state
-at a terminal prompt.
+let alone some sort of GUI.\footnote{That is, the (un)intended user
+  interface for Id ends up being a C debugger!} Faced with the necessity
+of adding \emph{some} UI, it seemed a waste of effort to end up with a
+system that must be continually polled for its current state at a
+terminal prompt.
 
@@
 
-\begin{figure}[h]
+\begin{figure}
   \centering
-  \includegraphics[width=0.5\textwidth]{orom-html.png}
+  \includegraphics[width=\linewidth]{orom-html.png}
   \caption{Part of Id{}/HTML: Obj-dicts are rendered as resizable tables, which reference
@@
 mappings, I will refer to them as \emph{obj-dicts}.
-Figure\textasciitilde{}\ref{fig:orom-html} shows the Id{}/HTML
-implementation \cite{orom-html}, in which they take the form of HTML
-tables.
+Figure\textasciitilde{}\ref{fig:orom-html} shows the Id{}/HTML\footnote{This
+  is read ``Id over HTML'', to emphasise the specific substrate on which
+  it sits.} implementation \cite{orom-html}, in which they take the form
+of HTML tables.
 
@@
 
-\begin{figure}[h]
+\begin{figure}
   \centering
-  \includegraphics[width=0.5\textwidth]{orom-svg.png}
+  \includegraphics[width=\linewidth]{orom-svg.png}
   \caption{Part of Id{}/SVG: Obj-dicts are moveable and resizable nested boxes,
@@
 \texttt{\textless{}rect\textgreater{}}s and other elements, reminiscent
-of diSessa's Boxer \cite{boxer}. This was a significant departure from
-the table representation, and even though SVG supports (some) nested
-HTML via \texttt{\textless{}foreignObject\textgreater{}}, I actually
-preferred the possibility of multiple levels of nesting.
+of Boxer \cite{boxer}. This was a significant departure from the table
+representation, and even though SVG supports (some) nested HTML via
+\texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
+the possibility of multiple levels of nesting.
 
@@
 
-\hypertarget{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}{%
-\section{Typical Requirements Of Common Software (and the work we must
-do to meet
-them)}\label{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}}
+\hypertarget{typical-requirements-of-common-software}{%
+\section{Typical Requirements Of Common
+Software}\label{typical-requirements-of-common-software}}
 
@@
 
-\begin{figure}[h]
+\begin{figure*}
   \centering
@@
   \caption{Fully rigid rods transmit changes in one end to the other end. \label{fig:delta-transmission}}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[h]
+\begin{figure*}
   \centering
-  \includegraphics[width=\textwidth]{../rods.png}
+  \includegraphics[width=0.85\textwidth]{rods.png}
   \caption{The yellow rods on the box border are ``half-rigid'': they transmit horizontal or vertical changes, but not both.
@@
 \label{fig:rods}}
-\end{figure}
+\end{figure*}
 
@@
 in a basic facility for translational rigidity. This is only available
-as a tree hierarchy\footnote{This highlights the mismatch between the
-  tree-based DOM and any system that is graph-structured.}, but it is
+as a tree hierarchy,\footnote{This highlights the mismatch between the
+  tree-based DOM and any system that is graph-structured.} but it is
 still useful. However, it conflicts with my early decision to have Point
@@
 simple relations, such as a point following the mouse pointer, are not
-infuriating to express.) Still, it was necessary in the case of certain
+infuriating to express). Still, it was necessary in the case of certain
 elements---especially those which must transcend the tree structure
@@
 In SVG, we express a rectangle by
-\texttt{\textless{}rect x="10" y="10" width="600" height="400"\textgreater{}}.
+
+\begin{lstlisting}
+<rect x="10" y="10" width="600" height="400">
+\end{lstlisting}
+
 The SVG specification \emph{defines} to the user that a rect simply
@@
 \texttt{\textless{}rect\textgreater{}}. The \texttt{x} and \texttt{y}
-are subscribed to the top-left Point\footnote{In the usual case where
+are subscribed to the top-left Point;\footnote{In the usual case where
   the \texttt{\textless{}rect\textgreater{}} is part of a box, the
@@
   \texttt{\textless{}g\textgreater{}}'s \texttt{transform}---but the
-  idea is the same.}; \texttt{width} and \texttt{height} are subscribed
+  idea is the same.} \texttt{width} and \texttt{height} are subscribed
 to Rod lengths.
@@
 \texttt{subscribe(point.position,\ pointer.position)}. The Point has an
-\texttt{is-considering-me?} Observable wrapping
-\texttt{onmouseover}/\texttt{onmouseout}, and the LMB is reified into
-\texttt{left\_mouse\_button\_is\_down} to be explicit. The
-aforementioned subscription is set up whenever
-\texttt{is-considering-me?} and \texttt{left\_mouse\_button\_is\_down}
-become true, and torn down otherwise. I tended to think of this in the
-form ``subscribe to pointer \emph{only when} pointer is-considering-me
-\emph{and} LMB is down'', but I could live with this notation as a
-future polyfill in JavaScript.
+\texttt{is-considering-me?} Observable wrapping \texttt{onmouseover} and
+\texttt{onmouseout}, while a \texttt{LMB\_is\_down} Observable reifies
+the LMB state. The aforementioned subscription is set up whenever
+\texttt{is-considering-me?} and \texttt{LMB\_is\_down} become true, and
+torn down otherwise. I tended to think of this in the form ``subscribe
+to pointer \emph{only when} pointer is-considering-me \emph{and} LMB is
+down'', but I could live with this notation as a future polyfill in
+JavaScript.
 
@@
 \cite{prog21-bbox}. It amounts to translating information from the Web's
-ontology into that of my substrate, as early as possible:
-
-\begin{lstlisting}
+ontology into that of my substrate, as early as possible
+(Listing\textasciitilde{}\ref{lst:drvs}).
+
+\begin{lstlisting}[
+  label={lst:drvs},float,captionpos=b,
+  caption={``Device drivers'' in the code translate the Web's ontology into a more structured form.},
+]
 svg.onmousedown = e =>
-  if (e.button === 0) change(left_mouse_button_is_down, true);
+  if (e.button === 0)
+    change(left_mouse_button_is_down, true);
 
 svg.onmouseup = e =>
-  if (e.button === 0) change(left_mouse_button_is_down, false);
+  if (e.button === 0)
+    change(left_mouse_button_is_down, false);
 
 svg.onmousemove = e => {
+  let client_pos = [e.clientX, e.clientY];
   let r = svg.getBoundingClientRect();
-  let pos = vsub([e.clientX, e.clientY], [r.left, r.top]);
+  let pos = vsub(client_pos, [r.left, r.top]);
   change(pointer.position, pos);
@@
 problem is, this reduction is still described in \emph{natural language}
-for a human implementor, missing the potential of such a
-formulation\footnote{The point here is that a machine-readable
+for a human implementor, missing the potential in such a
+formulation.\footnote{The point here is that a machine-readable
   description could easily remain human-readable, but a natural-language
-  description is not easily machine-readable.}.
+  description is not easily machine-readable.}
 
@@
 author's view that ``which language'' ought to be an
-\emph{implementation detail}, that languages ought to be treatable as
+\emph{implementation detail}, that languages ought to be merely
 \emph{different views} onto the same thing.
@@
 system state: trees or graphs of dictonaries, a.k.a. Maps, a.k.a.
-associative arrays. I am talking about filesystem paths
-\texttt{/path/to/some/file}, Python / Java modules
-\texttt{com.example.pkg.subpkg}, JavaScript objects
-\texttt{window.my\_obj.component}. The common case is an association of
-a textual name to an arbitrary value. I find it useful to see this as a
-mathematical ``function'' defined \emph{extensionally} by listing its
-input/output mappings---this opposed to an \emph{intensional} definition
-such as \(x \mapsto 2x+3\), or a computer program.
+associative arrays. I am talking about the following pattern:
+
+\begin{itemize}
+\tightlist
+\item
+  Filesystem paths \texttt{/path/to/some/file}
+\item
+  Python / Java modules \texttt{com.example.pkg.subpkg}
+\item
+  JavaScript objects \texttt{window.my\_obj.component}
+\end{itemize}
+
+The common case is an association of a textual name to an arbitrary
+value. I find it useful to see this as a mathematical ``function''
+defined \emph{extensionally} by listing its input/output mappings---this
+opposed to an \emph{intensional} definition such as \(x \mapsto 2x+3\),
+or a computer program.
 
@@
 wheels' spokes are silver'' straightforwardly translates to a function
-equation
-
-\begin{lstlisting}
-root (bicycle) (wheels) (spokes) (colour) = root (silver)
-\end{lstlisting}
+equation: \begin{equation*}
+\mathrm{root} (\mathrm{bicycle}) (\mathrm{wheels}) (\mathrm{spokes}) (\mathrm{colour}) = \mathrm{root} (\mathrm{silver})
+\end{equation*}
 
@@
 Normally, this can be as simple as autosave to the filesystem. But the
-Web platform is very wary of this\footnote{As anyone who learns WebGL
+Web platform is very wary of this,\footnote{As anyone who learns WebGL
   can attest to, when they discover they must run a local Web server to
-  provide image files for textures since any filesystem requests will be
-  rejected for security.}, so the solution I turned to was manually
+  provide image files for textures (since any filesystem requests will
+  be rejected for security).} so the solution I turned to was manually
 copying the markup in the browser's inspector
@@
 
-\begin{figure}[h]
+\begin{figure}
   \centering
-  \includegraphics[width=0.8\textwidth]{../html-inspector.png}
+  \includegraphics[width=\linewidth]{../html-inspector.png}
   \caption{To save the state of the system, the SVG group corresponding to the \texttt{root} box is copied and pasted inside the HTML file's \texttt{\textless{}svg\textgreater{}} element.\label{fig:html-inspector}}
@@
 on some distant and busy\footnote{Not to mention \emph{unelected?}
-  ``Take Back Control!'' (jokes aside, in this context it makes a lot of
+  ``Take Back Control!'' (Jokes aside, in this context it makes a lot of
   sense.)} language designer. It seems plausible that this is a step on
@@
 The Id{} paper was part of the STEPS project of Alan Kay's
-VPRI\footnote{Viewpoints Research Institute}. This project argued that
+VPRI.\footnote{Viewpoints Research Institute} This project argued that
 the immense level of ``accidental complexity'' present in software
@@
 T-shirt. The obvious \emph{reductio ad absurdum} is where we encapsulate
-these equations under a single symbol\footnote{In fact, this is almost
+these equations under a single symbol,\footnote{In fact, this is almost
   achieved by the formalism of Geometric Calculus, which reduces them to
-  only one equation.}, \(\mathcal{M}\). \(\mathcal{M}\) is defined as
+  only one equation.} \(\mathcal{M}\). \(\mathcal{M}\) is defined as
 ``Maxwell's Equations are true''. Ta-da---this fits on a coin, but good
@@
 those specific to the Web, but also issues that occur far more widely
-(e.g.~monopoly ontologies.) I showed how, in the short-term, we can
+(e.g.~monopoly ontologies). I showed how, in the short-term, we can
 overcome these with solutions intended to be domain-appropriate to their

### 2020-05-28 convivial-2020/acm-style/manuscript-acm.tex
IT JUST WON'T STAY DEAD ... KEEPS COMING BACK!! pls accept now thx

@@
 \newcommand{\OROM}{Id}
-\title{What it takes to create with domain-appropriate tools}
-\subtitle{Reflections on implementing the ``Id{}'' system}
-\author[ ]{Joel Jakubovic}
-\affiliation[ ]{University of Kent, Canterbury}
+\title{What It Takes to Create with Domain-Appropriate Tools}
+\subtitle{Reflections on Implementing the ``Id{}'' System}
+\author{Joel Jakubovic}
+\affiliation{University of Kent, Canterbury, UK}
+\email{jdj9@kent.ac.uk}
 
@@
 Section\textasciitilde{}\ref{conclusions-and-future-work} addresses the
-question in the title\footnote{The original title of this paper was
-  ``What does it take to create with domain-appropriate tools?'', but it
-  had to be changed for technical reasons.}, and concludes with next
-steps for the project.
+question in the title,\footnote{The original title was ``What Does It
+  Take to Create with Domain-Appropriate Tools?'', but it has been
+  changed for technical reasons.} and concludes with next steps for the
+project.
 
@@
 \hypertarget{id-as-html-tables}{%
-\subsection{Id as HTML tables}\label{id-as-html-tables}}
+\subsection{Id as HTML Tables}\label{id-as-html-tables}}
 
@@
 \hypertarget{the-radical-concept-of-arrows-that-stay-on-the-shapes}{%
-\subsubsection{The Radical Concept of Arrows That Stay On The
+\subsubsection{The Radical Concept of Arrows that Stay on the
 Shapes}\label{the-radical-concept-of-arrows-that-stay-on-the-shapes}}
@@
 \hypertarget{id-as-svg-trees}{%
-\subsection{Id as SVG trees}\label{id-as-svg-trees}}
+\subsection{Id as SVG Trees}\label{id-as-svg-trees}}
 
@@
 \hypertarget{typical-requirements-of-common-software}{%
-\section{Typical Requirements Of Common
+\section{Typical Requirements of Common
 Software}\label{typical-requirements-of-common-software}}
@@
 \hypertarget{basic-assumptions-about-physical-objects}{%
-\subsection{Basic Assumptions About Physical
+\subsection{Basic Assumptions about Physical
 Objects}\label{basic-assumptions-about-physical-objects}}
@@
 \hypertarget{maintaining-relationships-over-time}{%
-\subsection{Maintaining Relationships Over
+\subsection{Maintaining Relationships over
 Time}\label{maintaining-relationships-over-time}}
@@
 \hypertarget{nut-cracking-with-sledgehammers}{%
-\subsection{Nut-Cracking With
+\subsection{Nut-Cracking with
 Sledgehammers}\label{nut-cracking-with-sledgehammers}}
@@
 \hypertarget{wrapping-and-user-data}{%
-\subsection{Wrapping and ``user data''}\label{wrapping-and-user-data}}
+\subsection{Wrapping and ``User Data''}\label{wrapping-and-user-data}}
 
@@
 \hypertarget{context-appropriate-ontologies}{%
-\subsection{Context-appropriate
-ontologies}\label{context-appropriate-ontologies}}
+\subsection{Context-Appropriate
+Ontologies}\label{context-appropriate-ontologies}}
 
@@
 \hypertarget{the-id-system-as-a-part-of-the-solution}{%
-\section{The Id system as a part of the
-solution}\label{the-id-system-as-a-part-of-the-solution}}
+\section{The Id System as a Part of the
+Solution}\label{the-id-system-as-a-part-of-the-solution}}
 
@@
 \hypertarget{minimal-descriptions}{%
-\subsection{Minimal descriptions}\label{minimal-descriptions}}
+\subsection{Minimal Descriptions}\label{minimal-descriptions}}
 
@@
 \hypertarget{self-implementation}{%
-\subsection{Self-implementation}\label{self-implementation}}
+\subsection{Self-Implementation}\label{self-implementation}}
 
@@
 \hypertarget{conclusions-and-future-work}{%
-\section{Conclusions and future
-work}\label{conclusions-and-future-work}}
+\section{Conclusions and Future
+Work}\label{conclusions-and-future-work}}
 
 \hypertarget{what-did-it-take-and-why}{%
-\subsection{What did it take, and why?}\label{what-did-it-take-and-why}}
+\subsection{What Did It Take, and Why?}\label{what-did-it-take-and-why}}
 
@@
 \hypertarget{what-next}{%
-\subsection{What next?}\label{what-next}}
+\subsection{What Next?}\label{what-next}}
 

### 2020-10-08 ppig-2020/ppig20-doctoral-consortium.tex
ppig20 dump

\begin{abstract}
It is comparatively easy to find new \emph{languages} and derivative work within academic publishing, but somewhat harder for more general programming "systems" or "environments" which encompass more than that. This is a shame, since some of these have a dedicated following and have influenced programming and computing at large. We concentrate on some examples of such that, for various reasons, do not have much material written about them. We suggest some reasons to expect this, and draw attention to some characteristics (both "cognitive" and "technical") marking them worthy of further study. We conclude with a sketch of further steps to make the most of these software artefacts from a research perspective.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Programming language theory, implementation and evaluation is an
expansive and well-established field. Even entire applications,
``environments'' or ``systems'', developed \emph{for the purposes of
research}, contribute to progress by being discussed and evaluated
within a certain academic scope (for example, HCI). However, there exist
innovative, influential or otherwise noteworthy systems that
nevertheless fall outside of these realms. As we expand on later, this
is in part due to originating outside of the academic environment, and
it is also pushed against by the nature of the publishing medium. We
will lead in with three such systems to illustrate our later points.

\hypertarget{smalltalk}{%
\subsection{Smalltalk}\label{smalltalk}}

Smalltalk could be described as a self-sufficient desktop environment
programmed in a language of the same name. It was influential on the
rise of object-orientation and graphical user interfaces, and was itself
intended to be widely embraced. However, it did not manage to achieve
this goal, and lives on in its own niche on a par with other programming
language ecosystems.

Whenever a language or IDE feature spreads across the mainstream,
Smalltalkers have a reputation for pointing out that Smalltalk was able
to do the same thing back in the '80s. This is a plausible claim, owing
to the deliberate high-level, flexible and totalising architecture of
the system. Noting its following by a community of devotees, this is
evidence of something special and valuable to learn from and build upon.
However, despite having a literature presence in virtual-machine
optimisation and related \emph{implementation} technologies, insight
into Smalltalk's \emph{design} seems to be mostly scattered around
historical magazine articles \cite{BYTEmag}, web pages and blog posts.

Without a well-structured design discussion, it is only clear that there
is ``something about Smalltalk'' that is worth improving on. In other
words, it is difficult to distinguish which specific aspects of
Smalltalk are essential to its value and which are incidental. We can
identify characteristics such as \emph{meta-circularity} and
\emph{self-sufficiency}, by which the software within lives in a world
where ``everything is Smalltalk''. Another identifying feature is that
instead of transient ``applications'' and manual saving to ``files'',
the entire system state is persisted as a continuously evolving
``image''. Smalltalk also embeds its programming language within a
larger prescribed context of graphical interface and device utilities,
encompassing the same roles (and scope) of an entire Operating System.
Such prescription of a graphical interface is characteristic of what we
are referring to as ``systems'', but this last ``encompass the world''
aspect is more unusual.

\hypertarget{hypercard}{%
\subsection{HyperCard}\label{hypercard}}

HyperCard was a platform for exploring and creating interlinked sets of
multimedia pages, often regarded as a precursor to the Web. Explicitly
designed with the goal of end-user empowerment, with a slogan of
``programming for the rest of us'', HyperCard was a popular platform
both to share and to author among teachers, businesspeople, and other
non-programmers. This was no doubt in part due to its (initial) default
inclusion in Macs. However, the existence of other default apps (such as
Terminal) \emph{not} enthusiastically adopted by end-users means that
HyperCard itself still stands on its own merit.

HyperCard applications were organised into ``stacks'', with typical UI
elements (text, pictures, buttons) added and edited by direct
manipulation. Every such element could have a script attached to it in
HyperTalk, a language designed to resemble English. HyperCard also
distinguished itself in having separate ``beginner'' / ``advanced''
layers, allowing novices to make their way through straightforwardly
\emph{using} and maybe editing stacks superficially, before digging into
more advanced authoring and scripting.

Unfortunately, despite its popularity and originality, HyperCard fell
out of use due to corporate and marketing decisions and subsequent
declining technical support, eventually falling out of compatibility
with newer versions of the Mac OS. As HyperCard was primarily a
\emph{commercial} product, not much has been written about its design,
but its popularity suggests that it is worth studying and possibly
adapting for future software systems. Sample design characteristics for
HyperCard include its layered design allowing progression from
\emph{using} to \emph{developing} card stacks, and the fact that such
developing is not considered something \emph{far away} but instead
occurs in the same user interface.

\hypertarget{flash}{%
\subsection{Flash}\label{flash}}

Adobe Flash was a widely popular browser plugin providing ``rich''
content (video, audio, and games) for the primitive early Web. Such
applications were created using Flash Builder: an advanced graphical
workspace for various forms of vector graphics and animation, scripted
with an event-driven language called ActionScript.

Flash gave rise to an explosion of animations, games and websites across
the early Web, often by individuals and within online communities.
Because it was used so widely for websites and games, Flash can be
considered a success case, to some extent, of ``end-user'' software
development. Unfortunately, Flash was thought to be too insecure and
poorly-optimised for the emerging mobile platforms of the 2010s. Apple's
decision not to support Flash on the iPhone ultimately spelled its
decline and planned official demise at the end of 2020.

Following from our previous two examples, what are the interesting
design characteristics of Flash? Its success may be attributed to the
fact that it merely \emph{filled a gap} and achieved rapid ubiquity as a
result, even defining a new \emph{medium} of interactive Web content.
This is certainly true, but suggests this role could have been filled by
any other provider of rich content support. Flash Builder, though, does
seem to be an effectively designed tool for its audience. It is a
graphical editor centering around a WYSIWYG ``stage'' metaphor,
surrounded by many specialised sub-interfaces (for example, managing the
broad and narrow details of keyframe animation), and with an attached
scripting sub-editor. Are there more specific design principles hiding
here than merely ``use the right interface for the job''?

\hypertarget{what-to-do}{%
\section{What to do?}\label{what-to-do}}

\hypertarget{rational-reconstruction}{%
\subsection{Rational Reconstruction}\label{rational-reconstruction}}

The main point of focusing on such systems is that they exist outside of
the academic literature, having been commercial or educational ventures
rather than research artefacts. This means that theoretical analysis or
design rationale, if it exists at all, may have never been published or
be scattered around documents for internal use. In some cases, there may
not be any such written material or what existed has been lost.

This means that it is currently hard to do rigorous follow-up work on
these artefacts. If we wish to build on their successes and learn from
their failures, in a scholarly way, there needs to be more than the
artefact itself and documentation. They need to be made ``legible'' to
academic literature by \emph{reconstructing} what their design and
theoretical analysis \emph{would} have been in such a counterfactual
world.

\hypertarget{cognitive-and-technical-dimensions}{%
\subsection{Cognitive and Technical
Dimensions}\label{cognitive-and-technical-dimensions}}

One component of such a ``rational reconstruction'' could be an analysis
in terms of a common vocabulary of named characteristics. For example,
the Cognitive Dimensions of Notation framework \cite{cogdims} suggests a
number of ``dimensions'' along which a user interface or ``notation''
may be measured, along with linkages and tradeoffs between different
dimensions. This is certainly useful, but its emphasis on the
\emph{cognitive} aspect of \emph{interfaces} might leave uncharted other
aspects -- what we suggest to be \emph{technical} dimensions -- of the
``rest'' of a software system.

For example, a programming language in which ``code'' is expressed in
the same form as ``data'' is said to exhibit ``homo-iconicity''. We
could extend this to systems-in-general, perhaps, as a measure of how
much \emph{automating} a task resembles manually performing it.
Similarly, ``self-sufficiency'' measures the extent to which a system
can be improved and evolved from within, without relying on external
tools. Smalltalk was designed to be quite high on this measure (intended
to ``obsolete itself''), while HyperCard and Flash (along with most
other software) facilitate creation of \emph{separate} artefacts
(HyperCard stacks, Flash applications) and can only be changed via
in-application preferences or by editing their source code.

Our desire for ``technical'' dimensions comes from a feeling that there
are also properties of ``the system itself'' rather than merely its
\emph{interface}. If \emph{cognitive} dimensions are about how a system
is \emph{received} or \emph{experienced} by users, then \emph{technical}
dimensions are about how the system was \emph{intended} or
\emph{designed}. However, this does rest on an assumption that
separating the ``interface'' from the ``system itself'' is a meaningful
or useful thing in the first place, and we invite further discussion of
this.

\hypertarget{the-medium-is-the-message}{%
\subsection{The Medium is the Message}\label{the-medium-is-the-message}}

Another important thing to note is that academic publishing is by and
large a paper (virtual or physical) medium; as such, the ease with which
one can publish varies according to the medium of the subject matter.
Programming languages, being strings of characters equipped with syntax
and semantics, lend themselves very well to describing in a paper.
However, our systems of interest is that tend to be graphical,
interactive \emph{systems} rather than \emph{languages} alone. They
define much more of a given computing environment than languages; they
additionally state how they are graphically presented and manipulated.
So, in order to properly present or discuss these systems in full, one
needs to at least provide pictures and ideally provide a running
version.

While videos or runnable artefacts can be part of academic submissions,
they are still seen as ``supplementary material'' to accompany a written
paper. It is plausible that this makes it less attractive to present
general ``systems'' in academic publishing, and the non-academic nature
of our examples is easier to understand in light of this. This is
recognised in \cite{eval-prog-sys}, which begins a conversation about
how format, submission and peer-review ought to look for such
interactive systems and work on their \emph{design}. Such a model would
support a ``gallery of interactions'' with (possibly simplified)
re-creations of systems as one possible corrective.

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

In summary, there exist interesting holistic ``systems'' which seem to
have been ignored in published research. To be able to do follow-up work
on them, we need a deeper and more rigorous understanding. But their
typical commercial or educational nature means there was no initial body
of literature to seed such a process. Additionally, the (virtual) paper
medium is not well suited to ``systems'' in the first place. We propose
to re-create the design analysis in light of innovations like the
Cognitive Dimensions framework, and taking advantage of more appropriate
media like video or interactive essays. A Smalltalker might say that
``everything was already invented in the 1970s''; we agree that there
was little explicit building upon the visionary work of the past. To
remedy this, we need to take a fresh, academically rigorous look at such
old work on programming systems.

### 2021-06-13 plop-2021/plop21-master.tex
archive and tidyup work so far

\documentclass{article}
\usepackage{epsfig}
\usepackage{booktabs}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}

% The titlebox defines how much vertical space is given for
% the authors' list. If you need extra space to show all the
% authors, uncomment the line below and increase the value. Please
% do not make the titlebox smaller than the original size of 5cm.
%\setlength\titlebox{5cm}

\title{Technical Dimensions of Programming Systems}

\author{Joel Jakubovic}
\affiliation{%
  \institution{University of Kent}
  \city{Canterbury}
  \country{UK}}
\email{jdj9@kent.ac.uk}
\author{Jonathan Edwards}
\affiliation{}
\email{jonathanmedwards@gmail.com}
\author{Tomas Petricek}
\affiliation{%
  \institution{University of Kent}
  \city{Canterbury}
  \country{UK}}
\email{T.Petricek@kent.ac.uk}

\authorsaddresses{}

\date{}

\begin{document}
\maketitle
\thispagestyle{empty}

\newcommand{\joel}[1]{}
\newcommand{\note}[1]{}
\newcommand{\tp}[1]{}
\newcommand{\hypertarget}[1]{}

\input{plop21-manuscript.tex}

\bibliography{plop21}
\bibliographystyle{apacite}
\end{document}

### 2021-06-15 plop-2021/plop21-master.tex
tidy for initial submission

@@
-\documentclass{article}
-\usepackage{epsfig}
-\usepackage{booktabs}
-\usepackage{ucs}
-\usepackage[utf8x]{inputenc}
+\documentclass[acmlarge, review=true]{acmart}
 
-% The titlebox defines how much vertical space is given for
-% the authors' list. If you need extra space to show all the
-% authors, uncomment the line below and increase the value. Please
-% do not make the titlebox smaller than the original size of 5cm.
-%\setlength\titlebox{5cm}
+\acmConference{PLoP}{2021}{online}
+
+\markboth{}{}
 
@@
 \author{Joel Jakubovic}
-\affiliation{%
-  \institution{University of Kent}
-  \city{Canterbury}
-  \country{UK}}
+\affiliation{
+\institution{University of Kent} \city{Canterbury} \country{UK}
+}
 \email{jdj9@kent.ac.uk}
@@
 \author{Tomas Petricek}
-\affiliation{%
-  \institution{University of Kent}
-  \city{Canterbury}
-  \country{UK}}
+\affiliation{
+\institution{University of Kent} \city{Canterbury} \country{UK}
+}
 \email{T.Petricek@kent.ac.uk}
 
-\authorsaddresses{}
-
-\date{}
+\begin{abstract}
+Programming systems often go beyond programming languages. Programming
+is often done in the context of a stateful environment, through a
+graphical user interface, by interacting with the system rather than
+just by writing code. Much research effort focuses on building
+programming systems that are easier to use, accessible to non-experts,
+moldable and/or powerful, but such efforts are often disconnected. They
+are informal, guided by the personal vision of the authors and thus are
+only evaluable and comparable on the basis of individual experience
+using them. In other words, they fail to form a coherent body of
+research. It isn't clear how to build on what has been done. Much has
+been said and done that allows comparison of \emph{programming
+languages}, yet no similar theories exist for \emph{programming
+systems}. We believe that programming systems deserve a theory too. We
+examine some influential past programming systems and review their
+stated design principles, technical capabilities and styles of user
+interaction. We propose a framework of \emph{technical dimensions},
+which captures the underlying system characteristics and provide a means
+for conceptualising and comparing programming systems. We thus break the
+holistic view of programming systems down into characteristics that can
+be compared or advanced independently. It makes it possible to talk
+about programming systems in a way that can be shared and agreed or
+constructively disagreed upon, rather than relying just on individual
+experience of using them. The aim is to bring programming systems into
+the focus of programming research by providing the foundations that
+allow such research to proceed in a more systematic way. In other words,
+we want to allow the designers of future programming systems to
+(finally) stand on the shoulders of the giants.
+\end{abstract}
 
 \begin{document}
+
 \maketitle
-\thispagestyle{empty}
+%\thispagestyle{empty}
 
@@
 \newcommand{\tp}[1]{}
-\newcommand{\hypertarget}[1]{}
+\providecommand{\tightlist}{}% Don't want Pandoc's tight lists
+%\newcommand{\hypertarget}[1]{}
 
@@
 
+\appendix
+
+\input{plop21-appendix.tex}
+
+\citestyle{acmauthoryear}
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{plop21}
-\bibliographystyle{apacite}
 \end{document}

### 2021-08-25 plop-2021/plop21-master.tex
First revision

@@
 \acmConference{PLoP}{2021}{online}
+\citestyle{acmauthoryear}
 
@@
 \begin{abstract}
-Programming systems often go beyond programming languages. Programming
-is often done in the context of a stateful environment, through a
-graphical user interface, by interacting with the system rather than
-just by writing code. Much research effort focuses on building
-programming systems that are easier to use, accessible to non-experts,
-moldable and/or powerful, but such efforts are often disconnected. They
-are informal, guided by the personal vision of the authors and thus are
-only evaluable and comparable on the basis of individual experience
-using them. In other words, they fail to form a coherent body of
-research. It isn't clear how to build on what has been done. Much has
-been said and done that allows comparison of \emph{programming
-languages}, yet no similar theories exist for \emph{programming
-systems}. We believe that programming systems deserve a theory too. We
-examine some influential past programming systems and review their
-stated design principles, technical capabilities and styles of user
-interaction. We propose a framework of \emph{technical dimensions},
-which captures the underlying system characteristics and provide a means
-for conceptualising and comparing programming systems. We thus break the
-holistic view of programming systems down into characteristics that can
-be compared or advanced independently. It makes it possible to talk
-about programming systems in a way that can be shared and agreed or
-constructively disagreed upon, rather than relying just on individual
-experience of using them. The aim is to bring programming systems into
-the focus of programming research by providing the foundations that
-allow such research to proceed in a more systematic way. In other words,
-we want to allow the designers of future programming systems to
-(finally) stand on the shoulders of the giants.
+Many programming systems go beyond programming languages.
+Programming is usually done in the context of a stateful environment, beyond just writing code, by interacting with a system through a graphical user interface.
+Much research effort focuses on building programming systems that are easier to use, accessible to non-experts, moldable and/or powerful, but such efforts are often disconnected.
+They are informal, guided by the personal vision of the authors and thus are only evaluable and comparable on the basis of individual experience using them.
+In other words, they fail to form a coherent body of research.
+It isn't clear how to build on what has been done.
+Much has been said and done that allows comparison of \emph{programming languages}, yet no similar theories exist for \emph{programming systems}.
+We believe that programming systems deserve a theory too.
+We examine some influential past programming systems and review their stated design principles, technical capabilities and styles of user interaction.
+We propose a framework of \emph{technical dimensions}, capturing the underlying system characteristics and providing a means for conceptualising and comparing programming systems.
+We thus break the holistic view of programming systems down into characteristics that can be compared or advanced independently.
+These make it possible to talk about programming systems in a way that can be shared and agreed or constructively disagreed upon, rather than relying just on individual experience of using them.
+Our aim is to bring programming systems into the focus of programming research by providing the foundations that allow such research to proceed in a more systematic way.
+In other words, we wish the designers of future programming systems to finally stand on the shoulders of giants.
 \end{abstract}
@@
 
-\citestyle{acmauthoryear}
 \bibliographystyle{ACM-Reference-Format}

### 2021-09-14 plop-2021/plop21-master.tex
Pre writers'w workshop version

@@
-\documentclass[acmlarge, review=true]{acmart}
+\documentclass[acmlarge%, review=true]
+]{acmart}
 
@@
 We believe that programming systems deserve a theory too.
-We examine some influential past programming systems and review their stated design principles, technical capabilities and styles of user interaction.
+We examine some influential past programming systems and review their stated design principles, technical capabilities, and styles of user interaction.
 We propose a framework of \emph{technical dimensions}, capturing the underlying system characteristics and providing a means for conceptualising and comparing programming systems.

### 2022-01-21 prog-2022/prog22-master.tex
prog22 etc
fix typos / edit tomas, rework abstract a bit

--- 2021-09-14 plop-2021/plop21-master.tex
@@
-\documentclass[acmlarge%, review=true]
+\documentclass[acmlarge, review=true
 ]{acmart}
 
-\acmConference{PLoP}{2021}{online}
+\acmConference{}{2021}{}
 \citestyle{acmauthoryear}
@@
 They are informal, guided by the personal vision of the authors and thus are only evaluable and comparable on the basis of individual experience using them.
-In other words, they fail to form a coherent body of research.
-It isn't clear how to build on what has been done.
-Much has been said and done that allows comparison of \emph{programming languages}, yet no similar theories exist for \emph{programming systems}.
-We believe that programming systems deserve a theory too.
+This fails to form a coherent body of research, since it is unclear how to build on past work.
+In the research world, much has been said and done that allows comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems}; we believe that programming systems deserve a theory too.
 We examine some influential past programming systems and review their stated design principles, technical capabilities, and styles of user interaction.
-We propose a framework of \emph{technical dimensions}, capturing the underlying system characteristics and providing a means for conceptualising and comparing programming systems.
-We thus break the holistic view of programming systems down into characteristics that can be compared or advanced independently.
-These make it possible to talk about programming systems in a way that can be shared and agreed or constructively disagreed upon, rather than relying just on individual experience of using them.
-Our aim is to bring programming systems into the focus of programming research by providing the foundations that allow such research to proceed in a more systematic way.
-In other words, we wish the designers of future programming systems to finally stand on the shoulders of giants.
+We propose a framework of \emph{technical dimensions} which capture the underlying system characteristics and provide a means for conceptualizing and comparing programming systems.
+Since these characteristics may be compared or advanced independently, it should be easier to talk about programming systems in a way that can be shared and constructively debated rather than relying solely on personal impressions.
+By providing foundations for more systematic research in this area, we can help the designers of future programming systems to stand, at last, on the shoulders of giants.
 \end{abstract}
@@
 
-\input{plop21-manuscript.tex}
+\input{prog22-manuscript.tex}
 
-\appendix
-
-\input{plop21-appendix.tex}
+% \appendix
+% \input{prog22-appendix.tex}
 
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{plop21}
+\bibliography{prog22}
 \end{document}

### 2022-01-26 prog-2022/prog22-master.tex
Rework Conceptualization -> Architecture and merge in FoC
Paper now 40pg long with <programming> class.

@@
-\documentclass[acmlarge, review=true
-]{acmart}
+\documentclass[english,submission]{programming}
 
-\acmConference{}{2021}{}
-\citestyle{acmauthoryear}
+%\citestyle{acmauthoryear}
+%\markboth{}{}
 
-\markboth{}{}
+% Class scrartcl Warning: seems someone has broken package `auxhook'.
+% Usually this happens, if `auxhook' is loaded or used implicitly or explicitly
+% by patching \document or via etoolbox command \AtEndPreamble. Trying an
+% emergency workaround. You can avoid this warning adding:
+\usepackage{auxhook}
+% before \begin{document} on input line 6.
+\begin{document}
+
+\paperdetails{perspective=art, area=tbd}
 
@@
 \affiliation{
-\institution{University of Kent} \city{Canterbury} \country{UK}
+University of Kent, Canterbury, UK
+\email{jdj9@kent.ac.uk}
 }
-\email{jdj9@kent.ac.uk}
 \author{Jonathan Edwards}
-\affiliation{}
-\email{jonathanmedwards@gmail.com}
+\affiliation{\email{jonathanmedwards@gmail.com}}
 \author{Tomas Petricek}
 \affiliation{
-\institution{University of Kent} \city{Canterbury} \country{UK}
+University of Kent, Canterbury, UK
+\email{T.Petricek@kent.ac.uk}
 }
-\email{T.Petricek@kent.ac.uk}
+
+\keywords{paper, lorem ipsum}
+
+% Please go to https://dl.acm.org/ccs/ccs.cfm and generate your Classification
+% System [view CCS TeX Code] stanz and copy _all of it_ to this place.
+%% From HERE
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10002944.10011122.10003459</concept_id>
+<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10010405.10010476.10010477</concept_id>
+<concept_desc>Applied computing~Publishing</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}
+
+\maketitle
 
@@
 
-\begin{document}
-
-\maketitle
 %\thispagestyle{empty}
@@
 \newcommand{\tp}[1]{}
+\newcommand{\mybox}[1]{\noindent\fbox{\parbox{\textwidth}{#1}}}
 \providecommand{\tightlist}{}% Don't want Pandoc's tight lists
@@
 
-\bibliographystyle{ACM-Reference-Format}
 \bibliography{prog22}

### 2022-01-28 prog-2022/summary-table.tex
Insert table (too big) and copy concise cluster summaries from it

\begin{table}
%\centering
\newcommand{\summ}[1]{\multicolumn{2}{p{0.7\linewidth}|}{#1} \\ \hline}

\setlength\extrarowheight{2pt}
\begin{adjustwidth}{-2cm}{-2cm}
\begin{tabular}{| >{\raggedleft\arraybackslash}p{0.25\linewidth} | p {0.35\linewidth} | >{\raggedright\arraybackslash}p{0.35\linewidth} |}
\hline
Dimension (CLUSTER) & Summary & Range of examples \\ \hline
\hline INTERACTION &
\summ{How do users execute their ideas, evaluate the result, and generate new ideas in response?}
\def\arraystretch{1}
Feedback Loops &
How wide are the various gulfs of \emph{execution} and \emph{evaluation} and how do they relate to each other? &
Immediate Feedback (short) vs batch mode (long) evaluation \\
Modes of Interaction &
TODO &
TODO \\
Abstraction Construction &
How do we go from abstractions to concrete examples and vice versa? &
TODO \\
\hline NOTATION &
\summ{How are the different textual / visual programming notations related?}
Notational Structure &
What notations are used to program the system and how are they related? &
Notations overlap and need sync vs. complement each other \\
Surface / Internal Notations &
What is the connection between what a user sees and what a computer program sees? &
Sequence Editing vs. Rendering, Structure Editing vs. Recovery \\
Primary / Secondary Notations &
Is one notation more important than others? &
Build scripts are secondary vs. equal importance of visual editor and code in Flash \\
Expression Geography &
Do similar expressions encode similar programs? &
Concise yet error-prone vs. explicit yet verbose \\
Uniformity of Notations &
Does the notation use a small or a large number of basic concepts? &
Lisp S-expressions vs. English-like textual notations \\
\hline ARCHITECTURE &
\summ{What is the system's attitude to its own organization and that of the outside world?}
Conceptual Outlook &
Does the system present as \emph{designed} or \emph{improvised}? &
Integrity (Everything is a X) vs. openness (compatible mixtures) \\
Composability &
What are the primitives? How can they be combined to achieve novel behaviors? &
TODO \\
Convenience &
Which wheels do users not need to reinvent? &
TODO \\
Commonality &
How much is \emph{common structure} explicitly marked as such? &
Common structure is redundantly flattened vs. factored out\\
\hline CUSTOMIZABILITY &
\summ{Once a program exists in the system, how can it be extended and modified?}
Staging of Customization &
Must we customize \emph{running} programs differently to \emph{inert} ones? Do these changes last beyond termination? &
TODO \\
Addressability &
How many actual and potential \emph{system states} can customizations refer to? &
TODO \\
Additive Authorship &
How far can the system's behavior be changed by \emph{adding} expressions? &
TODO \\
Self-Sustainability &
How far can the system’s behavior be changed from within? &
TODO \\
\hline AUTOMATION &
\summ{To what extent, and in what ways, does the system remove the need to spell out implementation in minute detail?}
Degrees of Automation &
What part of program logic does not need to be explicitly specified? &
Automatic memory management is a low-end example; Prolog a higher-end one. \\
\hline ERRORS &
\summ{What does the system consider to be an error, and how does it approach their prevention and handling?}
Error Detection &
What errors can be detected in which feedback loops and what is the detection mechanism? &
In live coding, errors are detected by humans; in static typing, this is partly automated. \\
Error Response &
How does the system respond when an error is detected? &
Does it stop, recover automatically, ignore the error or ask the user how to continue? \\
\hline ADOPTABILITY &
\summ{How does the system facilitate or obstruct adoption by both individuals and communities?}
Learnability &
TODO &
TODO \\
Sociability &
TODO &
TODO \\
\hline
\end{tabular}
\end{adjustwidth}
\caption{\label{summary-table} Quick reference sheet for our set of Technical Dimensions.}
\end{table}

### 2022-01-28 prog-2022/summary-table.tex
Split table in 2 and respect style reqs; looks worse!

@@
+\newcommand{\summ}[1]{\multicolumn{2}{p{0.8\linewidth}}{#1} \\ \hline}
+\setlength\extrarowheight{2pt}
 \begin{table}
 %\centering
-\newcommand{\summ}[1]{\multicolumn{2}{p{0.7\linewidth}|}{#1} \\ \hline}
-
-\setlength\extrarowheight{2pt}
-\begin{adjustwidth}{-2cm}{-2cm}
-\begin{tabular}{| >{\raggedleft\arraybackslash}p{0.25\linewidth} | p {0.35\linewidth} | >{\raggedright\arraybackslash}p{0.35\linewidth} |}
+\begin{tabular}{ >{\raggedleft\arraybackslash}p{0.2\linewidth}  p {0.4\linewidth}  >{\raggedright\arraybackslash}p{0.4\linewidth} }
 \hline
-Dimension (CLUSTER) & Summary & Range of examples \\ \hline
+Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
 \hline INTERACTION &
 \summ{How do users execute their ideas, evaluate the result, and generate new ideas in response?}
-\def\arraystretch{1}
 Feedback Loops &
-How wide are the various gulfs of \emph{execution} and \emph{evaluation} and how do they relate to each other? &
-Immediate Feedback (short) vs batch mode (long) evaluation \\
+How wide are the various gulfs of \emph{execution} and \emph{evaluation} and how are they related? &
+Immediate Feedback (short) vs. batch mode (long) gulf of evaluation \\
 Modes of Interaction &
-TODO &
-TODO \\
+Which sets of feedback loops only occur together? &
+Setup vs. editing vs. debugging \\
 Abstraction Construction &
 How do we go from abstractions to concrete examples and vice versa? &
-TODO \\
+Programming by Example vs. first principles \\
 \hline NOTATION &
@@
 Is one notation more important than others? &
-Build scripts are secondary vs. equal importance of visual editor and code in Flash \\
+Secondary build scripts vs. visual editor and code on equal footing in Flash \\
 Expression Geography &
@@
 What are the primitives? How can they be combined to achieve novel behaviors? &
-TODO \\
+Sequence, selection, repetition, function abstraction, recursion, logical connectives \\
 Convenience &
 Which wheels do users not need to reinvent? &
-TODO \\
+Small vs. expansive standard libraries \\
 Commonality &
@@
 Common structure is redundantly flattened vs. factored out\\
+\hline
+\end{tabular}
+\caption{\label{summary-table1} Quick reference sheet for our set of Technical Dimensions.}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ >{\raggedleft\arraybackslash}p{0.2\linewidth}  p {0.4\linewidth}  >{\raggedright\arraybackslash}p{0.4\linewidth} }
+\hline
+Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
 \hline CUSTOMIZABILITY &
@@
 TODO \\
-Addressability &
-How many actual and potential \emph{system states} can customizations refer to? &
+Externalizability &
+Which portions of the system's state can be referenced and transferred to/from it? &
 TODO \\
-Additive Authorship &
+Open Authoring &
 How far can the system's behavior be changed by \emph{adding} expressions? &
@@
 \hline AUTOMATION &
-\summ{To what extent, and in what ways, does the system remove the need to spell out implementation in minute detail?}
+\summ{How far does the system remove the need to spell out implementation in minute detail?}
 Degrees of Automation &
 What part of program logic does not need to be explicitly specified? &
-Automatic memory management is a low-end example; Prolog a higher-end one. \\
+Garbage collection (low-tech) vs. Prolog engine (hi-tech) \\
 \hline ERRORS &
-\summ{What does the system consider to be an error, and how does it approach their prevention and handling?}
+\summ{What does the system consider to be an error; how are they prevented and handled?}
 Error Detection &
-What errors can be detected in which feedback loops and what is the detection mechanism? &
-In live coding, errors are detected by humans; in static typing, this is partly automated. \\
+What errors can be detected in which feedback loops, and how? &
+Human inspection in live coding vs. partial automation in static typing \\
 Error Response &
@@
 Learnability &
-TODO &
-TODO \\
+What is the attitude towards the \emph{learning curve} and what is the target audience? &
+HyperCard for the general public vs. FORTRAN for scientists \\
 Sociability &
-TODO &
+What are the social and economic factors that make the system the way it is? &
 TODO \\
@@
 \end{tabular}
-\end{adjustwidth}
-\caption{\label{summary-table} Quick reference sheet for our set of Technical Dimensions.}
+\caption{\label{summary-table2} Quick reference sheet for our set of Technical Dimensions.}
 \end{table}

### 2022-01-30 prog-2022/summary-table.tex
Fix remaining TODOs in table

@@
 Must we customize \emph{running} programs differently to \emph{inert} ones? Do these changes last beyond termination? &
-TODO \\
+Source code vs. config files, Developer Tools tab, auto image-based persistence, scripting language \\
 Externalizability &
 Which portions of the system's state can be referenced and transferred to/from it? &
-TODO \\
-Open Authoring &
+None (state is private) vs. all state exposed as human-legible, CSS-like addressing \\
+Additive Authoring &
 How far can the system's behavior be changed by \emph{adding} expressions? &
-TODO \\
+None (requires power to change original) vs. full (anything can be overridden repeatedly) \\
 Self-Sustainability &
 How far can the system’s behavior be changed from within? &
-TODO \\
+None (rely on extenal tools) vs. self-sufficient (contains everything needed) \\
 \hline AUTOMATION &
@@
 What are the social and economic factors that make the system the way it is? &
-TODO \\
+Funding, volunteers, code sharing, Q/A sites vs. documentation, sense of belonging \\
 \hline

### 2022-02-01 prog-2022/prog22-master.tex
Insert table (too big) and copy concise cluster summaries from it
Incorporate last night's discussion

@@
 % before \begin{document} on input line 6.
+\usepackage{changepage} % For the single-page summary table PDF insert.
+\usepackage{pdfpages}
 \begin{document}
 
-\paperdetails{perspective=art, area=tbd}
+\paperdetails{perspective=art, area={Programming systems}}
 
@@
 
-\keywords{paper, lorem ipsum}
+\keywords{Programming Systems, Dimensions, Design, Framework, Analysis}
 
@@
    <concept>
-<concept_id>10002944.10011122.10003459</concept_id>
-<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
+       <concept_id>10011007.10011006.10011066.10011069</concept_id>
+       <concept_desc>Software and its engineering~Integrated and visual development environments</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003120.10003121.10003129</concept_id>
+       <concept_desc>Human-centered computing~Interactive systems and tools</concept_desc>
        <concept_significance>300</concept_significance>
@@
    <concept>
-<concept_id>10010405.10010476.10010477</concept_id>
-<concept_desc>Applied computing~Publishing</concept_desc>
+       <concept_id>10003120.10003121.10003122</concept_id>
+       <concept_desc>Human-centered computing~HCI design and evaluation methods</concept_desc>
        <concept_significance>300</concept_significance>
@@
 \end{CCSXML}
-\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}
+\ccsdesc[500]{Software and its engineering~Integrated and visual development environments}
+\ccsdesc[300]{Human-centered computing~Interactive systems and tools}
+\ccsdesc[300]{Human-centered computing~HCI design and evaluation methods}
 
@@
 They are informal, guided by the personal vision of the authors and thus are only evaluable and comparable on the basis of individual experience using them.
-This fails to form a coherent body of research, since it is unclear how to build on past work.
+They fail to form a coherent body of research, since it is unclear how to build on past work.
 In the research world, much has been said and done that allows comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems}; we believe that programming systems deserve a theory too.
@@
 By providing foundations for more systematic research in this area, we can help the designers of future programming systems to stand, at last, on the shoulders of giants.
+
+NOTE TO REVIEWERS: The main contribution of the paper is a comprehensive survey of 21 design dimensions of programming systems.
+Each one comes with detailed discussion, including known uses, to properly motivate it.
+We also elect to include the \emph{full set} of these dimensions in the main text.
+While these two aspects do make the paper longer than the recommended page limit, we prefer to err on this side than to privilege certain dimensions and relegate others to an appendix.
+We invite the reader to focus on the ones that best suit their interest.
 \end{abstract}

### 2022-05-18 prog-2022/summary-table.tex
Add factoring of complexity and rename automation to complexity

@@
 None (rely on extenal tools) vs. self-sufficient (contains everything needed) \\
-\hline AUTOMATION &
-\summ{How far does the system remove the need to spell out implementation in minute detail?}
-Degrees of Automation &
+\hline COMPLEXITY &
+\summ{How does the system structure complexity and what level of detail is required?}
+Factoring of Complexity &
+What programming details are hidden in reusable components and how? &
+Domain-specific (more hiding) vs. general-purpose (less hiding) programming \\
+Level of Automation &
 What part of program logic does not need to be explicitly specified? &

### 2022-05-23 prog-2022/prog22-master.tex
Edits in sections 1 and 2

@@
 \begin{abstract}
-Many programming systems go beyond programming languages.
-Programming is usually done in the context of a stateful environment, beyond just writing code, by interacting with a system through a graphical user interface.
-Much research effort focuses on building programming systems that are easier to use, accessible to non-experts, moldable and/or powerful, but such efforts are often disconnected.
+Programming requires much more than just writing code in a programming language.
+It is usually done in the context of a stateful environment, by interacting with a system through a graphical user interface.
+Much research effort focuses on building programming systems that are easier to use, accessible to non-experts, moldable and/or powerful, but such efforts are disconnected.
 They are informal, guided by the personal vision of the authors and thus are only evaluable and comparable on the basis of individual experience using them.
-They fail to form a coherent body of research, since it is unclear how to build on past work.
-In the research world, much has been said and done that allows comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems}; we believe that programming systems deserve a theory too.
-We examine some influential past programming systems and review their stated design principles, technical capabilities, and styles of user interaction.
-We propose a framework of \emph{technical dimensions} which capture the underlying system characteristics and provide a means for conceptualizing and comparing programming systems.
-Since these characteristics may be compared or advanced independently, it should be easier to talk about programming systems in a way that can be shared and constructively debated rather than relying solely on personal impressions.
-By providing foundations for more systematic research in this area, we can help the designers of future programming systems to stand, at last, on the shoulders of giants.
+In other words, work on programming systems fails to form a coherent body of research, making it hard to improve on past work and advance the state of the art.
 
-NOTE TO REVIEWERS: The main contribution of the paper is a comprehensive survey of 21 design dimensions of programming systems.
+In computer science, much has been said and done to allows comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems}; we believe that programming systems deserve a theory too.
+We present a framework of \emph{technical dimensions} which capture the underlying characteristics of programming systems and provide a means for conceptualizing and comparing them.
+We identify technical dimensions by examining past influential programming systems and reviewing their design principles, technical capabilities, and styles of user interaction.
+
+Technical dimensions capture characteristics that may be studied, compared and advanced indepently.
+This makes it possible to talk about programming systems in a way that can be shared and constructively debated rather than relying solely on personal impressions.
+We outline two concrete ways of using our framework.
+First, we show how to use it to analyze a recently developed novel programming system, and, second, we use it to identify an interesting unexplored point in the design space of programming systems.
+By providing foundations for more systematic research, we can help programming systems researchers to stand, at last, on the shoulders of giants.
+
+%NOTE TO REVIEWERS: The main contribution of the paper is a comprehensive survey of 21 design dimensions of programming systems.
 Each one comes with detailed discussion, including known uses, to properly motivate it.

### 2022-05-27 prog-2022/prog22-master.tex
copyedit, tweak etc.

@@
 
-In computer science, much has been said and done to allows comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems}; we believe that programming systems deserve a theory too.
+In computer science, much has been said and done to allow comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems}; we believe that programming systems deserve a theory too.
 We present a framework of \emph{technical dimensions} which capture the underlying characteristics of programming systems and provide a means for conceptualizing and comparing them.
@@
 
-Technical dimensions capture characteristics that may be studied, compared and advanced indepently.
+Technical dimensions capture characteristics that may be studied, compared and advanced independently.
 This makes it possible to talk about programming systems in a way that can be shared and constructively debated rather than relying solely on personal impressions.

### 2022-05-31 prog-2022/prog22-master.tex
Start methodology Appendix + minor review changes
Perform appendectomy
Remove most material from body and reattach as appendix

@@
 \usepackage{pdfpages}
+\usepackage{pifont} % For checkmark / cross symbols for Appendix table :)
+
+% Thanks https://tex.stackexchange.com/a/32687
+\NewDocumentCommand{\rot}{O{45} O{1em} m}{\makebox[#2][l]{\rotatebox{#1}{#3}}}%
+
 \begin{document}
@@
 
-%NOTE TO REVIEWERS: The main contribution of the paper is a comprehensive survey of 21 design dimensions of programming systems.
+NOTE TO REVIEWERS: The main contribution of the paper is a comprehensive survey of 21 design dimensions of programming systems.
 Each one comes with detailed discussion, including known uses, to properly motivate it.
-We also elect to include the \emph{full set} of these dimensions in the main text.
-While these two aspects do make the paper longer than the recommended page limit, we prefer to err on this side than to privilege certain dimensions and relegate others to an appendix.
-We invite the reader to focus on the ones that best suit their interest.
+Regrettably, we only have space for a small number of these dimensions in the main paper body.
+To give a broader picture of what we are contributing, we invite the reader to explore Appendix\ \ref{dimensions-catalogue} for dimensions that suit their interest.
 \end{abstract}
@@
 
-% \appendix
-% \input{prog22-appendix.tex}
+\appendix
+\input{prog22-appendix.tex}
 

### 2022-06-01 prog-2022/prog22-master.tex
Submission-ready

@@
 We outline two concrete ways of using our framework.
-First, we show how to use it to analyze a recently developed novel programming system, and, second, we use it to identify an interesting unexplored point in the design space of programming systems.
+First, we show how it can analyze a recently developed novel programming system.
+Then, we use it to identify an interesting unexplored point in the design space of programming systems.
 By providing foundations for more systematic research, we can help programming systems researchers to stand, at last, on the shoulders of giants.

### 2022-06-07 prog-2022/prog22-master.tex
Rework Abstract

@@
 \begin{abstract}
-Programming requires much more than just writing code in a programming language.
+  \emph{Context.} Programming requires much more than just writing code in a programming language.
 It is usually done in the context of a stateful environment, by interacting with a system through a graphical user interface.
-Much research effort focuses on building programming systems that are easier to use, accessible to non-experts, moldable and/or powerful, but such efforts are disconnected.
-They are informal, guided by the personal vision of the authors and thus are only evaluable and comparable on the basis of individual experience using them.
-In other words, work on programming systems fails to form a coherent body of research, making it hard to improve on past work and advance the state of the art.
+Yet, this wide space of possibilities lacks a common structure for navigation.
+Work on programming systems fails to form a coherent body of research, making it hard to improve on past work and advance the state of the art.
 
-In computer science, much has been said and done to allow comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems}; we believe that programming systems deserve a theory too.
-We present a framework of \emph{technical dimensions} which capture the underlying characteristics of programming systems and provide a means for conceptualizing and comparing them.
-We identify technical dimensions by examining past influential programming systems and reviewing their design principles, technical capabilities, and styles of user interaction.
+  \emph{Inquiry.} In computer science, much has been said and done to allow comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems;} we believe that programming systems deserve a theory too.
 
+  \emph{Approach.} We present a framework of \emph{technical dimensions} which capture the underlying characteristics of programming systems and provide a means for conceptualizing and comparing them.
+
+  \emph{Knowledge.} We identify technical dimensions by examining past influential programming systems and reviewing their design principles, technical capabilities, and styles of user interaction.
 Technical dimensions capture characteristics that may be studied, compared and advanced independently.
 This makes it possible to talk about programming systems in a way that can be shared and constructively debated rather than relying solely on personal impressions.
+
+  \emph{Grounding.} Our framework is derived using a qualitative analysis of past programming systems.
 We outline two concrete ways of using our framework.
@@
 Then, we use it to identify an interesting unexplored point in the design space of programming systems.
+
+  \emph{Importance.} Much research effort focuses on building programming systems that are easier to use, accessible to non-experts, moldable and/or powerful, but such efforts are disconnected.
+They are informal, guided by the personal vision of the authors and thus are only evaluable and comparable on the basis of individual experience using them.
 By providing foundations for more systematic research, we can help programming systems researchers to stand, at last, on the shoulders of giants.
 
-NOTE TO REVIEWERS: The main contribution of the paper is a comprehensive survey of 21 design dimensions of programming systems.
+NOTE TO REVIEWERS: The main contribution of the paper is a comprehensive survey of 22 design dimensions of programming systems.
 Each one comes with detailed discussion, including known uses, to properly motivate it.

### 2022-09-14 onward-2022/onward22-master.tex
Onward! '22 submission

\documentclass[sigplan,review]{acmart}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{amsthm}

\acmConference{Onward! Essays}{December 5--10, 2022}{Auckland, New Zealand}

\title{Ascending the Ladder to Self-Sustainability}
\subtitle{Achieving open evolution in an interactive graphical system}
\author{Joel Jakubovic}
\affiliation{\institution{University of Kent} \city{Canterbury} \country{United Kingdom}}
\email{jdj9@kent.ac.uk}
\author{Tomas Petricek}
\affiliation{\institution{Charles University} \city{Prague} \country{Czechia}}
\email{tomas@tomasp.net}

\begin{document}

\lstset{breaklines=true, columns=fullflexible, breakatwhitespace=true, basicstyle=\ttfamily\footnotesize}
\lstdefinelanguage{JavaScript}{
  keywords={break, case, catch, continue, debugger, default, delete, do, else, finally, for, function, if, in, instanceof, new, return, switch, this, throw, try, typeof, var, void, while, with, let, const},
  morecomment=[l]{//},
  morecomment=[s]{/*}{*/},
  morestring=[b]',
  morestring=[b]",
  sensitive=true
}

\newcommand{\joel}[1]{}
\newcommand{\tomas}[1]{}
\newcommand{\note}[1]{}
\newcommand{\notes}[1]{}
\newcommand{\todo}[1]{\textbf{TODO: #1}}
\newcommand{\delete}[1]{\textbf{DELETE:}#1}
\providecommand{\tightlist}{}% Don't want Pandoc's tight lists
% Nor do we want long tables by default
\newenvironment{longtable}[2]{\begin{tabular}}{\end{tabular}}
\newenvironment{head}{}{}
\theoremstyle{definition}
\newtheorem{force}{Force}
\newtheorem{heuristic}{Heuristic}
\newtheorem{requirement}{Requirement}
\newcommand{\naive}{na\"ive}
%\newcommand{\hypertarget}[1]{}

\input{onward22-manuscript.tex}

\appendix
\input{onward22-appendix.tex}

\bibliographystyle{plainurl}
\bibliography{onward22}
\end{document}

### 2022-09-21 prog-2022/changelog.tex

\section*{Technical Dimensions of Programming Systems: major revisions}

We graciously thank the reviewers for their detailed feedback. We've
made many minor changes to the paper and several major changes to both
structure and content.

Two issues particularly stood out:

\paragraph{Definition of ``system''.}

All three reviewers requested clarification on what constitutes a
``system'', mentioning IDEs amongst other things as having unclear
membership. Unfortunately, we truly don't have a definition behind the
scenes which we're using to include or exclude examples. It's more of a
``family resemblance'' between things that generalize ``programming
languages'' to ``software which lets one program, yet doesn't have to be
a language''. Nevertheless, we thought it might be worth adding a new
classification scheme for the examples we mention, which should get
across the key ``reference classes'' from which we're drawing. In sum,
these are: things that are like operating systems (O-type), things that
are like domain-specific applications (A-type) and the systems that are
based around a programming language (L-type). We also went into detail
about how a programming language (Java) can be interpreted as a system
in several different ways. These were all additions to Section 3.

\paragraph{But how do you use it?}

Two reviewers were concerned that it's hard to see how to use the
framework. We've rectified this by including a new, concrete
``Discussion'' section with two halves: one using the framework for
analysis, the other for synthesis. Specifically, we analyze Dark, a
recent programming system, along many of the dimensions. Then, we plot
two dimensions (self-sustainability vs.~notational diversity) against
each other to reveal an unexplored gap. The methodology we used to turn
qualitative dimensions into numbers is given in a new Appendix. Various
points earlier in the paper (e.g.~the Abstract and Figure 1) have been
updated to set up for this new section.

Other changes:

\begin{itemize}
\tightlist
\item
  We've significantly reworked Section 3 to be more to-the-point about
  the example systems and less of a historical narrative.
\item
  We've argued for why we think ``sociability'' belongs as a technical
  dimension in the relevant section.
\item
  We've reorganized and re-worded parts of Section 2. We've also made
  the part about how the framework relates to Cognitive Dimensions in
  Section 2, under ``Already-known characteristics'', more explicit.
\item
  We've clarified some of the points about Direct Manipulation
  (primitive actions) and Immediate Feedback (pure code previews).
\item
  Under ``Addressing and Externalizability'', we've made it explicit
  that CSS provides Additive Authoring for the Web programming system.
\item
  To make room for the ``Discussion'' section, we've had to move more
  material to the existing Appendix with the bulk of the dimensions.
  Now, the main body only contains the Customizability dimensions.
\item
  We've elevated the singleton Automation cluster to a doublet by adding
  a Factoring of Complexity dimension.
\end{itemize}

### 2022-09-21 prog-2022/diff-prog22-manuscript.tex

\begin{quote}
A systematic presentation removes ideas from the ground that made them
grow and arranges them in an artificial pattern.

--- Paul Feyerabend, \emph{The Tyranny of Science}, Polity Press (2011)
\end{quote}

\begin{quote}
Irony is said to allow the artist to continue his creative production
while immersed in a sociocultural context of which he is critical.

--- Emmanuel Petit, \emph{Irony or, the Self-Critical Opacity of
Postmodernist Architecture}, Yale (2013)
\end{quote}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Many forms of software have been developed to enable programming. The
classic form consists of a \emph{programming language}, a text editor to
enter source code, and a compiler to turn it into an executable program.
Instances of this form are differentiated by the syntax and semantics of
the language, along with the implementation techniques in the compiler
or runtime environment. Since the advent of graphical user interfaces
(GUIs), programming languages can be found embedded within graphical
environments that increasingly define how programmers work with the
language---for instance, by directly supporting debugging or
refactoring. Beyond this, the rise of GUIs also permits diverse visual
forms of programming, including visual languages and GUI-based end-user
programming tools. This paper advocates a shift of attention from
\emph{programming languages} to the more general notion of ``software
that enables programming''---in other words, \emph{programming systems}.

A \emph{programming system} may include tools, protocols, notations,
interfaces, and languages. It is a software artifact that makes it
possible to construct programs, debug them, and turn them into
operational, maintained, and evolvable artifacts running on appropriate
hardware. This notion covers classic programming languages together with
their editors, debuggers, compilers, and other tools. Yet it is
intentionally broad enough to accommodate image-based programming
environments like Smalltalk, operating systems like UNIX, and hypermedia
authoring systems like Hypercard, in addition to various other examples
we will mention.

\hypertarget{what-is-the-problem}{%
\subsection{What is the problem?}\label{what-is-the-problem}}

There is a growing interest in broader forms of \emph{programming
systems}, both in the programming research community and in industry. On
the one hand, researchers are studying topics such as \emph{programming
experience} and \emph{live programming} that require considering not
just the \emph{language}, but further aspects of a given system. On the
other hand, commercial companies are building new programming
environments like Replit \cite{ReplitWeb} or low-code tools like Dark
\cite{DarkWeb} and Glide \cite{GlideWeb}. Yet, such topics remain at the
sidelines of mainstream programming research. While \emph{programming
languages} are a well-established concept, analysed and compared in a
common vocabulary, no similar foundation exists for the wider range of
\emph{programming systems}.

The academic research on programming suffers from this lack of common
vocabulary. While we may thoroughly assess programming \emph{languages},
as soon as we add interaction or graphics into the picture, we often get
stuck on how the resulting system is vaguely ``interesting''. Moreover,
when designing new systems, inspiration is often drawn from the same few
standalone sources of ideas. These might be influential past systems
like Smalltalk, programmable end-user applications like spreadsheets, or
motivational illustrations by thinkers like Victor~\cite{BretVictor}.

Instead of forming a solid body of work, the ideas that emerge are
difficult to relate to each other. The research methods used to study
programming systems lack the rigorous structure of programming language
research methods. They tend to rely on singleton examples, which
demonstrate the author's ideas, but are inadequate methods for comparing
new ideas with the work of others. This makes it hard to build on top
and thereby advance the state of the art.

Studying \emph{programming systems} is not merely about taking a
programming language and looking at the tools that surround it. It
presents a \emph{paradigm shift} to a perspective that is, at least
partly, \emph{incommensurable} with that of languages. When studying
programming languages, everything that matters is in the program code;
when studying programming systems, everything that matters is in the
\emph{interaction} between the programmer and the system. As documented
by Gabriel \cite{PLrev}, looking at a \emph{system} from a
\emph{language} perspective makes it impossible to think about concepts
that arise from interaction with a system, but are not reflected in the
language. Thus, we must proceed with some caution. As we will see, when
we talk about Lisp as a programming system, we mean something very
different from a parenthesis-heavy programming language!

\hypertarget{contributions}{%
\subsection{Contributions}\label{contributions}}

We propose a common language as an initial step towards a more
progressive research on programming systems. Our set of \emph{technical
dimensions} seeks to break down the holistic view of systems along
various specific ``axes''. The dimensions identify a range of possible
design choices, characterized by two extreme points in the design space.
They are not quantitative, but they allow comparison by locating systems
on a common axis. We do not intend for the extreme points to represent
``good'' or ``bad'' designs; we expect any position to be a result of
design trade-offs. At this stage we encourage agreement on descriptions
of systems first, in order to settle any normative judgements later.

The set of dimensions can be understood as a map of the design space of
programming systems (Figure~\ref{fig:tech-dims-diagram}). Past and
present systems will serve as landmarks, and with enough of them,
unexplored or overlooked possibilities will reveal themselves. So far,
the field has not been able to establish a virtuous cycle of feedback;
it is hard for practitioners to situate their work in the context of
others' so that subsequent work can improve on it. Our aim is to provide
foundations for the study of programming systems that would allow such
development.

This paper is intended as a reference on the current state of the
technical dimensions framework and it is meant to be \emph{used} rather
than \emph{read}. We present the dimensions in detail, but encourage the
reader to skim through the details on the first read. Subsequently, we
suggest revisiting dimensions which seem relevant to a concrete system
known to the reader. The main contributions of this paper are organized
as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  In Section~\ref{programming-systems}, we characterize what a
  programming system is and review landmark programming systems of the
  past that are used as examples throughout this paper, as well as to
  delineate our notion of a programming system.
\item
  In Section~\ref{technical-dimensions}, we present the technical
  dimensions in details, organised into related clusters:
  \emph{interaction}, \emph{notation}, \emph{conceptual structure},
  \emph{customizability}, \emph{complexity}, \emph{errors}, and
  \emph{adoptability}. For each technical dimension, we give examples
  that illustrate a number of values along the axis identified by the
  dimension.
\item
  In Section~\ref{discussion}, we sketch two ways of using the technical
  dimensions framework. In Section~\ref{evaluating-programming-systems},
  we use it to evaluate a recent interesting programming system and in
  Section~\ref{exploring-the-design-space}, we use it to identify an
  unexplored point in the design space and envision a potential novel
  programming system.
\end{enumerate}

\begin{figure}
  \centering
  \includegraphics[width=0.6\linewidth]{plot-figure0.pdf}
  \caption{One 2-dimensional slice of the space of possible systems, to be examined in more detail in Section\ \ref{exploring-the-design-space}.\label{fig:tech-dims-diagram}}
\end{figure}

\hypertarget{related-work}{%
\section{Related work}\label{related-work}}

While we do have new ideas to propose, part of our contribution is
integrating a wide range of \emph{existing} concepts under a common
umbrella. This work is spread out across different domains, but each
part connects to programming systems or focuses on a specific
characteristic they may have.

\paragraph{From languages to systems.}

Our approach lies between a narrow focus on programming languages and a
broad focus on programming as a socio-political and cultural subject.
Our concept of a programming system is technical in scope, although we
acknowledge the technical side often has important social implications
as in the case of the ``Adoptability'' dimension
(Section~\ref{adoptability}). This contrasts with the more
socio-political focus found in Tchernavskij \cite{TcherDiss} or in
software studies \cite{SwStudies}. It overlaps with Kell's
conceptualization of UNIX, Smalltalk, and Operating Systems
generally~\cite{KellOS}, and we have ensured that UNIX has a place in
our framework.

The distinction between more narrow \emph{programming languages} and
broader \emph{programming systems} is more subtle. Richard Gabriel noted
an invisible paradigm shift from the study of ``systems'' to the study
of ``languages'' in computer science \cite{PLrev}, and this observation
informs our distinction here. One consequence of the change is that a
\emph{language} is often formally specified apart from any specific
implementations, while \emph{systems} resist formal specification and
are often \emph{defined by} an implementation. We recognize programming
language implementations as a \emph{small region} of the space of
possible systems (Figure~\ref{fig:tech-dims-diagram}). Hence we refer to
the \emph{interactive programming system} aspects of languages, such as
text editing and command-line workflow.

\paragraph{Programming systems research.}

There is renewed interest in programming systems in both industry and
research, but the holistic programming systems view is more often
adopted in work on non-traditional programming tools:

\begin{itemize}
\tightlist
\item
  Computational notebooks such as Jupyter simplify data analysis by
  combining code snippets with text and visual output. They are backed
  by stateful ``kernels'' and used interactively.
\item
  ``Low code'' end-user programming systems allow application
  development (mostly) through a GUI. One example is Coda
  \cite{CodaWeb}, which combines tables, formulas, and scripts to enable
  non-technical people to build ``applications as documents''.
\item
  Domain-specific programming systems such as Dark, which claims a
  ``holistic'' programming experience for cloud API services. This
  includes a language, a direct manipulation editor, and
  near-instantaneous building and deployment.
\item
  Even for general purpose programming with conventional tools, systems
  like Replit have demonstrated the benefits of integrating all needed
  languages, tools, and user interfaces into a seamless experience,
  available from the browser, that requires no setup.
\end{itemize}

Research that follows the programming systems perspective can be found
in a number of research venues. Those include Human-Computer Interaction
conferences such as \href{https://uist.acm.org/}{UIST}\footnote{ACM
  Symposium on User Interface Software and Technology} and
\href{https://conferences.computer.org/VLHCC/}{VL/HCC}\footnote{IEEE
  Symposium on Visual Languages and Human-Centric Computing}. However,
work in those often emphasizes the user experience over technical
description. Programming systems are often presented in workshops such
as \href{https://liveprog.org/}{LIVE} and
\href{https://2021.programming-conference.org/home/px-2021}{PX}\footnote{Programming
  eXperience}. However, work in those venues is often limited to the
authors' individual perspectives and suffers from the aforementioned
difficulty of comparing to other systems.

Concrete examples of systems appear throughout the paper. Recent systems
which motivated some of our dimensions include Subtext \cite{Subtext},
which combines code with its live execution in a single editable
representation; Sketch-n-sketch \cite{SnS}, which can synthesize code by
direct manipulation of its outputs; Hazel \cite{Hazel}, a live
functional programming environment with typed holes to enable execution
of incomplete or ill-typed programs; and Webstrates \cite{Webstrates},
which extends Web pages with real-time sharing of state.

\paragraph{Already-known characteristics.}

There are several existing projects identifying characteristics of
programming systems. Some revolve around a single one, such as levels of
liveness \cite{Liveness}, or plurality and communicativity
\cite{KellComm}. Others propose an entire collection. \emph{Memory
Models of Programming Languages}~\cite{MemMod} identifies the
``everything is an X'' metaphors underlying many programming languages;
the \emph{Design Principles of Smalltalk}~\cite{STdesign} documents the
philosophical goals and dicta used in the design of Smalltalk; the
``Gang of Four'' \emph{Design Patterns}~\cite{DesPats} catalogues
specific implementation tactics; and the \emph{Cognitive Dimensions of
Notation}~\cite{CogDims} introduces a common vocabulary for software's
\emph{notational surface} and for identifying their trade-offs.

The latter two directly influence our proposal. Firstly, the Cognitive
Dimensions are a set of qualitative properties which can be used to
analyze \emph{notations}. We wish to extend this approach to the
``rest'' of a system, beyond its notation, with \emph{Technical}
Dimensions. Secondly, our individual dimensions naturally fall under
larger \emph{clusters} that we present in a regular format, similar to
the presentation of the classic Design Patterns. As for characteristics
identified by others, part of our contribution is to integrate them
under a common umbrella: liveness, pluralism, and uniformity metaphors
(``everything is an X''), which have already been identified by others,
become dimensions in our framework.

\paragraph{Methodology.}

We follow the attitude of \emph{Evaluating Programming
Systems}~\cite{EvProgSys} in distinguishing our work from HCI methods
and empirical evaluation. We are generally concerned with
characteristics that are not obviously amenable to statistical analysis
(e.g.~mining software repositories) or experimental methods like
controlled user studies, so numerical quantities are generally not
featured.

Similar development seems to be taking place in HCI research focused on
user interfaces. The UIST guidelines \cite{UISTAuthor} instruct authors
to evaluate system contributions holistically, and the community has
developed heuristics for such evaluation, such as \emph{Evaluating User
Interface Systems Research}~\cite{EvUISR}. Our set of dimensions offers
similar heuristics for identifying interesting aspects of programming
systems, though they focus more on underlying technical properties than
the surface interface.

Finally, we believe that the aforementioned paradigm shift from
programming systems to programming languages has hidden many ideas about
programming that are worthwhile recovering and developing further
\cite{ComplementaryBasic}. Thus our approach is related to the idea of
\emph{complementary science} developed by Chang~\cite{Chang} in the
context of history and philosophy of science. Chang argues that even in
disciplines like physics, superseded or falsified theories may still
contain interesting ideas worth documenting. In the field of
programming, where past systems are discarded for many reasons besides
empirical failure, Chang's \emph{complementary science} approach seems
particularly suitable.

\paragraph{Programming systems deserve a theory too!}

In short, while there is a theory for programming languages, programming
\emph{systems} deserve a theory too. It should apply from the small
scale of language implementations to the vast scale of operating
systems. It should be possible to analyse the common and unique features
of different systems, to reveal new possibilities, and to build on past
work in an effective manner. In Kuhnian terms, it should enable a body
of ``normal science'': filling in the map of the space of possible
systems (Figure \ref{fig:tech-dims-diagram}), thereby forming a
knowledge repository for future designers.

\hypertarget{programming-systems}{%
\section{Programming systems}\label{programming-systems}}

We introduce the notion of a \emph{programming system} through a number
of examples, classified following the style of Lehman \cite{SPEPrograms}
into three broad types. These are loosely inspired by the notions of
\emph{languages}, \emph{operating systems} and \emph{applications}:

\begin{itemize}
\item
  \textbf{L-type programming systems} are software ecosystems built
  around a text-based programming \emph{language}. They consist of a set
  of tools such as compilers, debuggers, and profilers. These tools may
  exist as separate command-line programs, or within an Integrated
  Development Environment.
\item
  \textbf{O-type programming systems} resemble \emph{operating systems}
  in that they structure the execution environment and encompass the
  resources of an entire machine (physical or virtual). They provide a
  common interface for communication, both between the user and the
  computer, and between programs themselves.
\item
  \textbf{A-type programming systems} are programmable
  \emph{applications}. They are typically optimized for a specific
  domain, and offer a limited degree of programmability which may be
  increased with newer versions.
\end{itemize}

We will illustrate each of these types with several examples. This will
provide an intuition for the notion of a programming system and
establish a collection of go-to examples for the rest of the paper.

\hypertarget{l-type-programming-systems}{%
\subsection{L-type programming
systems}\label{l-type-programming-systems}}

We see programming systems as (collections of) software artifacts that
make it possible to construct programs, debug them, and turn them into
operational, maintained, and evolvable artifacts running on an
appropriate hardware. Text-based programming languages sit within
programming systems whose boundaries are not explicitly defined. To
speak of a programming system, we need to consider a language with, at
minimum, an editor and a compiler or interpreter. However, the exact
boundaries are a design choice that significantly affects our analysis.

\paragraph{Java with the Eclipse ecosystem.}

Java cannot be viewed as a programming system on its own, but it is one
if we consider it as embedded in an ecosystem of tools. There are
multiple ways to delineate this, resulting in different analyses. A
minimalistic programming system would consist of a text editor to write
Java code and a command line compiler. A more realistic system is Java
as embedded in the Eclipse IDE. The programming systems view allows us
to see all there is beyond the textual code. In the case of Eclipse,
this includes the debugger, refactoring tools, testing and modelling
tools, GUI designers, and so on. This delineation yields a programming
system that is powerful and convenient, but has a large number of
concepts and secondary notations.

\paragraph{Haskell tools ecosystem.}

Haskell is an even more language-focused programming system. It is used
through the command-line \emph{GHC} compiler and \emph{GHCi} REPL,
alongside a text editor that provides features like syntax highlighting
and auto-completion. In general, we can consider any editor that
supports the Language Server Protocol \cite{LSP}.

Haskell is mathematically rooted and relies on mathematical intuition
for understanding many of its concepts. This background is also
reflected in the notations it uses. In addition to the concrete language
syntax for writing code, the ecosystem also uses an informal
mathematical notation for writing about Haskell (e.g.~in academic papers
or on the whiteboard). This provides an additional tool for manipulating
Haskell programs. Experiments on paper can provide a kind of rapid
feedback that other systems may provide through live programming.

\paragraph{From REPLs to notebooks.}

A different kind of developer ecosystem that evolved around a
programming language is the Jupyter notebook platform. In Jupyter, data
scientists write scripts divided into notebook cells, execute them
interactively and see the resulting data and visualizations directly in
the notebook itself. This brings together the REPL, which dates back to
conversational implementations of Lisp in the 1960s, with literate
programming \cite{LiterateProg} used in the late 1980s in Mathematica
1.0.

As a programming system, Jupyter has a number of interesting
characteristics. The primary outcome of programming is the notebook
itself, rather than a separate application to be compiled and run. The
code lives in a document format, interleaved with other notations. Code
is written in small parts that are executed quickly, offering the user
more rapid feedback than in conventional programming. A notebook can be
seen as a trace of how the result has been obtained, yet one often
problematic feature of notebooks is that some allow the user to run code
blocks out-of-order. The code manipulates mutable state that exists in a
``kernel'' running in the background. Thus, retracing one's steps in a
notebook is more subtle than in, say, Common Lisp, where the
\texttt{dribble} function would directly record the user's session to a
file.

\hypertarget{o-type-programming-systems}{%
\subsection{O-type programming
systems}\label{o-type-programming-systems}}

The first O-type programming systems emerged in the 1960s when it became
possible to interact one-on-one with a computer. First, time-sharing
systems enabled interactive shared use of a computer via a teletype;
smaller computers such as the PDP-1 and PDP-8 provided similar direct
interaction, while 1970s workstations such as the Alto and Lisp Machines
added graphical displays and mouse input.

\paragraph{MacLisp and Interlisp.}

LISP 1.5 \cite{LISP15} arrived before the rise of interactive computers,
but the existence of an interpreter and the absence of declarations made
it natural to use Lisp interactively, with the first such
implementations appearing in the early 1960s. Two branches of the Lisp
family \cite{LispEvolve}, MacLisp and the later Interlisp, embraced the
interactive ``conversational'' way of working, first through a teletype
and later using the screen and keyboard.

Both MacLisp and Interlisp adopted the idea of \emph{persistent address
space}. Both program code and program state were preserved when powering
off the system, and could be accessed and modified interactively as well
as programmatically using the \emph{same means}. Lisp Machines embraced
the idea that the machine runs continually and saves the state to disk
when needed. Today, this is widely seen in cloud-based services like
Google Docs and online IDEs. Another idea pioneered in MacLisp and
Interlisp was the use of \emph{structure editors}. These let programmers
work with Lisp data structures not as sequences of characters, but as
nested lists. In Interlisp, the programmer would use commands such as
\texttt{*P} to print the current expression, or \texttt{*(2\ (X\ Y))} to
replace its second element with the argument \texttt{(X\ Y)}. The PILOT
system \cite{Pilot} offered even more sophisticated conversational
features. For typographical errors and other slips, it would offer an
automatic fix for the user to interactively accept, modifying the
program in memory and resuming execution.

\paragraph{Smalltalk.}

Smalltalk appeared in the 1970s with a distinct ambition of providing
``dynamic media which can be used by human beings of all ages''
\cite{PersonalDynMedia}. The authors saw computers as \emph{meta-media}
that could become a range of other media for education, discourse,
creative arts, simulation and other applications not yet invented.
Smalltalk was designed for single-user workstations with a graphical
display, and pioneered this display not just for applications but also
for programming itself. In Smalltalk 72, one wrote code in the bottom
half of the screen using a structure editor controlled by a mouse, and
menus to edit definitions. In Smalltalk-76 and later, this had switched
to text editing embedded in a \emph{class browser} for navigating
through classes and their methods.

Similarly to Lisp systems, Smalltalk adopts the persistent address space
model of programming where all objects remain in memory, but based on
\emph{objects} and \emph{message passing} rather than \emph{lists}. Any
changes made to the system state by programming or execution are
preserved when the computer is turned off. Lastly, the fact that much of
the Smalltalk environment is implemented in itself makes it possible to
extensively modify the system from within.

\paragraph{UNIX.}

We included Lisp and Smalltalk in the O-type because they function as
operating systems in many ways. On specialized machines, like the Xerox
Alto and Lisp machines, the user started their machine directly in the
Lisp or Smalltalk environment and was able to do everything they needed
from \emph{within} the system. Nowadays, however, this experience is
associated with UNIX and its descendants on a vast range of commodity
machines.

UNIX illustrates the fact that many aspects of programming systems are
shaped by their intended target audience. Built for computer hackers,
its abstractions and interface are close to the machine. Although
historically linked to the C language, UNIX developed a
language-agnostic set of abstractions that make it possible to use
multiple programming languages in a single system. While everything is
an object in Smalltalk, the ontology of the UNIX system consists of
files, memory, executable programs, and running processes. Note the
explicit ``stage'' distinction here: UNIX distinguishes between volatile
\emph{memory} structures, which are lost when the system is shut down,
and non-volatile \emph{disk} structures that are preserved. This
distinction between types of memory is considered, by Lisp and
Smalltalk, to be an implementation detail to be abstracted over by their
persistent address space. Still, this did not prevent the UNIX ontology
from supporting a pluralistic ecosystem of different languages and
tools.

\paragraph{Early and late Web.}

The Web evolved from a system for sharing and organizing information to
a \emph{programming system}. Today, it consists of a wide range of
server-side programming tools, JavaScript and languages that compile to
it, and notations like HTML and CSS. As a programming system, the
``modern 2020s web'' is reasonably distinct from the ``early 1990s
web''. In the early web, JavaScript code was distributed in a form that
made it easy to copy and re-use existing scripts, which led to
enthusiastic adoption by non-experts---recalling the birth of
microcomputers like Commodore 64 with BASIC a decade earlier.

In the ``modern web'', multiple programming languages treat JavaScript
as a compilation target, and JavaScript is also used as a language on
the server-side. This web is no longer simple enough to encourage
copy-and-paste remixing of code from different sites. However, it does
come with advanced developer tools that provide functionality resembling
early interactive programming systems like Lisp and Smalltalk. The
\emph{Document Object Model (DOM)} structure created by a web page is
transparent, accessible to the user and modifiable through the built-in
browser inspector tools. Third-party code to modify the DOM can be
injected via extensions. The DOM almost resembles the tree/graph model
of Smalltalk and Lisp images, lacking the key persistence property. This
limitation, however, is being addressed by Webstrates~\cite{Webstrates}.

\hypertarget{a-type-programming-systems}{%
\subsection{A-type programming
systems}\label{a-type-programming-systems}}

The previously discussed programming systems were either universal, not
focusing on any particular kind of application, or targeted at broad
fields, such as Artificial Intelligence and symbolic data manipulation
in Lisp's case. In contrast, the A-type programming systems focus on
more narrow kinds of applications that need to be built. Many support
programming based on rich interactions with specialized visual and
textual notations.

\paragraph{Spreadsheets.}

The first system, VisiCalc, became available in 1979 and helped analysts
perform budget calculations. As programming systems, spreadsheets are
notable for their programming substrate (a two-dimensional grid) and
evaluation model (automatic re-evaluation). The programmability of
spreadsheets developed over time, acquiring features that made them into
powerful programming systems in a way VisiCalc was not. The final step
was the 1993 inclusion of \emph{macros} in Excel, later further extended
with \emph{Visual Basic for Applications}.

\paragraph{HyperCard.}

While spreadsheets were designed to solve problems in a specific
application area, HyperCard \cite{HyperCard} was designed around a
particular application format. Programs are ``stacks of cards''
containing multimedia components and controls such as buttons. These
controls can be programmed with pre-defined operations like ``navigate
to another card'', or via the HyperTalk scripting language for anything
more sophisticated.

As a programming system, HyperCard is interesting for a couple of
reasons. It effectively combines visual and textual notation. Programs
appear the same way during editing as they do during execution. Most
notably, HyperCard supports gradual progression from the ``user'' role
to ``developer'': a user may first use stacks, then go on to edit the
visual aspects or choose pre-defined logic until, eventually, they learn
to program in HyperTalk.

\hypertarget{technical-dimensions}{%
\section{Technical dimensions}\label{technical-dimensions}}

We present our proposed technical dimensions grouped under
\emph{clusters}. The clusters may be regarded as ``topics of interest''
or ``areas of inquiry'' when studying a given system, grouping together
related dimensions against which to measure it. We also include a
concise reference sheet on the next page, though it will make more sense
after reading the relevant sections.

Each cluster is named and opens with a boxed \emph{summary}, followed by
a longer \emph{discussion}, and closes with a list of any
\emph{relations} to other clusters along with any \emph{references} if
applicable. Within the main discussion, individual \emph{dimensions} are
listed. Sometimes, a particular value along a dimension (or a
combination of values along several dimensions) can be recognized as a
familiar pattern---perhaps with a name already established in the
literature. These are marked as \emph{examples}. Finally, interspersed
discussion that is neither a \emph{dimension} nor an \emph{example} is
introduced as a \emph{remark}.

For space reasons, we are only able to include one cluster,
\emph{customizability}, in the main body of this paper. We encourage the
reader to peruse Appendix~\ref{dimensions-catalogue} for the full extent
of our proposed dimensions.

\hypertarget{customizability}{%
\subsection{Customizability}\label{customizability}}

\mybox{Once a program exists in the system, how can it be extended and modified?}

Programming is a gradual process. We start either from nothing, or from
an existing program, and gradually extend and refine it until it serves
a given purpose. Programs created using different programming systems
can be refined to different extents, in different ways, at different
stages of their existence.

\includepdf[pages={2}]{table.pdf}

Consider three examples. First, a program in a conventional programming
language like Java can be refined only by modifying its source code.
However, you may be able to do so by just adding new code, such as a new
interface implementation. Second, a spreadsheet can be modified at any
time by modifying the formulas or data it contains. There is no separate
programming phase. However, you have to modify the formulas directly in
the cell---there is no way of modifying it by specifying a change in a
way that is external to the cell. Third, a \emph{self-sustaining}
programming system, such as Smalltalk, does not make an explicit
distinction between ``programming'' and ``using'' phases, and it can be
modified and extended via itself. It gives developers the power to
experiment with the system and, in principle, replace it with a better
system from within.

\hypertarget{dimension-staging-of-customization}{%
\subsubsection{Dimension: staging of
customization}\label{dimension-staging-of-customization}}

For systems that distinguish between different stages, such as writing
source code versus running a program, customization methods may be
different for each stage. In traditional programming languages,
customization is done by modifying or adding source code at the
programming stage, but there is no (automatically provided) way of
customizing the created programs once they are running.

There are a number of interesting questions related to staging of
customization. First, what is the notation used for customization? This
may be the notation in which a program was initially created, but a
system may also use a secondary notation for customization (consider
Emacs using Emacs Lisp). For systems with a stage distinction, an
important question is whether such changes are \emph{persistent}.

\emph{Smalltalk, Interlisp and similar.} In image-based programming
systems, there is generally no strict distinction between stages and so
a program can be customized during execution in the same way as during
development. The program image includes the programming environment.
Users of a program can open this, navigate to a suitable object or a
class (which serve as the \emph{addressable extension points}) and
modify that. Lisp-based systems such as \emph{Interlisp} follow a
similar model. Changes made directly to the image are persistent. The
PILOT system for Lisp \cite{Pilot} offers an interactive way of
correcting errors when a program fails during execution. Such
corrections are then applied to the image and are thus persistent.

\emph{Document Object Model (DOM) and Webstrates}: In the context of Web
programming, there is traditionally a stage distinction between
programming (writing the code and markup) and running (displaying a
page). However, the DOM can be also modified by browser Developer
Tools---either manually, by running scripts in a console, or by using a
userscript manager such as Greasemonkey. Such changes are not persistent
in the default browser state, but are made so by Webstrates
\cite{Webstrates} which synchronize the DOM between the server and the
client. This makes the DOM collaborative, but not (automatically)
\emph{live} because of the complexities this implies for event handling.

\hypertarget{dimension-addressing-and-externalizability}{%
\subsubsection{Dimension: addressing and
externalizability}\label{dimension-addressing-and-externalizability}}

Programs in all programming systems have a representation that may be
exposed through notation such as source code. When customizing a
program, an interesting question is whether a customization needs to be
done by modifying the original representation, or whether it can be done
by \emph{adding} something alongside the original structure.

In order to support customization through addition, a programming system
needs a number of characteristics introduced by Basman et
al.~\cite{Externalize,OpenAuthorial}. First, the system needs to support
\emph{addressing}: the ability to refer to a part of the program
representation from the outside. Next, \emph{externalizability} means
that a piece of addressed state can be exhaustively transferred between
the system and the outside world. Finally, \emph{additive authoring}
requires that system behaviours can be \emph{changed} by simply
\emph{adding} a new expression containing addresses---in other words,
anything can be \emph{overriden} without being \emph{erased}. Of
particular importance is how addresses are specified and what extension
points in the program they can refer to. The system may offer an
automatic mechanism that makes certain parts of a program addressable,
or this task may be delegated to the programmer.

\emph{Cascading Style Sheets (CSS)}: CSS is a prime example of additive
authoring within the Web programming system. It provides rich
addressability mechanisms that are partly automatic (when referring to
tag names) and partly manual (when using element IDs and class names).
Given a web page, it is possible to modify almost any aspect of its
appearance by simply \emph{adding} additional rules to a CSS file. The
Infusion project \cite{Infusion} offers similar customizability
mechanisms, but for behaviour rather than just styling. There is also
th4e recent programming system Varv~\cite{varv}, which embodies additive
authoring as a core principle.

\emph{Object Oriented Programming and Aspect Oriented Programming}: in
conventional programming languages, customization is done by modifying
the code itself. OOP and AOP make it possible to do so by adding code
independently of existing program code. In OOP, this requires manual
definition of extension points, i.e.~interfaces and abstract methods.
Functionality can then be added to a system by defining a new class
(although injecting the new class into existing code without
modification requires some form of configuration such as a dependency
injection container). AOP systems such as AspectJ \cite{AspectJ}
provides a richer addressing mechanism. In particular, it makes it
possible to add functionality to the invocation of a specific method
(among other options) by using the \emph{method call pointcut}. This
functionality is similar to \emph{advising} in Pilot \cite{Pilot}.

\hypertarget{dimension-self-sustainability}{%
\subsubsection{Dimension:
self-sustainability}\label{dimension-self-sustainability}}

For most programming languages, programming systems, and ordinary
software applications, if one wants to customize beyond a certain point,
one must go beyond the facilities provided in the system itself. Most
programming systems maintain a clear distinction between the \emph{user
level}, where the system is used, and \emph{implementation level}, where
the source code of the system itself resides. If the user level does not
expose control over some property or feature, then one is forced to go
to the implementation level. In the common case this will be a
completely different language or system, with an associated learning
cost. It is also likely to be lower-level---lacking expressive
functions, features or abstractions of the user level---which makes for
a more tedious programming experience.

It is possible, however, to carefully design systems to expose deeper
aspects of their implementation \emph{at the user level}, relaxing the
formerly strict division between these levels. For example, in the
research system \emph{3-Lisp} \cite{PRinPLs}, ordinarily built-in
functions like the conditional \texttt{if} and error handling
\texttt{catch} are implemented in 3-Lisp code at the user level.

The degree to which a system's inner workings are accessible to the user
level, we call \emph{self-sustainability}. At the maximal degree of this
dimension would reside ``stem cell''-like systems: those which can be
progressively evolved to arbitrary behavior without having to ``step
outside'' of the system to a lower implementation level. In a sense, any
difference between these systems would be merely a difference in initial
state, since any could be turned into any other.

The other end, of minimal self-sustainability, corresponds to minimal
customizability: beyond the transient run-time state changes that make
up the user level of any piece of software, the user cannot change
anything without dropping down to the means of implementation of the
system. This would resemble a traditional end-user ``application''
focused on a narrow domain with no means to do anything else.

The terms ``self-describing'' or ``self-implementing'' have been used
for this property, but they can invite confusion: how can a thing
describe itself? Instead, a system that can \emph{sustain itself} is an
easier concept to grasp. The examples that we see of high
self-sustainability all tend to be \emph{Operating System-like}. UNIX is
widely established as an operating system, while Smalltalk and Lisp have
been branded differently. Nevertheless, all three have shipped as the
operating systems of custom hardware, and have similar responsibilities.
Specifically: they support the execution of ``programs''; they define an
interface for accessing and modifying state; they provide standard
libraries of common functionality; they define how programs can
communicate with each other; they provide a user interface.

\emph{UNIX}: Self-sustainability of UNIX is owed to the combination of
two factors. First, the system is implemented in binary files (via
ELF\footnote{Executable and Linkable Format.}) and text files (for
configuration). Second, these files are part of the user-facing
filesystem, so users can replace and modify parts of the system using
UNIX file interfaces.

\emph{Smalltalk and COLAs}: Self-sustainability in Smalltalk is similar
to UNIX, but at a finer granularity and with less emphasis on whether
things reside in volatile (process) or non-volatile (file) storage. The
analogous points are that (1) the system is implemented as objects with
methods containing Smalltalk code, and (2) these are modifiable using
the class browser and code editor. Combined Object Lambda Architectures,
or COLAs~\cite{COLAs}, are a theoretical system design to improve on the
self-sustainability of Smalltalk. This is achieved by generalizing the
object model to support relationships beyond classes.

\hypertarget{references}{%
\subsubsection{References}\label{references}}

In addition to the examples discussed above, the proceedings of
self-sustaining systems workshops
\cite{SelfSustaining2008,SelfSustaining2010} provides numerous examples
of systems and languages that are able to bootstrap, implement, modify,
and maintain themselves; Gabriel's analysis of programming language
revolutions \cite{PLrev} uses \emph{advising} in PILOT, related Lisp
mechanisms, and ``mixins'' in OOP to illustrate the difference between
the ``languages'' and ``systems'' paradigms.

\hypertarget{relations}{%
\subsubsection{Relations}\label{relations}}

\begin{itemize}
\tightlist
\item
  \emph{Flattening and factoring} (Section
  \ref{examples-flattening-and-factoring}): related in that
  ``customizability'' is a form of creating new programs from existing
  ones; factoring repetitive aspects into a reusable standard component
  library facilitates the same thing.
\item
  \emph{Interaction} (Section \ref{interaction}): this determines
  whether there are separate stages for running and writing programs and
  may thus influence what kind of customization is possible.
\end{itemize}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The technical dimensions framework is intended to assist designers of
programming systems. This section illustrates two ways in which it can
be used. First, we use the dimensions to analyze the recent programming
system Dark~\cite{DarkWeb}, explaining how it relates to past work and
how it contributes to the state of the art. Second, we use technical
dimensions to identify a new unexplored point in the design space of
programming systems and envision a new design that could emerge from the
analysis.

\hypertarget{evaluating-programming-systems}{%
\subsection{Evaluating programming
systems}\label{evaluating-programming-systems}}

Dark is a programming system for building ``serverless backends'',
i.e.~services that are used by web and mobile applications. It aims to
make building such services easier by ``removing accidental
complexity''\footnote{https://roadmap.darklang.com/goals-of-dark-v2.html}
resulting from the large number of systems typically involved in their
deployment and operation. This includes infrastructure for
orchestration, scaling, logging, monitoring and versioning. Dark
provides integrated tooling (Figure~\ref{fig:dark}) for development and
is described as \emph{deployless}, meaning that deploying code to
production is instantaneous.

\begin{figure}
  \centering
  \includegraphics[width=1\linewidth]{dark.png}
  \caption{The Dark programming environment showing a simple web service comprising a database, two HTTP endpoints and a worker.\label{fig:dark}}
  \note{FROM https://medium.com/@wilk/dark-lang-an-uncommon-step-towards-the-future-of-programming-921cf7f38baf}
\end{figure}

Dark illustrates the need for the broader perspective of programming
systems. Of course, it contains a programming language, which is
inspired by OCaml and F\#. But Dark's distinguishing feature is that it
eliminates the many secondary systems needed for deployment of modern
cloud-based services. Those exist outside of a typical programming
language, yet form a major part of the complexity of the overall
development process.

With technical dimensions, we can go beyond the ``sales pitch'', look
behind the scenes, and better understand the interesting technical
aspects of Dark as a programming system.

\hypertarget{dimension-analysis-of-dark}{%
\subsubsection{Dimension analysis of
Dark}\label{dimension-analysis-of-dark}}

\paragraph{Modes of interaction and feedback loops.}

Conventional \emph{modes of interaction}
(\ref{dimension-modes-of-interaction}) include running, editing and
debugging. For modern web services, running refers to operation in a
cloud-based environment that typically comes with further kinds of
feedback (logging and monitoring). The key design decision of Dark is to
integrate all these different modes of interaction into a single one.
This tight integration allows Dark to provide a more immediate
\emph{feedback loop} (\ref{dimension-feedback-loops}) where code changes
become immediately available not just to the developer, but also to
external users. The integrated mode of interaction is reminiscent of the
image-based environment in Smalltalk; Dark advances the state of art by
using this model in a multi-user, cloud-based context.

\paragraph{Feedback loops and error response.}

The integration of development and operation also makes it possible to
use \emph{errors} occurring during operation to drive development.
Specifically, when a Dark service receives a request that is not
supported, the user can build a handler to provide a response---taking
advantage of the live data that was sent as part of the request. In
terms of our dimensions, this is a kind of \emph{error response}
(\ref{dimension-error-response}) that was pioneered by the PILOT system
for Lisp~\cite{Pilot}. Dark does this not just to respond to errors, but
also as the primary development mechanism, which we might call
\emph{Error-Driven Development.} This way, Dark users can construct
programs with respect to sample input values.

\paragraph{Conceptual structure and learnability.}

Dark programs are expressed using high-level concepts that are specific
to the domain of server-side web programming: HTTP request handlers,
databases, workers and scheduled jobs. These are designed to reduce
accidental complexity and aim for high \emph{conceptual integrity}
(\ref{dimension-conceptual-integrity-vs.-openness}). At the level of
code, Dark uses a general-purpose functional language that emphasizes
certain concepts, especially records and pipelines. The high-level
concepts contribute to \emph{learnability}
(\ref{dimension-learnability}) of the system, because they are highly
domain-specific and will already be familiar to its intended users.

\paragraph{Notational structure and uniformity.}

Dark uses a combination of graphical editor and code. The two aspects of
the notation follow the \emph{complementing notations}
(\ref{dimension-notational-structure}) pattern. The windowed interface
is used to work with the high-level concepts and code is used for
working with low-level concepts. At the high level, code is structured
in freely positionable boxes on a 2D surface. Unlike Boxer \cite{Boxer},
these boxes do not nest and the space cannot be used for other content
(say, for comments, architectural illustrations or other media). Code at
the low level is manipulated using a syntax-aware structure editor,
showing inferred types and computed live values for pure functions. It
also provides special editing support for records and pipelines,
allowing users to add fields and steps respectively.

\paragraph{Factoring of complexity and automation.}

One of the advertised goals of Dark is to remove accidental complexity.
This is achieved by collapsing the heterogeneous stack of technologies
that are typically required for development, cloud deployment,
orchestration and operation. Dark hides this via \emph{factoring of
complexity} (\ref{dimension-factoring-of-complexity}). The advanced
infrastructure is provided by the Dark platform and is hidden from the
user. The infrastructure is programmed explicitly and there is no need
for sophisticated automation (\ref{dimension-level-of-automation}). This
factoring of functionality that was previously coded manually follows a
similar pattern as the development of garbage collection in high-level
programming languages.

\paragraph{Customizability.}

The Dark platform makes a clear distinction between the platform itself
and the user application, so \emph{self-sustainability}
(\ref{dimension-self-sustainability}) is not an objective. The strict
division between the platform and user (related to its aforementioned
\emph{factoring of complexity}) means that changes to Dark require
modifying the platform source code itself, which is available under a
license that solely allows using it for the purpose of contributing.
Similarly, applications themselves are developed by modifying and adding
code, requiring destructive access to it---so \emph{additive authoring}
(\ref{dimension-addressing-and-externalizability}) is not exhibited at
either level. Thanks to the integration of execution and development,
persistent changes may be made during execution (c.f. \emph{staging of
customization}, \ref{dimension-staging-of-customization}) but this is
done through the Dark editor, which is separate from the running
service.

\hypertarget{technical-innovations-of-dark}{%
\subsubsection{Technical innovations of
Dark}\label{technical-innovations-of-dark}}

This analysis reveals a number of interesting aspects of the Dark
programming system. The first is the tight integration of different
\emph{modes of interaction} which collapses a heterogeneous stack of
technologies, makes Dark \emph{learnable}, and allows quick feedback
from deployed services. The second is the use of \emph{error response}
to guide the development of HTTP handlers. Thanks to the technical
dimensions framework, each of these can be more precisely described. It
is also possible to see how they may be supported in other programming
systems. The framework also points to possible alternatives (and perhaps
improvements) such as building a more self-sustainable system that has
similar characteristics to Dark, but allows greater flexibility in
modifying the platform from within itself.

\hypertarget{exploring-the-design-space}{%
\subsection{Exploring the design
space}\label{exploring-the-design-space}}

With a little work, technical dimensions can let us see patterns or gaps
in the design space by plotting their values on a simple scatterplot.
Here, we will look at two dimensions, \emph{notational
diversity}\footnote{This is simply the dimension we named as
  \emph{uniformity of notations}
  (\ref{dimension-uniformity-of-notations}), but flipped in the opposite
  direction.} and \emph{self-sustainability}, for the following
programming systems: Haskell, Jupyter notebooks, Boxer, HyperCard, the
Web, spreadsheets, Lisp, Smalltalk, UNIX, and COLAs.

While our choice to identify and describe dimensions as qualitative
concepts was necessary for coming up with them, \emph{some} way of
generating numbers is clearly necessary for visualizing their
relationships like this. For simplicity,\footnote{There are undoubtedly
  many ways to turn our descriptions into various measures, with
  strengths and weaknesses for different purposes, but this is beyond
  the scope of the present paper. Here, we merely wish to demonstrate
  that such a thing is possible and show what one can do with the
  results.} we adopt the following scheme. For each dimension, we
distill the main idea into several yes/no questions
(Appendix~\ref{making-dimensions-quantitative}) that capture enough of
the distinctions we observe between the systems we wish to plot. Then,
for each system, we add up the number of ``yes'' answers and obtain a
plausible score for the dimension.

\begin{figure}
  \centering
  \includegraphics[width=0.6\linewidth]{plot-figure0.pdf}
  \caption{Example programming systems (or system families) measured against \emph{self-sustainability} and \emph{notational diversity}, revealing an absence of systems with a high degree of both. \label{fig:design-space-plot}}
\end{figure}

Figure~\ref{fig:design-space-plot} shows the results we obtained with
our sets of questions. It shows that the A-types span a range of
notational diversity, but only within fairly low self-sustainability.
The O-types cluster in an ``island'' at the right, sharing identical
notational diversity and near-identical self-sustainability. There is
also a conspicuous blank space at the top-right, representing an
unexplored combination of high values on both dimensions. With other
pairs of dimensions, we might take this as evidence of an oppositional
relationship, such that more of one inherently means less of the other
(perhaps looking for a single new dimension that describes this better.)
In this case, though, there is no obvious conflict between having many
notations and being able to change a system from within. Therefore, we
interpret the gap as a new opportunity to try out: combine the
self-sustainability of COLAs with the notational diversity of Boxer and
Web development. In fact, this is more or less the forthcoming
dissertation of the primary author.

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

There is a renewed interest in developing new programming systems. Such
systems go beyond the simple model of code written in a programming
language using a more or less sophisticated text editor. They combine
textual and visual notations, create programs through rich graphical
interactions, and challenge accepted assumptions about program editing,
execution and debugging. Despite the growing number of novel programming
systems, it remains difficult to evaluate the design of programming
systems and see how they improve over work done in the past. To address
the issue, we proposed a framework of ``technical dimensions'' that
captures essential characteristics of programming systems in a
qualitative but rigorous way.

The framework of technical dimensions puts the vast variety of
programming systems, past and present, on a common footing of
commensurability. This is crucial to enable the strengths of each to be
identified and, if possible, combined by designers of the next
generation of programming systems. As more and more systems are assessed
in the framework, a picture of the space of possibilities will gradually
emerge. Some regions will be conspicuously empty, indicating unrealized
possibilities that could be worth trying. In this way, a domain of
``normal science'' is created for the design of programming systems.

\acks

(To be completed for publication.)

### 2022-09-22 prog-2022/prog22-master.tex
Reinclude appendix in main and final minor changes

@@
 \usepackage{pifont} % For checkmark / cross symbols for Appendix table :)
+\usepackage{amsthm} % For numberless Definition
+\usepackage{epigraph} % For the opening quotes
 
@@
 
-NOTE TO REVIEWERS: The main contribution of the paper is a comprehensive survey of 22 design dimensions of programming systems.
+NOTE TO REVIEWERS: The main contribution of the paper is a comprehensive catalogue of 22 design dimensions of programming systems.
 Each one comes with detailed discussion, including known uses, to properly motivate it.
-Regrettably, we only have space for a small number of these dimensions in the main paper body.
-To give a broader picture of what we are contributing, we invite the reader to explore Appendix\ \ref{dimensions-catalogue} for dimensions that suit their interest.
+We feel these belong in the main paper body rather than in an appendix, even though this increases the page count, and have received permission to do this.
+We intend the catalogue as a reference to be used \emph{as necessary,} not to be read from start to finish.
+This means that Section\ \ref{technical-dimensions-catalogue} can be considered to be \textbf{outside} of the normal reviewable material, leaving the rest of the paper within the journal's 22 page limit.
 \end{abstract}
@@
 \providecommand{\tightlist}{}% Don't want Pandoc's tight lists
+\newtheorem*{defn}{Definition}
 %\newcommand{\hypertarget}[1]{}

### 2022-09-27 prog-2022/changelog.tex
New cover letter, ensure nbsp pre cites

@@
 
-We graciously thank the reviewers for their detailed feedback. We've
-made many minor changes to the paper and several major changes to both
-structure and content.
-
-Two issues particularly stood out:
+We thank the reviewers for their continued attention and feedback. The
+major changes we have made are as follows.
 
@@
 
-All three reviewers requested clarification on what constitutes a
-``system'', mentioning IDEs amongst other things as having unclear
-membership. Unfortunately, we truly don't have a definition behind the
-scenes which we're using to include or exclude examples. It's more of a
-``family resemblance'' between things that generalize ``programming
-languages'' to ``software which lets one program, yet doesn't have to be
-a language''. Nevertheless, we thought it might be worth adding a new
-classification scheme for the examples we mention, which should get
-across the key ``reference classes'' from which we're drawing. In sum,
-these are: things that are like operating systems (O-type), things that
-are like domain-specific applications (A-type) and the systems that are
-based around a programming language (L-type). We also went into detail
-about how a programming language (Java) can be interpreted as a system
-in several different ways. These were all additions to Section 3.
+We worked out a sensible explicit definition of what a programming
+system is and included it as early as possible in the paper.
 
-\paragraph{But how do you use it?}
+\paragraph{L/O/A typology.}
 
-Two reviewers were concerned that it's hard to see how to use the
-framework. We've rectified this by including a new, concrete
-``Discussion'' section with two halves: one using the framework for
-analysis, the other for synthesis. Specifically, we analyze Dark, a
-recent programming system, along many of the dimensions. Then, we plot
-two dimensions (self-sustainability vs.~notational diversity) against
-each other to reveal an unexplored gap. The methodology we used to turn
-qualitative dimensions into numbers is given in a new Appendix. Various
-points earlier in the paper (e.g.~the Abstract and Figure 1) have been
-updated to set up for this new section.
+We intended the three-way classification in section~3 merely to provide
+some rhyme and reason as to where our example systems were coming from.
+We did not intend it as a rigid scheme where membership in a category
+means we are making some profound or bold claim. The L/O/A-type naming
+seems to have misled readers into anticipating the latter, so we have
+relaxed the naming to ``language-based'', ``OS-like'' and
+``application-focused''. We did refer back to this scheme in one
+specific context, in section~5.2, noting that most OS-likes (Unix, Lisp
+and Smalltalk) cluster in a region to the right.
 
-Other changes:
+\paragraph{Dark \& evaluation.}
 
-\begin{itemize}
-\tightlist
-\item
-  We've significantly reworked Section 3 to be more to-the-point about
-  the example systems and less of a historical narrative.
-\item
-  We've argued for why we think ``sociability'' belongs as a technical
-  dimension in the relevant section.
-\item
-  We've reorganized and re-worded parts of Section 2. We've also made
-  the part about how the framework relates to Cognitive Dimensions in
-  Section 2, under ``Already-known characteristics'', more explicit.
-\item
-  We've clarified some of the points about Direct Manipulation
-  (primitive actions) and Immediate Feedback (pure code previews).
-\item
-  Under ``Addressing and Externalizability'', we've made it explicit
-  that CSS provides Additive Authoring for the Web programming system.
-\item
-  To make room for the ``Discussion'' section, we've had to move more
-  material to the existing Appendix with the bulk of the dimensions.
-  Now, the main body only contains the Customizability dimensions.
-\item
-  We've elevated the singleton Automation cluster to a doublet by adding
-  a Factoring of Complexity dimension.
-\end{itemize}
+Section~5 has been clarified as an evaluation of the framework by means
+of the two example usages. We've included a summary table for the
+dimensional analysis of Dark in section~5.1 and annotated the screenshot
+for convenience. We appreciate the reviewer's desire for examples of the
+specific interactions, but we thought it'd be pointless to write down
+something already explained in one of Dark's tutorial videos, so we
+cited the latter instead.
+
+\paragraph{References to strengthen claims.}
+
+Our statement about ``cool/interesting'' was entirely from experience,
+so we relaxed it to something we could more easily support with
+references. These can be seen in a footnote with our remarks linking
+them to the point we're trying to make. Elsewhere we have addressed this
+by deleting the claim (e.g.~for ``holistic view'') or adding references,
+for example to the various example systems like Java and Jupyter. We're
+afraid we don't understand the request to put explicit examples in the
+section on venues for publishing programming systems research; we'd
+already listed UIST, VL/HCC, LIVE and PX. If this was about specific
+publications instead, we've now referred to some in the footnote on page
+3.
+
+\paragraph{Socio-Technical dimensions?}
+
+The fact of the matter is that almost all of our 22 dimensions are
+technical, with the possible exception of Learnability and Sociability
+being socio-technical, and some remarks in Integrity vs.~Openness. On
+this basis, we strongly disagree with calling the contribution
+``socio-technical'' dimensions or similar, as this would give a false
+expectation to readers. We appeal to the reviewer to permit us to keep
+the name ``technical''---we put a lot of effort into thinking of names
+early on, but ``technical dimensions'' was the best out of perhaps only
+two viable choices. If this is still unacceptable, we will fall back on
+the other name, ``design dimensions'', but if this is still rejected we
+would need some help determining an alternative.
+
+\paragraph{On the new length of the paper.}
+
+Following approval from the editors, this revision includes the full
+catalogue of dimensions in the main body of the paper. While we
+acknowledge this makes it very long, we do not intend reviewers or
+readers to read this section from start to finish. Instead, it is meant
+as a reference to be used as needed, for example to get further detail
+on the dimensions we reference later for the Dark programming system.
+This lets us present the material as we originally intended and
+addresses a reviewer concern that it was awkward to refer to appendix
+material in the section on Dark. In summary, we always saw the
+dimensions catalogue as belonging in the main body, but for review
+purposes it may still be treated like an appendix.

### 2022-09-29 prog-2022/changelog.tex
Penultimate pre-submission stuff

@@
 
-Section~5 has been clarified as an evaluation of the framework by means
+Section~5 has been reframed as an evaluation of the framework by means
 of the two example usages. We've included a summary table for the
@@
 for convenience. We appreciate the reviewer's desire for examples of the
-specific interactions, but we thought it'd be pointless to write down
-something already explained in one of Dark's tutorial videos, so we
-cited the latter instead.
+specific interactions. We feel these are best demonstrated through one
+of Dark's tutorial videos, so we added a reference to one that addresses
+this concern.
 
@@
 
-Our statement about ``cool/interesting'' was entirely from experience,
-so we relaxed it to something we could more easily support with
-references. These can be seen in a footnote with our remarks linking
-them to the point we're trying to make. Elsewhere we have addressed this
-by deleting the claim (e.g.~for ``holistic view'') or adding references,
-for example to the various example systems like Java and Jupyter. We're
-afraid we don't understand the request to put explicit examples in the
-section on venues for publishing programming systems research; we'd
-already listed UIST, VL/HCC, LIVE and PX. If this was about specific
-publications instead, we've now referred to some in the footnote on page
-3.
+Our statement about ``vaguely interesting'' was entirely from
+experience, so we relaxed it to something we could more easily support
+with references. These can be seen in a footnote with our remarks
+linking them to the point we're trying to make. Elsewhere we have
+addressed this by deleting the claim (e.g.~for ``holistic view'') or
+adding references, for example to the various example systems like Java
+and Jupyter. As for explicit examples of venues for publishing
+programming systems research: we already listed several examples in
+section~2 (UIST, VL/HCC, LIVE, PX) and while further venues could be
+listed, we believe these illustrate the range well enough. If this was
+about specific publications instead, we've now referred to some in the
+footnote on page 3.
 
@@
 
-The fact of the matter is that almost all of our 22 dimensions are
-technical, with the possible exception of Learnability and Sociability
-being socio-technical, and some remarks in Integrity vs.~Openness. On
-this basis, we strongly disagree with calling the contribution
+Almost all of our 22 dimensions are straightforwardly technical, with
+the possible exception of Learnability and Sociability being
+socio-technical, and some remarks in Integrity vs.~Openness. Even though
+there are some social aspects, we hope to emphasise the technical side.
+On this basis, we would rather not call the contribution
 ``socio-technical'' dimensions or similar, as this would give a false
-expectation to readers. We appeal to the reviewer to permit us to keep
-the name ``technical''---we put a lot of effort into thinking of names
-early on, but ``technical dimensions'' was the best out of perhaps only
-two viable choices. If this is still unacceptable, we will fall back on
-the other name, ``design dimensions'', but if this is still rejected we
-would need some help determining an alternative.
+expectation to readers. Our preferred naming remains ``technical''---we
+put a lot of effort into thinking of names early on, but ``technical
+dimensions'' was the best out of perhaps only two viable choices. If
+this is unacceptable, we could fall back on the other name, ``design
+dimensions'', but if that doesn't work we would need some help
+determining an alternative.
 

### 2022-09-30 prog-2022/prog22-master.tex
Final pre-submission tweaks

@@
   \emph{Importance.} Much research effort focuses on building programming systems that are easier to use, accessible to non-experts, moldable and/or powerful, but such efforts are disconnected.
-They are informal, guided by the personal vision of the authors and thus are only evaluable and comparable on the basis of individual experience using them.
+They are informal, guided by the personal vision of their authors and thus are only evaluable and comparable on the basis of individual experience using them.
 By providing foundations for more systematic research, we can help programming systems researchers to stand, at last, on the shoulders of giants.

### 2022-11-01 onward-2022/onward22-master.tex
Camera-ready corrections / tweaks

@@
-\documentclass[sigplan,review]{acmart}
+\documentclass[sigplan,screen]{acmart}
+\usepackage[utf8]{inputenc}
+%\usepackage[T1]{fontenc}
 \usepackage{listings}
@@
 \usepackage{amsthm}
-
-\acmConference{Onward! Essays}{December 5--10, 2022}{Auckland, New Zealand}
+%\usepackage{pifont} % For checkmark / cross symbols
+\usepackage{microtype}
+\usepackage{flushend}
 
 \title{Ascending the Ladder to Self-Sustainability}
-\subtitle{Achieving open evolution in an interactive graphical system}
+\subtitle{Achieving Open Evolution in an Interactive Graphical System}
 \author{Joel Jakubovic}
@@
 
+\setcopyright{acmlicensed}
+\acmPrice{15.00}
+\acmDOI{10.1145/3563835.3568736}
+\acmYear{2022}
+\copyrightyear{2022}
+\acmSubmissionID{onward22essays-p6-p}
+\acmISBN{978-1-4503-9909-8/22/12}
+\acmConference[Onward!
+'22]{Proceedings of the 2022 ACM SIGPLAN International Symposium on New Ideas, New Paradigms, and Reflections on Programming and Software}{December 8--10, 2022}{Auckland, New Zealand}
+\acmBooktitle{Proceedings of the 2022 ACM SIGPLAN International Symposium on New Ideas, New Paradigms, and Reflections on Programming and Software (Onward!
+'22), December 8--10, 2022, Auckland, New Zealand}
+\received{2022-09-05}
+\received[accepted]{2022-10-03}
+
 \begin{document}
+
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10011007.10011006.10011066.10011069</concept_id>
+       <concept_desc>Software and its engineering~Integrated and visual development environments</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011050</concept_id>
+       <concept_desc>Software and its engineering~Context specific languages</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011041.10011048</concept_id>
+       <concept_desc>Software and its engineering~Runtime environments</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Integrated and visual development environments}
+\ccsdesc[500]{Software and its engineering~Context specific languages}
+\ccsdesc[300]{Software and its engineering~Runtime environments}
+
+\keywords{self-sustaining, self-sustainable, interactive, bootstrapping, visual programming, context-specific, domain-specific}
 
@@
 
-\bibliographystyle{plainurl}
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{onward22}
+
 \end{document}

### 2023-01-26 prog-2022/prog22-master.tex
Prog22 camera-ready changes

@@
-\documentclass[english,submission]{programming}
+\documentclass[english,crc]{programming}
 
@@
 
+%%% The following is specific to Programming Journal, Volume 7, Issue 3 and the paper
+%%% 'Technical Dimensions of Programming Systems'
+%%% by Joel Jakubovic, Jonathan Edwards, and Tomas Petricek.
+%%%
+\paperdetails{
+  submitted=2022-09-30,
+  published=2023-02-15,
+  year=2023,
+  volume=7,
+  issue=3,
+  articlenumber=13,
+}
+
 \begin{document}
@@
 \affiliation{
-University of Kent, Canterbury, UK
-\email{T.Petricek@kent.ac.uk}
+University of Kent, Canterbury, UK \\
+Charles University, Prague, Czechia
+\email{tomas@tomasp.net}
 }
@@
 By providing foundations for more systematic research, we can help programming systems researchers to stand, at last, on the shoulders of giants.
-
-NOTE TO REVIEWERS: The main contribution of the paper is a comprehensive catalogue of 22 design dimensions of programming systems.
-Each one comes with detailed discussion, including known uses, to properly motivate it.
-We feel these belong in the main paper body rather than in an appendix, even though this increases the page count, and have received permission to do this.
-We intend the catalogue as a reference to be used \emph{as necessary,} not to be read from start to finish.
-This means that Section\ \ref{technical-dimensions-catalogue} can be considered to be \textbf{outside} of the normal reviewable material, leaving the rest of the paper within the journal's 22 page limit.
 \end{abstract}

### 2023-02-17 prog-2022/prog22-master.tex
Final camera-ready changes

@@
 
-% Class scrartcl Warning: seems someone has broken package `auxhook'.
-% Usually this happens, if `auxhook' is loaded or used implicitly or explicitly
-% by patching \document or via etoolbox command \AtEndPreamble. Trying an
-% emergency workaround. You can avoid this warning adding:
-\usepackage{auxhook}
-% before \begin{document} on input line 6.
-\usepackage{changepage} % For the single-page summary table PDF insert.
+%\usepackage{changepage}
+%\usepackage{lscape} % For the summary table
 \usepackage{pdfpages}
@@
 
-\author{Joel Jakubovic}
-\affiliation{
-University of Kent, Canterbury, UK
-\email{jdj9@kent.ac.uk}
-}
-\author{Jonathan Edwards}
-\affiliation{\email{jonathanmedwards@gmail.com}}
-\author{Tomas Petricek}
-\affiliation{
-University of Kent, Canterbury, UK \\
-Charles University, Prague, Czechia
-\email{tomas@tomasp.net}
-}
+\author[a]{Joel Jakubovic}
+\authorinfo[joel]{is Tomas' final-year PhD student interested in overcoming pervasive ``premature commitment'' in programming.
+His forthcoming dissertation is about creating programming systems that combine self-sustainability and notational freedom.
+Send him an e-mail at \email{jdj9@kent.ac.uk} and read his blog at \href{https://programmingmadecomplicated.wordpress.com}{\texttt{programmingmadecomplicated.wordpress.com}}.}
+\author[b]{Jonathan Edwards}
+\authorinfo[jonathan]{is an independent researcher working on drastically simplifying programming.
+He is known for his Subtext series of programming language experiments and his blog at \href{https://alarmingdevelopment.org}{\texttt{alarmingdevelopment.org}}.
+He has been a researcher at MIT CSAIL and CDG/HARC.
+He tweets \texttt{@jonathoda} and can be reached at \email{jonathanmedwards@gmail.com}.}
+\author[a,c]{Tomas Petricek}
+\authorinfo[tomas]{is an assistant professor at Charles University.
+He is interested in finding easier and more accessible ways of thinking about programming.
+To do so, he combines technical work on programming systems and tools with research into history and philosophy of science.
+His work can be found at \href{https://tomasp.net}{\texttt{tomasp.net}} and he can be reached at \email{tomas@tomasp.net}.}
+\affiliation[a]{University of Kent, Canterbury, UK}
+\affiliation[b]{Independent}
+\affiliation[c]{Charles University, Prague, Czechia}
 
@@
 
+\newcommand{\ciakgi}[1]{}
+
 \begin{abstract}
-  \emph{Context.} Programming requires much more than just writing code in a programming language.
+  \ciakgi{Context} Programming requires much more than just writing code in a programming language.
 It is usually done in the context of a stateful environment, by interacting with a system through a graphical user interface.
@@
 
-  \emph{Inquiry.} In computer science, much has been said and done to allow comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems;} we believe that programming systems deserve a theory too.
+  \ciakgi{Inquiry} In computer science, much has been said and done to allow comparison of \emph{programming languages}, yet no similar theory exists for \emph{programming systems;} we believe that programming systems deserve a theory too.
 
-  \emph{Approach.} We present a framework of \emph{technical dimensions} which capture the underlying characteristics of programming systems and provide a means for conceptualizing and comparing them.
+  \ciakgi{Approach} We present a framework of \emph{technical dimensions} which capture the underlying characteristics of programming systems and provide a means for conceptualizing and comparing them.
 
-  \emph{Knowledge.} We identify technical dimensions by examining past influential programming systems and reviewing their design principles, technical capabilities, and styles of user interaction.
+  \ciakgi{Knowledge} We identify technical dimensions by examining past influential programming systems and reviewing their design principles, technical capabilities, and styles of user interaction.
 Technical dimensions capture characteristics that may be studied, compared and advanced independently.
@@
 
-  \emph{Grounding.} Our framework is derived using a qualitative analysis of past programming systems.
+  \ciakgi{Grounding} Our framework is derived using a qualitative analysis of past programming systems.
 We outline two concrete ways of using our framework.
@@
 
-  \emph{Importance.} Much research effort focuses on building programming systems that are easier to use, accessible to non-experts, moldable and/or powerful, but such efforts are disconnected.
+  \ciakgi{Importance} Much research effort focuses on building programming systems that are easier to use, accessible to non-experts, moldable and/or powerful, but such efforts are disconnected.
 They are informal, guided by the personal vision of their authors and thus are only evaluable and comparable on the basis of individual experience using them.
@@
 \newcommand{\tp}[1]{}
-\newcommand{\mybox}[1]{\noindent\fbox{\parbox{\textwidth}{#1}}}
+%\newcommand{\mybox}[1]{\noindent\fbox{\parbox{\textwidth}{#1}}}
+\newcommand{\mybox}[1]{\textbf{#1}}
 \providecommand{\tightlist}{}% Don't want Pandoc's tight lists

### 2023-02-17 prog-2022/summary-table-old.tex
Final camera-ready changes

--- 2022-05-18 prog-2022/summary-table.tex
@@
 Source code vs. config files, Developer Tools tab, auto image-based persistence, scripting language \\
-Externalizability &
-Which portions of the system's state can be referenced and transferred to/from it? &
-None (state is private) vs. all state exposed as human-legible, CSS-like addressing \\
-Additive Authoring &
-How far can the system's behavior be changed by \emph{adding} expressions? &
-None (requires power to change original) vs. full (anything can be overridden repeatedly) \\
+Externalizability / Additive Authoring &
+Which portions of the system's state can be referenced and transferred to/from it? How far can the system's behavior be changed by \emph{adding} expressions? &
+State is: (\emph{i}) private and hidden; (\emph{ii}) externalizable and overwriteable; (\emph{iii}) externalized and \emph{overridable} by addition \\
 Self-Sustainability &

### 2023-02-17 prog-2022/tech-dims-table-portrait.tex
Final camera-ready changes

--- 2023-02-17 prog-2022/summary-table-old.tex
@@
-\newcommand{\summ}[1]{\multicolumn{2}{p{0.8\linewidth}}{#1} \\ \hline}
+\newcommand{\summ}[1]{\multicolumn{2}{>{\hsize=\dimexpr2\hsize+2\tabcolsep+\arrayrulewidth\relax}X}{#1} \\ \hline}
+%\newcommand{\summ}[1]{\\}
+
 \setlength\extrarowheight{2pt}
+
 \begin{table}
-%\centering
-\begin{tabular}{ >{\raggedleft\arraybackslash}p{0.2\linewidth}  p {0.4\linewidth}  >{\raggedright\arraybackslash}p{0.4\linewidth} }
+\caption{Technical Dimensions cheat sheet.}
+\begin{tabularx}{\linewidth}{ >{\raggedleft\arraybackslash}p{3.75cm} X X }
 \hline
@@
 \hline INTERACTION &
-\summ{How do users execute their ideas, evaluate the result, and generate new ideas in response?}
+\summ{How do users manifest their ideas, evaluate the result, and generate new ideas in response?}
 Feedback Loops &
@@
 Lisp S-expressions vs. English-like textual notations \\
-\hline ARCHITECTURE &
-\summ{What is the system's attitude to its own organization and that of the outside world?}
-Conceptual Outlook &
-Does the system present as \emph{designed} or \emph{improvised}? &
+\end{tabularx}
+\label{tab:summary} 
+\end{table}
+
+\begin{table}
+\caption{Technical Dimensions cheat sheet (continued).}
+\begin{tabularx}{\linewidth}{ >{\raggedleft\arraybackslash}p{3.5cm} X X }
+\hline
+Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
+\hline CONCEPTUAL STRUCTURE &
+\summ{How is meaning constructed? How are internal and external incentives balanced?}
+Conceptual Integrity vs. Openness &
+Does the system present as elegantly \emph{designed} or pragmatically \emph{improvised}? &
 Integrity (Everything is a X) vs. openness (compatible mixtures) \\
@@
 How much is \emph{common structure} explicitly marked as such? &
-Common structure is redundantly flattened vs. factored out\\
-\hline
-\end{tabular}
-\caption{\label{summary-table1} Quick reference sheet for our set of Technical Dimensions.}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ >{\raggedleft\arraybackslash}p{0.2\linewidth}  p {0.4\linewidth}  >{\raggedright\arraybackslash}p{0.4\linewidth} }
-\hline
-Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
+Common structure is redundantly flattened vs. factored out \\
 \hline CUSTOMIZABILITY &
@@
 How far can the system’s behavior be changed from within? &
-None (rely on extenal tools) vs. self-sufficient (contains everything needed) \\
+None (rely on external tools) vs. self-sufficient (contains everything needed) \\
+\end{tabularx}
+\label{tab:summary2}
+\end{table}
+
+\begin{table}
+\caption{Technical Dimensions cheat sheet (continued).}
+\begin{tabularx}{\linewidth}{ >{\raggedleft\arraybackslash}p{3.5cm} X X }
+\hline
+Dimension (CLUSTER) & Summary & Range of key examples \\ \hline
 \hline COMPLEXITY &
@@
 What programming details are hidden in reusable components and how? &
-Domain-specific (more hiding) vs. general-purpose (less hiding) programming \\
+Domain-specific (more hiding) vs. general-purpose (less hiding) \\
 Level of Automation &
@@
 \hline ERRORS &
-\summ{What does the system consider to be an error; how are they prevented and handled?}
+\summ{What does the system consider to be an \emph{error}? How are they prevented and handled?}
 Error Detection &
@@
 What are the social and economic factors that make the system the way it is? &
-Funding, volunteers, code sharing, Q/A sites vs. documentation, sense of belonging \\
-\hline
-\end{tabular}
-\caption{\label{summary-table2} Quick reference sheet for our set of Technical Dimensions.}
+Cathedral vs. Bazaar model \\
+\end{tabularx}
+\label{tab:summary3}
 \end{table}
